---
title: "Movie Data Analysis — Part 4: Summary and Reflection"
author: "TBD"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	comment = NA,
	fig.align = "center",
	fig.width = 10,
	fig.height = 5,
	out.width = "100%"
)

options(repos = c(CRAN = "https://cloud.r-project.org"))
```

```{r load-and-clean, include=FALSE}
# ── Data preparation (repeated from Part 1 so this file is self-contained) ──
#
# We reload the raw CSV and re-apply the cleaning rules established in
# Part 1. Part 4 does not need the full set of derived columns from Part 2
# or the models from Part 3 — it only needs the basic "before vs. after
# cleaning" numbers so that the summary table at the end of this document
# reports real, reproducible figures instead of numbers typed in by hand.

movie_profit <- read.csv("movie_profit.csv", stringsAsFactors = FALSE)

# Date conversion (same rule as Part 1)
release_date_raw          <- movie_profit$release_date
movie_profit$release_date <- as.Date(release_date_raw, format = "%m/%d/%Y")

# Row-level filter (same rule as Part 1):
#   • drop rows with any missing value outside of mpaa_rating
#   • drop rows where domestic_gross or worldwide_gross equals zero
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")
row_has_NA         <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])
gross_is_zero      <- movie_profit$domestic_gross  == 0 |
                      movie_profit$worldwide_gross == 0

movie_profit_clean <- movie_profit[!(row_has_NA | gross_is_zero), ]

# A few headline numbers reused throughout the narrative below
n_raw       <- nrow(movie_profit)
n_clean     <- nrow(movie_profit_clean)
n_removed   <- n_raw - n_clean
pct_removed <- round(100 * n_removed / n_raw, 1)

year_min <- format(min(movie_profit_clean$release_date, na.rm = TRUE), "%Y")
year_max <- format(max(movie_profit_clean$release_date, na.rm = TRUE), "%Y")

n_genres <- length(unique(movie_profit_clean$genre))
```

---

## Overview

This document covers **Part 4 — Summary and Reflection**, the closing
part of the Movie Data Analysis project. Parts 1–3 did the technical
work: cleaning the raw data, exploring it in depth, and building two
statistical models to investigate how production budget and genre
relate to worldwide box-office revenue, with a particular focus on the
Horror vs. non-Horror comparison.

This final part steps back from code and numbers and ties everything
together. It is organised into five sections:

1. **Project summary** — a concise recap of what was done in each part
2. **Most important findings** — the key results from the EDA and the
   models, brought together in one place
3. **Main conclusions** — what these findings mean in practical terms
4. **Limitations and difficulties** — what the data and models could
   *not* tell us, and why
5. **Reflection on the process** — what was hardest, what was most
   interesting, and what we would do differently with more time,
   knowledge, or resources

---

## 1  Project Summary

The project worked with a dataset of `r n_raw` films spanning the five
genres **Action, Adventure, Comedy, Drama, and Horror**, released
between `r year_min` and `r year_max`. The work was carried out in four
stages, each documented in its own part of the report.

**Part 1 — Data cleaning and preparation.** The raw CSV was loaded and
inspected for variable types, missing values, duplicates, and logical
inconsistencies (e.g. `worldwide_gross` smaller than `domestic_gross`).
`release_date` was converted from a text column to a proper `Date`. Rows
with missing values in any column other than `mpaa_rating` (which was
deliberately *kept*, since the rating is not needed for the core
analysis and dropping it would have removed many otherwise-usable rows),
or with a `domestic_gross`/`worldwide_gross` of exactly zero, were
removed. This step removed `r n_removed` rows (`r pct_removed`% of the
original data), leaving `r n_clean` films for analysis — a small enough
fraction that the dataset's overall structure was not materially
affected.

**Part 2 — Exploratory data analysis (EDA).** Every original column was
documented, and **12 new columns** were derived (above the required
minimum of 10), including `international_gross`, two ROI factors
(`roi_domestic_factor`, `roi_worldwide_factor`), two "top 100"
blockbuster flags, the `is_horror_movie` / `movie_type` grouping
variables, a `budget_band` categorical variable, two financial columns
rescaled to thousands of dollars for the interactive table, and a
formatted release date plus a `release_year` integer for time-series
charts. **13 charts** were produced, each accompanied by a short
explanation of why it was created and what it shows — covering genre
and rating distributions, releases over time, distributor market share,
and — most importantly — the Horror vs. non-Horror comparison across
counts, ROI, and budget bands.

**Part 3 — Modeling.** Building on the EDA, the research question *"Does
a higher production budget lead to higher worldwide revenue, and does
this relationship differ across genres?"* was formalised and tested with
two nested linear regression models on log-transformed budget and
revenue: **Model 1**, a main-effects model with a single budget slope
for all genres, and **Model 2**, an interaction model allowing the
budget slope (elasticity) to vary by genre. The two models were compared
on goodness of fit, interpretability, and practical usefulness.

**Part 4 — Summary and reflection.** Brings the threads
from Parts 1–3 together, draws overall conclusions, and reflects honestly
on the process — including what was difficult, what was interesting, and
what we would change with more time or resources.

---

## 2  Most Important Findings

### 2.1  What the exploratory analysis showed

- **Low-budget Horror films achieve unusually high ROI.** Among the
  cheapest Horror productions (budgets ≤ \$1M), several returned more
  than 100× their production cost, and the large majority returned at
  least 8× — a pattern not seen at the same scale in any other genre.
- **This advantage is structural, not a handful of lucky outliers.**
  Both the median ROI comparison and the budget-band distribution showed
  that Horror consistently occupies a different financial "regime" from
  Action, Adventure, Comedy, and Drama, even after accounting for
  outliers.
- **Bigger budgets do not help Horror the way they help other genres.**
  The scatter of budget vs. revenue for low-budget films suggested that,
  for Horror, spending more does not translate into proportionally more
  revenue — a pattern that motivated the modelling question in Part 3.

### 2.2  What the models showed

- **Budget is a strong, statistically significant predictor of worldwide
  revenue** for every genre (p < 2×10⁻¹⁶ in both models). On the
  log–log scale, a 1% increase in budget is associated with roughly a
  1% increase in revenue for most genres — close to a proportional
  relationship.
- **Horror is the clear outlier.** Its estimated budget elasticity
  (≈ 0.68) is well below that of the reference genre Drama (≈ 0.97) and
  below Comedy, Adventure, and Action, all of which showed elasticities
  at or above 1. In plain terms: for Horror, an extra dollar of budget
  buys noticeably less extra revenue than it does for other genres.
- **The interaction model (Model 2) fits the data better than the
  simpler main-effects model (Model 1)**, confirming that genre
  *moderates* the budget–revenue relationship rather than just shifting
  it up or down.
- **Budget and genre together explain roughly 43–44% of the variance in
  log-revenue** (adjusted R² ≈ 0.436 for the interaction model). This is
  a meaningful share, but it also means more than half of the variation
  in box-office revenue is driven by factors not captured in this
  dataset.

---

## 3  Main Conclusions

Putting the EDA and modelling results together, three practical
conclusions stand out:

1. **Low-budget Horror is a genuinely different segment of the film
   industry, not just a curiosity.** Both the descriptive statistics and
   the regression elasticities point the same direction: Horror films
   can be extremely profitable *relative to their cost*, even though
   they rarely achieve the largest absolute revenues. This explains why
   there are so many cheap Horror productions — money is simply not the
   main lever for this genre's success.

2. **For most genres, "spend more to earn more" roughly holds — but not
   universally.** Action, Adventure, and Comedy show budget elasticities
   at or above 1, so increased investment is, on average, rewarded with
   at least proportional revenue gains. Horror is the exception, which
   has direct implications for how studios might think about
   genre-specific investment strategy.

3. **A single "one model fits all genres" view of the budget–revenue
   relationship would be an oversimplification.** The interaction model
   consistently outperformed the additive model, supporting the idea
   that genre-aware models are more useful for anyone trying to use
   budget as a lever for predicting or planning revenue.

At the same time, the moderate R² values are a useful reminder that
production budget and genre — while informative — are far from the whole
story behind a film's box-office performance.

---

## 4  Limitations and Difficulties

Several limitations affect how far these conclusions can be generalised:

- **No marketing data.** Marketing spend can rival or exceed production
  budgets for major releases. Its absence likely inflates the apparent
  importance of the production budget coefficient, since some of
  marketing's effect is probably being absorbed into the budget
  variable.
- **Cross-sectional treatment of a 1936–2018 time span.** The models
  treat all years as equivalent, but ticket prices, inflation, streaming
  competition, and global distribution patterns have changed enormously
  over this period. A film's production budget and gross revenue are not
  directly comparable across decades without some form of inflation
  adjustment.
- **Genre as a single label.** Many films could reasonably be classified
  under more than one genre (e.g. an action-horror crossover), but the
  dataset assigns exactly one genre per film. This introduces some
  measurement error into the genre variable and, by extension, into the
  genre-specific elasticities.
- **Homoscedasticity assumption.** Even after the log transformation,
  residual variance may not be perfectly constant across the budget
  range. Robust standard errors (e.g. HC3) would give more reliable
  inference but were outside the scope of this project.
- **Missing `mpaa_rating` values.** `mpaa_rating` had a non-trivial
  number of missing entries and was therefore excluded from the
  row-completeness check in Part 1 (otherwise too many otherwise-usable
  rows would have been dropped). As a result, it was used descriptively
  in the EDA but not as a model predictor.
- **External validity.** The dataset is a curated selection of
  theatrically-released films. Straight-to-streaming releases,
  international productions outside major US distribution, and very
  recent films are largely absent, which limits how far these findings
  can be extended to "the movie industry" as a whole today.

---

## 5  Reflection on the Process

### 5.1  What was the most difficult part of the project?

The most challenging stage was the transition between **EDA and
modelling** — specifically, deciding *how* to formalise the pattern
spotted during EDA (cheap Horror films with huge ROI) into a model that
was both statistically sound and easy to interpret. Working on the
log–log scale made the elasticity interpretation cleaner, but it also
required being careful about strictly-positive values, choosing a
sensible reference genre, and explaining what an "elasticity" actually
means in terms a non-statistician would understand. Cleaning decisions
in Part 1 — particularly how to handle the missing `mpaa_rating` values
without discarding a large share of the dataset — also took more
iteration than expected, since every cleaning choice had knock-on
effects for the row counts reported throughout the rest of the report.

### 5.2  What was the most interesting part?

The most interesting moment was seeing the **Horror anomaly confirmed
statistically** after first noticing it informally during EDA. It is one
thing to see a handful of cheap Horror films with very high ROI in a
table; it is more convincing to see the same pattern show up as a
significantly lower budget elasticity in a regression model, controlling
for budget itself. Connecting a simple, intuitive observation
("cheap Horror movies seem to do really well relative to their cost")
to a formal statistical result was the most rewarding part of the
project.

### 5.3  What would we do differently if we started again?

If starting again, it would help to **plan the derived columns and the
research question earlier and together**, rather than designing the
EDA columns first and only formulating the modelling question afterwards.
A few of the columns created in Part 2 (e.g. the "top 100" blockbuster
flags) ended up being more useful for exploration than for the final
models. Sketching the modelling question at the start — even in a rough
form — would have made it easier to design derived columns that serve
both the EDA *and* the modelling stage, and would have reduced some
duplicated logic between Parts 2 and 3.

### 5.4  What would we improve with more time, knowledge, or resources?

With more time, knowledge, or data, the analysis could be extended in
several useful directions:

- **Add marketing budget data** (even an approximate figure) to separate
  the effect of production spending from promotional spending.
- **Adjust monetary values for inflation**, so that budgets and revenues
  from 1936 and 2018 are genuinely comparable, and consider adding
  release year as a covariate or fitting separate models by decade.
- **Allow multiple genre tags per film** instead of a single label, to
  reduce measurement error in the genre variable.
- **Use robust standard errors** (e.g. HC3) to make the significance
  tests more reliable in the presence of heteroscedasticity.
- **Try additional modelling approaches**, such as quantile regression
  to model the *spread* of ROI outcomes (not just the average), or a
  tree-based model (e.g. random forest) to explore non-linear
  interactions and rank predictor importance without assuming a log-log
  functional form in advance.

---

## 6  Project at a Glance

The table below summarises the project's headline numbers in one place.
The dataset-derived figures (rows, genres, years) are computed live from
the cleaned data so that they always stay in sync with Part 1; the
remaining rows describe the scope of the analytical work carried out in
Parts 2 and 3.

```{r project-at-a-glance}
glance_table <- data.frame(
	metric = c(
		"Films in raw dataset",
		"Films after cleaning (Part 1)",
		"Rows removed during cleaning",
		"Percent of rows removed",
		"Genres analysed",
		"Years covered (cleaned data)",
		"New derived columns created (Part 2)",
		"Charts created (Part 2)",
		"Statistical models built (Part 3)"
	),
	value = c(
		n_raw,
		n_clean,
		n_removed,
		paste0(pct_removed, "%"),
		paste(sort(unique(movie_profit_clean$genre)), collapse = ", "),
		paste0(year_min, "\u2013", year_max),
		"12 (minimum required: 10)",
		"13 (minimum required: 10)",
		"2 (main-effects and interaction linear regression)"
	)
)

knitr::kable(glance_table, caption = "Project at a glance")
```
