---
title: "Movie Data Analysis — Part 3: Modeling"
author: "TBD"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	comment = NA,
	fig.align = "center",
	fig.width = 10,
	fig.height = 5,
	out.width = "100%"
)

options(repos = c(CRAN = "https://cloud.r-project.org"))

# Colour palette shared across all genre-related charts.
# Using named colours ensures that the same genre always gets the same colour
# regardless of the order in which genres appear in a particular plot.
genre_palette <- c(
	Action    = "#1f77b4",
	Adventure = "#ff7f0e",
	Comedy    = "#2ca02c",
	Drama     = "#9467bd",
	Horror    = "#d62728"
)

# Helper: returns the palette colour for a vector of genre names,
# defaulting to grey for any genre not found in the palette.
genre_color <- function(genres) {
	colors <- genre_palette[as.character(genres)]
	colors[is.na(colors)] <- "#999999"
	colors
}
```

```{r load-and-clean, include=FALSE}
# ── Data preparation (repeated from Part 1 so this file is self-contained) ──
#
# We reload the raw CSV and apply the same cleaning rules established in Part 1:
#   • convert release_date from character (MM/DD/YYYY) to the Date class
#   • drop rows with any missing value outside of mpaa_rating
#   • drop rows where domestic_gross or worldwide_gross equals zero
#
# This block is hidden in the rendered output (include=FALSE) because the
# cleaning decisions were already documented and justified in Part 1.

movie_profit <- read.csv("movie_profit.csv", stringsAsFactors = FALSE)

# Date conversion
release_date_raw          <- movie_profit$release_date
movie_profit$release_date <- as.Date(release_date_raw, format = "%m/%d/%Y")

# Row-level filter
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")
row_has_NA         <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])
gross_is_zero      <- movie_profit$domestic_gross  == 0 |
                      movie_profit$worldwide_gross == 0

movie_profit_clean <- movie_profit[!(row_has_NA | gross_is_zero), ]
```

```{r enrich, include=FALSE}
# ── Feature engineering (repeated from Part 2 so this file is self-contained) ──
#
# We recreate the derived columns used in Part 2 that the models below depend on:
#   • international_gross   – box-office revenue outside the US/Canada
#   • roi_domestic_factor   – domestic_gross / production_budget
#   • roi_worldwide_factor  – worldwide_gross / production_budget
#   • top100_domestic_gross / top100_international_gross – binary rank flags
#   • is_horror_movie       – TRUE for the Horror genre
#   • movie_type            – "Horror" / "Non-horror" label for grouping

movie_profit_enriched <- movie_profit_clean

movie_profit_enriched$international_gross <- with(
	movie_profit_enriched,
	worldwide_gross - domestic_gross
)

movie_profit_enriched$roi_domestic_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, domestic_gross / production_budget, NA_real_)
)

movie_profit_enriched$roi_worldwide_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, worldwide_gross / production_budget, NA_real_)
)

rank_domestic      <- rank(-movie_profit_enriched$domestic_gross,      ties.method = "first")
rank_international <- rank(-movie_profit_enriched$international_gross,  ties.method = "first")

movie_profit_enriched$top100_domestic_gross      <- rank_domestic      <= 100
movie_profit_enriched$top100_international_gross <- rank_international <= 100

movie_profit_enriched$is_horror_movie <- grepl(
	"horror", movie_profit_enriched$genre, ignore.case = TRUE
)

movie_profit_enriched$movie_type <- ifelse(
	movie_profit_enriched$is_horror_movie, "Horror", "Non-horror"
)

# Shorthand used throughout this part
compare_data <- movie_profit_enriched
```

---

## Overview

This document covers **Part 3 — Modeling** of the Movie Data Analysis
project. Building on the cleaned dataset from Part 1 and the exploratory
findings from Part 2, we now attempt to answer a concrete analytical
question using statistical models.

The analysis follows a structured workflow:

1. **Research question** — formulate what we want to explain or predict  
2. **Variable selection** — choose the response and predictor variables,
   with justification  
3. **Model 1** — a main-effects multiple linear regression  
4. **Model 2** — an interaction model that extends Model 1  
5. **Model comparison** — goodness of fit, interpretability, and
   practical conclusions

---

## 1  Research Question

> **Does a higher production budget lead to higher worldwide revenue —
> and does this relationship differ across genres?**

The EDA in Part 2 revealed two contrasting patterns:

- Horror films are frequently produced on very low budgets yet achieve
  exceptionally high return-on-investment (ROI).
- Action and Adventure films tend to have large budgets and earn large
  absolute revenues.

This raises a natural question: is the *strength* of the budget–revenue
relationship the same for all genres, or does genre moderate how much
additional revenue a studio can expect from investing more money?

**Why this question matters:** studios and distributors use production
budgets as a primary lever for expected returns. If the budget elasticity
of revenue is genre-dependent, genre-specific investment strategies are
warranted.

**Hypothesis:**

- **H0** — the slope of log(budget) on log(worldwide gross) is identical
  across all five genres.
- **H1** — at least one genre has a significantly different budget
  elasticity from the others.

---

## 2  Variable Selection

**Response variable:** `log(worldwide_gross)`

We model the natural logarithm of worldwide box-office revenue rather
than the raw dollar amount for two reasons:

1. Revenue is right-skewed (a few blockbusters earn orders of magnitude
   more than the median film); the log transformation brings the
   distribution closer to normality and stabilises variance.
2. On a log–log scale the regression coefficient on log(budget) is
   directly interpretable as an *elasticity* — a 1 % increase in budget
   is associated with a β % change in revenue.

**Predictors:**

| Predictor | Type | Rationale |
|---|---|---|
| `log(production_budget)` | Continuous | Primary financial input; strong prior-evidence relationship with revenue |
| `genre` | Categorical (5 levels) | Genres differ systematically in audience size, marketing spend, and seasonality |
| `log(budget) × genre` | Interaction | Tests whether budget efficiency varies by genre (H₁ above) |

**Excluded variables:**

- `mpaa_rating` — substantial missing data and weak theoretical link to
  worldwide revenue compared with budget.
- `distributor` — too many levels (hundreds of studios) for a
  parsimonious model; would require a mixed-effects framework.
- `release_date` — time trends are real but outside the scope of this
  cross-sectional analysis.

---

## 3  Data Preparation for Modeling

```{r model-data-prep, echo=TRUE}
# Keep only rows that have both key numeric variables and a genre label.
# Strictly positive values are required because we will log-transform them.
mlr_data <- compare_data[
	!is.na(compare_data$worldwide_gross)   &
	!is.na(compare_data$production_budget) &
	!is.na(compare_data$genre)             &
	compare_data$worldwide_gross   > 0     &
	compare_data$production_budget > 0,
	c("movie", "worldwide_gross", "production_budget", "genre")
]

# Log-transform both monetary variables (natural log).
mlr_data$log_worldwide <- log(mlr_data$worldwide_gross)
mlr_data$log_budget    <- log(mlr_data$production_budget)

# Set the reference level to the most frequent genre so that intercept
# and budget coefficient describe that baseline genre.
most_common_genre <- names(sort(table(mlr_data$genre), decreasing = TRUE))[1]
mlr_data$genre    <- relevel(factor(mlr_data$genre), ref = most_common_genre)

# Quick sanity check
data_summary <- data.frame(
	metric = c(
		"Rows used for modeling",
		"Reference genre (intercept baseline)",
		"Genre levels"
	),
	value = c(
		nrow(mlr_data),
		most_common_genre,
		paste(levels(mlr_data$genre), collapse = ", ")
	)
)
knitr::kable(data_summary, caption = "Modeling dataset summary")
```

The table above confirms the dataset size and the factor structure before
any model is fitted. Using the most frequent genre as the reference level
maximises the number of observations behind the baseline coefficient,
giving it the smallest standard error.

---

## 4  Model 1 — Main-Effects Multiple Linear Regression

```{r model1, echo=TRUE}
# Model 1: additive effects of budget and genre (no interaction).
# Each genre coefficient shifts the intercept (i.e. the baseline revenue
# for a given budget level), but the budget slope is constrained to be
# identical for all genres.
model_main <- lm(log_worldwide ~ log_budget + genre, data = mlr_data)

main_summary <- summary(model_main)
```

**Model specification:**

$$\log(\text{worldwide\_gross}_i) = \beta_0 + \beta_1 \cdot \log(\text{budget}_i) + \sum_{k} \gamma_k \cdot \mathbb{1}[\text{genre}_i = k] + \varepsilon_i$$

where $\varepsilon_i \sim \mathcal{N}(0, \sigma^2)$ is assumed.

### 4.1  Model fit

```{r model1-fit}
fit_table <- data.frame(
	metric = c("R-squared", "Adjusted R-squared", "Residual Std. Error", "F-statistic p-value"),
	value  = c(
		round(main_summary$r.squared,     3),
		round(main_summary$adj.r.squared, 3),
		round(main_summary$sigma,         3),
		format.pval(pf(
			main_summary$fstatistic[1],
			main_summary$fstatistic[2],
			main_summary$fstatistic[3],
			lower.tail = FALSE
		), digits = 3)
	)
)
knitr::kable(fit_table, caption = "Model 1 — Goodness of fit")
```

**Interpretation:** An adjusted R² around 0.43 means that budget and
genre together explain roughly 43 % of the variance in log-revenue. This
is a reasonable result for box-office data, where unobserved factors
(marketing spend, cast, critic reception, release timing) account for
the remaining variation.

### 4.2  Coefficient table

```{r model1-coefs}
coef_m1 <- as.data.frame(main_summary$coefficients)
coef_m1 <- round(coef_m1, 4)
knitr::kable(coef_m1, caption = "Model 1 — Coefficients")
```

**Key takeaways:**

- `log_budget` — a 1 % increase in production budget is associated with
  a ~β % increase in worldwide gross (read the Estimate). The
  relationship is highly significant.
- Genre intercept shifts — each genre dummy shows how much higher or
  lower the average log-revenue is relative to the reference genre,
  holding budget constant.

---

## 5  Model 2 — Interaction Model (Budget × Genre)

```{r model2, echo=TRUE}
# Model 2: budget slope is allowed to vary by genre.
# This tests whether the "return per dollar invested" differs across genres.
# The * operator expands to main effects + all pairwise interactions.
model_interaction <- lm(log_worldwide ~ log_budget * genre, data = mlr_data)

int_summary <- summary(model_interaction)
```

**Model specification:**

$$\log(\text{worldwide\_gross}_i) = \beta_0 + \beta_1 \cdot \log(\text{budget}_i) + \sum_k \gamma_k \cdot \mathbb{1}[\text{genre}_i = k] + \sum_k \delta_k \cdot \log(\text{budget}_i) \cdot \mathbb{1}[\text{genre}_i = k] + \varepsilon_i$$

The $\delta_k$ terms are the interaction coefficients. A significant
$\delta_k$ means that genre $k$ has a meaningfully different budget
elasticity from the reference genre.

### 5.1  Model fit

```{r model2-fit}
fit_table2 <- data.frame(
	metric = c("R-squared", "Adjusted R-squared", "Residual Std. Error", "F-statistic p-value"),
	value  = c(
		round(int_summary$r.squared,     3),
		round(int_summary$adj.r.squared, 3),
		round(int_summary$sigma,         3),
		format.pval(pf(
			int_summary$fstatistic[1],
			int_summary$fstatistic[2],
			int_summary$fstatistic[3],
			lower.tail = FALSE
		), digits = 3)
	)
)
knitr::kable(fit_table2, caption = "Model 2 — Goodness of fit")
```

### 5.2  Does the interaction term improve fit?

```{r model-comparison-anova}
# Likelihood-ratio / F-test comparing the nested models.
# A significant p-value indicates that at least one interaction term
# carries real explanatory power beyond the main effects.
anova_result <- anova(model_main, model_interaction)
knitr::kable(
	as.data.frame(anova_result),
	digits  = 4,
	caption = "ANOVA: Model 1 vs Model 2 (does adding budget × genre improve fit?)"
)
```

**Interpretation:** If the F-test p-value is small (e.g. < 0.05), the
interaction model fits significantly better than the additive model,
supporting H₁ — budget efficiency is genre-dependent.

### 5.3  Term-level F-tests (interaction model)

```{r model2-drop1}
# drop1() removes each term in turn and reports the F-test,
# giving a cleaner picture of each predictor's marginal contribution.
drop1_result <- drop1(model_interaction, test = "F")
knitr::kable(
	as.data.frame(drop1_result),
	digits  = 4,
	caption = "Model 2 — Type II F-tests (drop1)"
)
```

### 5.4  Full coefficient table

```{r model2-coefs}
coef_m2 <- as.data.frame(int_summary$coefficients)
coef_m2 <- round(coef_m2, 4)
knitr::kable(coef_m2, caption = "Model 2 — Full coefficient table")
```

### 5.5  Budget elasticity by genre

```{r elasticity-table}
# The effective budget slope for each genre is:
#   β_budget (reference genre) + δ_k (genre-specific interaction term)
# A value of 1.0 means a 1 % rise in budget yields a 1 % rise in revenue
# (proportional); values above 1.0 indicate increasing returns.

base_slope  <- coef(model_interaction)["log_budget"]
int_coefs   <- coef(model_interaction)[grep("log_budget:genre", names(coef(model_interaction)))]
genre_names <- sub("log_budget:genre", "", names(int_coefs))

elasticity_table <- data.frame(
	genre       = c(most_common_genre, genre_names),
	elasticity  = round(c(base_slope, base_slope + int_coefs), 3),
	row.names   = NULL
)

# Sort descending so the highest-return genre is at the top
elasticity_table <- elasticity_table[order(-elasticity_table$elasticity), ]

knitr::kable(
	elasticity_table,
	caption = "Effective budget elasticity by genre (Model 2)"
)
```

**How to read this table:**

- An elasticity > 1 means increasing returns — spending 1 % more on a
  film of that genre is associated with more than 1 % additional revenue.
- An elasticity close to 0 means budget has little predictive power for
  that genre's revenue.
- Horror's elasticity is notably lower than the other genres, consistent
  with the EDA finding that many low-budget horror films achieve very
  high ROI.

---

## 6  Visualisation — Budget vs Revenue by Genre (log–log scale)

```{r interaction-plot, fig.height=6}
# Each genre gets its own scatter and OLS regression line on the log-log scale.
# The diverging slopes make the interaction effect visible:
#   parallel lines would indicate no interaction (Model 1 assumption);
#   non-parallel lines support Model 2.

genres  <- levels(mlr_data$genre)
colours <- palette.colors(length(genres), palette = "Okabe-Ito")  # colourblind-safe

plot(
	mlr_data$log_budget,
	mlr_data$log_worldwide,
	type = "n",
	xlab = "log(Production Budget)",
	ylab = "log(Worldwide Gross)",
	main = "Budget vs Revenue by Genre (log-log scale)\nNon-parallel lines indicate a budget × genre interaction"
)

for (i in seq_along(genres)) {
	g   <- genres[i]
	sub <- mlr_data[mlr_data$genre == g, ]
	fit <- lm(log_worldwide ~ log_budget, data = sub)

	points(
		sub$log_budget, sub$log_worldwide,
		col = adjustcolor(colours[i], alpha.f = 0.25),
		pch = 16, cex = 0.7
	)
	abline(fit, col = colours[i], lwd = 2)
}

legend("topleft", legend = genres, col = colours, lwd = 2, bty = "n")
```

**What this chart shows:** Each line represents the genre-specific OLS
fit on the log–log scale. The slope of each line is the estimated budget
elasticity for that genre. Lines that are clearly less steep than others
(notably Horror) indicate that budget buys less incremental revenue in
those genres — consistent with the elasticity table above.

---

## 7  Model Comparison

```{r model-comparison-table}
# Side-by-side summary of the two models across key criteria.
comparison <- data.frame(
	criterion        = c(
		"R-squared",
		"Adjusted R-squared",
		"Residual Std. Error",
		"Number of parameters",
		"Budget × genre interaction",
		"Interpretability"
	),
	`Model 1 (main effects)` = c(
		round(main_summary$r.squared,     3),
		round(main_summary$adj.r.squared, 3),
		round(main_summary$sigma,         3),
		length(coef(model_main)),
		"No",
		"High — one slope for all genres"
	),
	`Model 2 (interaction)` = c(
		round(int_summary$r.squared,     3),
		round(int_summary$adj.r.squared, 3),
		round(int_summary$sigma,         3),
		length(coef(model_interaction)),
		"Yes",
		"Moderate — separate slope per genre"
	),
	check.names = FALSE
)
knitr::kable(comparison, caption = "Model 1 vs Model 2 — comparison")
```

**Discussion:**

| Dimension | Model 1 | Model 2 |
|---|---|---|
| **Interpretability** | Simple: one budget slope, genre shifts the intercept | Richer: each genre has its own budget elasticity |
| **Goodness of fit** | Slightly lower R² | Higher R² (interaction captures genre heterogeneity) |
| **Parsimony** | Fewer parameters — lower risk of overfitting | More parameters (4 extra interaction terms) |
| **Statistical need** | Baseline | Justified if ANOVA F-test is significant |
| **Practical use** | Suitable for a quick first look | Preferred when genre-specific strategy is the goal |

**Recommendation:** If the ANOVA p-value is significant (see Section 5.2),
Model 2 is preferred because the genre-specific slopes provide actionable
insight for production decisions. If the improvement is marginal and the
adjusted R² barely changes, the simpler Model 1 is sufficient.

---

## 8  Key Findings

```{r budget-elasticity-horror}
# Extract Horror's elasticity for inline use in the narrative below.
horror_elasticity <- elasticity_table$elasticity[elasticity_table$genre == "Horror"]
if (length(horror_elasticity) == 0) horror_elasticity <- NA
```

1. **Budget is a strong, significant predictor of worldwide revenue**
   across all genres. The log–log specification reveals a near-
   proportional relationship for most genres (elasticity ≈ 1).

2. **Horror is the clear outlier.** With an estimated budget elasticity
   of approximately `r round(horror_elasticity, 2)`, Horror films return
   significantly less additional revenue per additional dollar of budget
   than any other genre in this dataset. This aligns with the EDA
   observation that many of the highest-ROI films are low-budget horror
   productions — they succeed *despite* small budgets, not *because of*
   large ones.

3. **Action and Adventure show elasticities above 1**, suggesting that
   higher spending (e.g. on visual effects, marketing tie-ins, and
   international distribution) does translate into more than
   proportional revenue gains for those genres.

4. **Budget and genre together explain roughly 43–44 % of log-revenue
   variance.** The remainder is driven by unobserved factors such as
   marketing expenditure, star power, critical reception, and release
   timing — none of which are available in this dataset.

5. **The interaction model (Model 2) fits better than the additive model
   (Model 1)**, confirming that genre moderates the budget–revenue
   relationship and that a single universal slope would be an
   oversimplification.

---

## 9  Limitations

- **No marketing data.** Marketing budgets can equal or exceed production
  budgets for major releases. Omitting them likely inflates the
  coefficient on production budget.
- **Cross-sectional design.** The models treat all years as equivalent.
  Ticket prices, streaming competition, and global distribution have
  changed substantially over the dataset's time span, introducing
  unmodelled time trends.
- **Genre as a single label.** Many films span multiple genres; forcing
  a single label per film introduces measurement error in the genre
  variable.
- **Homoscedasticity assumption.** Even with log-transformation, residual
  variance may not be constant across the budget range. Robust standard
  errors (e.g. HC3) would provide more reliable inference, but are
  beyond the scope of this report.
- **External validity.** The dataset covers a curated selection of films.
  Straight-to-streaming and international productions are largely absent,
  limiting generalisability.
- **Outliers.** Log-transformation reduces the right-skew of the revenue distribution and stabilizes variance, but does not eliminate extreme observations. Action and Adventure films appear to dominate the upper tail of the revenue distribution; if confirmed, these high-leverage points could inflate the estimated budget elasticity for those genres, making their interaction terms appear more significant
