---
title: "Movie Data Analysis"
author: "TBD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	comment = NA,
	fig.align = "center",
	fig.width = 10,
	fig.height = 5,
	out.width = "100%"
)
```

## Load and verify the data

```{r}
if (!requireNamespace("DT", quietly = TRUE)) {
	stop("Package 'DT' is required. Install it once with: install.packages('DT', repos = 'https://cloud.r-project.org')")
}

movie_profit <- read.csv("movie_profit.csv", stringsAsFactors = FALSE)

overview_table <- data.frame(
	Metric = c("Rows", "Columns", "Numeric columns", "Character columns"),
	Value = c(
		nrow(movie_profit),
		ncol(movie_profit),
		sum(sapply(movie_profit, is.numeric)),
		sum(sapply(movie_profit, is.character))
	)
)

knitr::kable(overview_table, caption = "Dataset overview")
```


## Check variable types and first 5 values

```{r}
variable_overview <- data.frame(
	variable = names(movie_profit),
	type = sapply(movie_profit, function(x) class(x)[1]),
	first_5_values = sapply(movie_profit, function(x) {
		paste(head(x, 5), collapse = ", ")
	}),
	row.names = NULL
)

knitr::kable(variable_overview, caption = "Variable types and first 5 values")
```

## Convert release_date to Date type and check conversion results

```{r}
release_date_raw <- movie_profit$release_date
movie_profit$release_date <- as.Date(release_date_raw, format = "%m/%d/%Y")

conversion_table <- data.frame(
	metric = c("Original type", "New type", "Invalid dates after conversion"),
	value = c(
		class(release_date_raw)[1],
		class(movie_profit$release_date)[1],
		sum(is.na(movie_profit$release_date) & !is.na(release_date_raw) & trimws(release_date_raw) != "")
	)
)

knitr::kable(conversion_table, caption = "release_date conversion summary")
```

## Check for missing values, duplicates, and logical consistency

```{r}
# Missing values (NA or blank strings)
missing_count <- sapply(movie_profit, function(x) {
	if (is.character(x)) {
		sum(is.na(x) | trimws(x) == "")
	} else {
		sum(is.na(x))
	}
})

missing_pct <- round(100 * missing_count / nrow(movie_profit), 2)
missing_summary <- data.frame(
	variable = names(missing_count),
	missing_count = as.integer(missing_count),
	missing_percent = missing_pct,
	row.names = NULL
)

# Duplicate rows
dup_rows <- duplicated(movie_profit)
dup_count <- sum(dup_rows)

# Numeric column checks
numeric_cols <- c("production_budget", "domestic_gross", "worldwide_gross")
numeric_checks <- data.frame()
for (col in numeric_cols) {
	if (col %in% names(movie_profit)) {
		x <- movie_profit[[col]]
		numeric_checks <- rbind(
			numeric_checks,
			data.frame(
				variable = col,
				negative_values = sum(x < 0, na.rm = TRUE),
				zero_values = sum(x == 0, na.rm = TRUE)
			)
		)
	}
}

# Logical consistency check
bad_gross_count <- NA
if (all(c("domestic_gross", "worldwide_gross") %in% names(movie_profit))) {
	bad_gross_rows <- which(movie_profit$worldwide_gross < movie_profit$domestic_gross)
	bad_gross_count <- length(bad_gross_rows)
}

# Date parsing check
bad_date_count <- sum(
	is.na(movie_profit$release_date) &
		!is.na(release_date_raw) &
		trimws(release_date_raw) != ""
)

knitr::kable(missing_summary, caption = "Missing values by variable")

dup_table <- data.frame(
	metric = "Duplicate rows",
	count = dup_count
)
knitr::kable(dup_table, caption = "Duplicate row summary")

knitr::kable(numeric_checks, caption = "Numeric consistency checks")

logic_table <- data.frame(
	check = c("Rows where worldwide_gross < domestic_gross", "Invalid release_date values"),
	count = c(bad_gross_count, bad_date_count)
)
knitr::kable(logic_table, caption = "Logical consistency checks")
```

## Estimate impact of removing rows with missing values or gross = 0

```{r}
# Row-level missing values (NA or blank strings) across all columns
row_has_missing <- apply(movie_profit, 1, function(row) {
	any(is.na(row) | trimws(as.character(row)) == "")
})

# Rows where either gross column is exactly zero
gross_is_zero <-
	movie_profit$domestic_gross == 0 |
	movie_profit$worldwide_gross == 0

overlap_count <- sum(row_has_missing & gross_is_zero)
remove_all_count <- sum(row_has_missing | gross_is_zero)
total_movies <- nrow(movie_profit)
remove_all_pct <- round(100 * remove_all_count / total_movies, 2)

# Same calculation, but ignore missing values in mpaa_rating
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")
row_has_missing_no_rating <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])

remove_no_rating_count <- sum(row_has_missing_no_rating | gross_is_zero)
remove_no_rating_pct <- round(100 * remove_no_rating_count / total_movies, 2)

impact_table <- data.frame(
	metric = c(
		"Rows with at least one missing value",
		"Rows with gross value 0 (domestic or worldwide)",
		"Overlap (missing and gross = 0)",
		"Rows removed (missing or gross = 0)",
		"Rows removed (ignoring missing mpaa_rating)"
	),
	count = c(
		sum(row_has_missing),
		sum(gross_is_zero),
		overlap_count,
		remove_all_count,
		remove_no_rating_count
	),
	percent = c(
		round(100 * sum(row_has_missing) / total_movies, 2),
		round(100 * sum(gross_is_zero) / total_movies, 2),
		round(100 * overlap_count / total_movies, 2),
		remove_all_pct,
		remove_no_rating_pct
	)
)

knitr::kable(impact_table, caption = "Impact of filtering rules")
```

## Remove rows with missing values (except rating) and gross = 0

```{r}
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")

row_has_NA <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])

gross_is_zero <-
	movie_profit$domestic_gross == 0 |
	movie_profit$worldwide_gross == 0

movie_profit_clean <- movie_profit[!(row_has_NA | gross_is_zero), ]

cleaning_table <- data.frame(
	metric = c("Rows before cleaning", "Rows after cleaning", "Rows removed"),
	count = c(nrow(movie_profit), nrow(movie_profit_clean), nrow(movie_profit) - nrow(movie_profit_clean))
)

knitr::kable(cleaning_table, caption = "Cleaning summary")
```

## Create new derived columns
New columns:
- International Gross (numerical)
- ROI domestic (numerical)
- ROI worldwide (numerical)
- Top-100 domestic gross flag (boolean)
- Top-100 international gross flag (boolean)
- Is Horror Movie (logical flag)

```{r}
movie_profit_enriched <- movie_profit_clean

# International gross is the non-domestic part of worldwide gross
movie_profit_enriched$international_gross <- movie_profit_enriched$worldwide_gross - movie_profit_enriched$domestic_gross

# ROI factors (gross divided by production budget)
movie_profit_enriched$roi_domestic_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, domestic_gross / production_budget, NA_real_)
)

movie_profit_enriched$roi_worldwide_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, worldwide_gross / production_budget, NA_real_)
)

# Top-100 flags
rank_domestic <- rank(-movie_profit_enriched$domestic_gross, ties.method = "first")
rank_international <- rank(-movie_profit_enriched$international_gross, ties.method = "first")

movie_profit_enriched$top100_domestic_gross <- rank_domestic <= 100
movie_profit_enriched$top100_international_gross <- rank_international <= 100

# Logical flag for horror movies based on genre text
movie_profit_enriched$is_horror_movie <-
	grepl("horror", movie_profit_enriched$genre, ignore.case = TRUE)

new_columns_preview <- movie_profit_enriched[, c(
	"movie",
	"production_budget",
	"domestic_gross",
	"worldwide_gross",
	"international_gross",
	"roi_domestic_factor",
	"roi_worldwide_factor",
	"top100_domestic_gross",
	"top100_international_gross",
	"is_horror_movie"
)]

knitr::kable(head(new_columns_preview, 5), caption = "Preview of new derived columns")
```

## Basic descriptive statistics

```{r}
analysis_data <- movie_profit_enriched

numeric_vars <- names(analysis_data)[sapply(analysis_data, is.numeric)]
# Exclude row ID from numeric summaries
numeric_vars <- setdiff(numeric_vars, "X")

# Focus categorical summaries on meaningful analysis columns
categorical_vars <- intersect(c("genre", "mpaa_rating", "distributor"), names(analysis_data))

release_dates <- analysis_data$release_date
profile_table <- data.frame(
	metric = c(
		"Rows",
		"Columns",
		"Numeric variables",
		"Categorical variables",
		"Unique movies",
		"Valid release dates",
		"Release year range"
	),
	value = c(
		nrow(analysis_data),
		ncol(analysis_data),
		length(numeric_vars),
		length(categorical_vars),
		length(unique(analysis_data$movie)),
		length(release_dates),
		paste(format(min(release_dates), "%Y"), "to", format(max(release_dates), "%Y"))
	)
)
knitr::kable(profile_table, caption = "Analysis dataset profile")

if (length(numeric_vars) > 0) {
	num_summary_mat <- do.call(cbind, lapply(analysis_data[, numeric_vars, drop = FALSE], summary))
	num_summary <- data.frame(
		statistic = rownames(num_summary_mat),
		num_summary_mat,
		row.names = NULL,
		check.names = FALSE
	)

	for (col in numeric_vars) {
		num_summary[[col]] <- format(
			round(as.numeric(num_summary[[col]]), 2),
			big.mark = ",",
			scientific = FALSE,
			trim = TRUE
		)
	}

	print(knitr::kable(num_summary, caption = "Numeric summary"))

	mean_median_table <- data.frame(
		variable = numeric_vars,
		mean = round(sapply(analysis_data[, numeric_vars, drop = FALSE], mean, na.rm = TRUE), 2),
		median = round(sapply(analysis_data[, numeric_vars, drop = FALSE], median, na.rm = TRUE), 2),
		row.names = NULL
	)
	mean_median_table$mean <- format(mean_median_table$mean, big.mark = ",", scientific = FALSE, trim = TRUE)
	mean_median_table$median <- format(mean_median_table$median, big.mark = ",", scientific = FALSE, trim = TRUE)
	print(knitr::kable(mean_median_table, caption = "Mean and median by numeric variable"))

	sd_table <- data.frame(
		variable = numeric_vars,
		sd = round(sapply(analysis_data[, numeric_vars, drop = FALSE], sd, na.rm = TRUE), 2),
		row.names = NULL
	)
 sd_table$sd <- format(sd_table$sd, big.mark = ",", scientific = FALSE, trim = TRUE)
	print(knitr::kable(sd_table, caption = "Standard deviation by variable"))

	outlier_stats <- do.call(rbind, lapply(numeric_vars, function(v) {
		x <- analysis_data[[v]]
		q1 <- quantile(x, 0.25, na.rm = TRUE)
		q3 <- quantile(x, 0.75, na.rm = TRUE)
		iqr_val <- q3 - q1
		lower <- q1 - 1.5 * iqr_val
		upper <- q3 + 1.5 * iqr_val
		low_outlier <- x < lower
		high_outlier <- x > upper
		is_outlier <- low_outlier | high_outlier

		data.frame(
			variable = v,
			outlier_count = sum(is_outlier, na.rm = TRUE),
			outlier_percent = round(100 * sum(is_outlier, na.rm = TRUE) / sum(!is.na(x)), 2),
			low_outlier_count = sum(low_outlier, na.rm = TRUE),
			high_outlier_count = sum(high_outlier, na.rm = TRUE)
		)
	}))

	print(knitr::kable(outlier_stats, caption = "IQR outlier summary"))
}


if (length(categorical_vars) > 0) {
	cat("\nCategorical variable details:\n")
	cat_table <- do.call(rbind, lapply(categorical_vars, function(v) {
		x <- analysis_data[[v]]
		x_non_missing <- x[!(is.na(x) | trimws(x) == "")]
		tab <- sort(table(x_non_missing), decreasing = TRUE)
		data.frame(
			variable = v,
			unique_levels = length(unique(x_non_missing)),
			most_common = names(tab)[1],
			most_common_count = as.integer(tab[1])
		)
	}))
	print(knitr::kable(cat_table, caption = "Categorical variable summary"))
}
```

## Simple distributions

```{r, include=FALSE}
analysis_data <- if (exists("movie_profit_clean")) movie_profit_clean else movie_profit
```

```{r}
if ("genre" %in% names(analysis_data)) {
	genre_tab <- sort(table(analysis_data$genre), decreasing = TRUE)
	top_n <- min(10, length(genre_tab))
	barplot(
		genre_tab[1:top_n],
		las = 2,
		col = "tan",
		main = "Top Genres",
		ylab = "Number of Movies"
	)
}
```

```{r}
if ("mpaa_rating" %in% names(analysis_data)) {
	rating_tab <- sort(table(analysis_data$mpaa_rating), decreasing = TRUE)
	barplot(
		rating_tab,
		col = "skyblue",
		main = "MPAA Rating Distribution",
		ylab = "Number of Movies"
	)
}
```

```{r}
release_dates <- analysis_data$release_date
release_year <- as.integer(format(release_dates, "%Y"))
year_tab <- table(release_year)
year_tab <- year_tab[order(as.integer(names(year_tab)))]
if (length(year_tab) > 0) {
	barplot(
		as.integer(year_tab),
		names.arg = names(year_tab),
		col = "darkorange",
		border = "white",
		las = 2,
		cex.names = 0.7,
		xlab = "Release Year",
		ylab = "Number of Movies",
		main = "Movies Released Per Year"
	)
}
```

```{r}
if ("distributor" %in% names(analysis_data)) {
	distributor_tab <- sort(table(analysis_data$distributor), decreasing = TRUE)
	top_n <- min(10, length(distributor_tab))
	barplot(
		distributor_tab[1:top_n],
		las = 2,
		cex.names = 0.8,
		col = "cornflowerblue",
		main = "Top 10 Distributors",
		ylab = "Number of Movies"
	)
}
```

## Horror vs non-horror comparison

```{r}
compare_data <- movie_profit_enriched

compare_data$movie_type <- ifelse(compare_data$is_horror_movie, "Horror", "Non-horror")

count_iqr_outliers <- function(x) {
	x <- x[!is.na(x)]
	if (length(x) < 4) {
		return(0)
	}
	q1 <- quantile(x, 0.25, na.rm = TRUE)
	q3 <- quantile(x, 0.75, na.rm = TRUE)
	iqr_val <- q3 - q1
	lower <- q1 - 1.5 * iqr_val
	upper <- q3 + 1.5 * iqr_val
	sum(x < lower | x > upper)
}

summary_table <- data.frame(
	metric = c(
		"movie_count",
		"budget_mean",
		"budget_q1",
		"budget_median",
		"budget_q3",
		"worldwide_mean",
		"worldwide_q1",
		"worldwide_median",
		"worldwide_q3",
		"roi_domestic_mean",
		"roi_domestic_q1",
		"roi_domestic_median",
		"roi_domestic_q3",
		"roi_domestic_outlier_count",
		"roi_worldwide_mean",
		"roi_worldwide_q1",
		"roi_worldwide_median",
		"roi_worldwide_q3",
		"roi_worldwide_outlier_count"
	),
	Horror = c(
		sum(compare_data$is_horror_movie),
		round(mean(compare_data$production_budget[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$production_budget[compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$production_budget[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$production_budget[compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		round(mean(compare_data$worldwide_gross[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$worldwide_gross[compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$worldwide_gross[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$worldwide_gross[compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		round(mean(compare_data$roi_domestic_factor[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_domestic_factor[compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$roi_domestic_factor[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_domestic_factor[compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		count_iqr_outliers(compare_data$roi_domestic_factor[compare_data$is_horror_movie]),
		round(mean(compare_data$roi_worldwide_factor[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_worldwide_factor[compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$roi_worldwide_factor[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_worldwide_factor[compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		count_iqr_outliers(compare_data$roi_worldwide_factor[compare_data$is_horror_movie])
	),
	`Non-horror` = c(
		sum(!compare_data$is_horror_movie),
		round(mean(compare_data$production_budget[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$production_budget[!compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$production_budget[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$production_budget[!compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		round(mean(compare_data$worldwide_gross[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$worldwide_gross[!compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$worldwide_gross[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$worldwide_gross[!compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		round(mean(compare_data$roi_domestic_factor[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_domestic_factor[!compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$roi_domestic_factor[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_domestic_factor[!compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		count_iqr_outliers(compare_data$roi_domestic_factor[!compare_data$is_horror_movie]),
		round(mean(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		count_iqr_outliers(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie])
	),
	check.names = FALSE
)

numeric_cols <- setdiff(names(summary_table), "metric")
for (col in numeric_cols) {
	summary_table[[col]] <- format(summary_table[[col]], big.mark = ",", scientific = FALSE, trim = TRUE)
}

knitr::kable(summary_table, caption = "Horror vs non-horror basic metrics")
```

## Horror vs non-horror movies over time

```{r}
year_vals <- as.integer(format(compare_data$release_date, "%Y"))
valid_rows <- !is.na(year_vals) & year_vals >= 1975
year_range <- range(year_vals[valid_rows], na.rm = TRUE)

horror_year_tab <- table(year_vals[valid_rows & compare_data$is_horror_movie])
horror_year_tab <- horror_year_tab[order(as.integer(names(horror_year_tab)))]

if (length(horror_year_tab) > 0) {
	plot(
		as.integer(names(horror_year_tab)),
		as.integer(horror_year_tab),
		type = "l",
		lwd = 2,
		col = "firebrick",
		xlim = year_range,
		xlab = "Release Year",
		ylab = "Number of Horror Movies",
		main = "Horror Movies per Year"
	)
}
```

```{r}
year_vals <- as.integer(format(compare_data$release_date, "%Y"))
valid_rows <- !is.na(year_vals) & year_vals >= 1975
year_range <- range(year_vals[valid_rows], na.rm = TRUE)

non_horror_year_tab <- table(year_vals[valid_rows & !compare_data$is_horror_movie])
non_horror_year_tab <- non_horror_year_tab[order(as.integer(names(non_horror_year_tab)))]

if (length(non_horror_year_tab) > 0) {
	plot(
		as.integer(names(non_horror_year_tab)),
		as.integer(non_horror_year_tab),
		type = "l",
		lwd = 2,
		col = "steelblue",
		xlim = year_range,
		xlab = "Release Year",
		ylab = "Number of Non-horror Movies",
		main = "Non-horror Movies per Year"
	)
}
```

## Average ROI by year: horror vs non-horror

```{r}
year_vals <- as.integer(format(compare_data$release_date, "%Y"))
roi_vals <- compare_data$roi_worldwide_factor

valid_rows <- !is.na(year_vals) & !is.na(roi_vals)
all_years <- sort(unique(year_vals[valid_rows]))
all_years <- all_years[all_years >= 1975]

horror_roi_by_year <- tapply(
	roi_vals[valid_rows & compare_data$is_horror_movie],
	year_vals[valid_rows & compare_data$is_horror_movie],
	mean
)

non_horror_roi_by_year <- tapply(
	roi_vals[valid_rows & !compare_data$is_horror_movie],
	year_vals[valid_rows & !compare_data$is_horror_movie],
	mean
)

horror_series <- as.numeric(horror_roi_by_year[as.character(all_years)])
non_horror_series <- as.numeric(non_horror_roi_by_year[as.character(all_years)])

if (length(all_years) > 0) {
	y_cap <- 20
	horror_plot_series <- pmin(horror_series, y_cap)
	non_horror_plot_series <- pmin(non_horror_series, y_cap)

	horror_outlier_idx <- which(!is.na(horror_series) & horror_series > y_cap)
	non_horror_outlier_idx <- which(!is.na(non_horror_series) & non_horror_series > y_cap)

	plot(
		all_years,
		horror_plot_series,
		type = "l",
		lwd = 2,
		col = "firebrick",
		xlab = "Release Year",
		ylab = "Average ROI (Worldwide)",
		main = "Average ROI per Year: Horror vs Non-horror (capped at 20)",
		ylim = c(0, y_cap)
	)

	lines(all_years, non_horror_plot_series, lwd = 2, col = "steelblue")

	if (length(horror_outlier_idx) > 0) {
		points(all_years[horror_outlier_idx], rep(y_cap, length(horror_outlier_idx)),
			pch = 24, bg = "firebrick", col = "firebrick", cex = 0.8)
	}

	if (length(non_horror_outlier_idx) > 0) {
		points(all_years[non_horror_outlier_idx], rep(y_cap, length(non_horror_outlier_idx)),
			pch = 24, bg = "steelblue", col = "steelblue", cex = 0.8)
	}

	legend("topright",
		legend = c("Horror", "Non-horror", paste0("Capped value (> ", y_cap, ")")),
		col = c("firebrick", "steelblue", "black"),
		lwd = c(2, 2, NA),
		pch = c(NA, NA, 24),
		pt.bg = c(NA, NA, "gray60"),
		bty = "n"
	)

	horror_outlier_note <- if (length(horror_outlier_idx) > 0) {
		paste(
			paste0(
				all_years[horror_outlier_idx],
				" (",
				round(horror_series[horror_outlier_idx], 2),
				"x)"
			),
			collapse = ", "
		)
	} else {
		"none"
	}

	non_horror_outlier_note <- if (length(non_horror_outlier_idx) > 0) {
		paste(
			paste0(
				all_years[non_horror_outlier_idx],
				" (",
				round(non_horror_series[non_horror_outlier_idx], 2),
				"x)"
			),
			collapse = ", "
		)
	} else {
		"none"
	}

	cat(paste0("Outlier years (mean ROI > ", y_cap, "):\n"))
	cat("- Horror:", horror_outlier_note, "\n")
	cat("- Non-horror:", non_horror_outlier_note, "\n")
}
```


## Median ROI by year: horror vs non-horror

```{r}
year_vals <- as.integer(format(compare_data$release_date, "%Y"))
roi_vals <- compare_data$roi_worldwide_factor

valid_rows <- !is.na(year_vals) & !is.na(roi_vals)
all_years <- sort(unique(year_vals[valid_rows]))
all_years <- all_years[all_years >= 1975]

horror_roi_by_year <- tapply(
	roi_vals[valid_rows & compare_data$is_horror_movie],
	year_vals[valid_rows & compare_data$is_horror_movie],
	median
)

non_horror_roi_by_year <- tapply(
	roi_vals[valid_rows & !compare_data$is_horror_movie],
	year_vals[valid_rows & !compare_data$is_horror_movie],
	median
)

horror_series <- as.numeric(horror_roi_by_year[as.character(all_years)])
non_horror_series <- as.numeric(non_horror_roi_by_year[as.character(all_years)])

if (length(all_years) > 0) {
	y_cap <- 20
	horror_plot_series <- pmin(horror_series, y_cap)
	non_horror_plot_series <- pmin(non_horror_series, y_cap)

	horror_outlier_idx <- which(!is.na(horror_series) & horror_series > y_cap)
	non_horror_outlier_idx <- which(!is.na(non_horror_series) & non_horror_series > y_cap)

	plot(
		all_years,
		horror_plot_series,
		type = "l",
		lwd = 2,
		col = "firebrick",
		xlab = "Release Year",
		ylab = "Median ROI (Worldwide)",
		main = "Median ROI per Year: Horror vs Non-horror (capped at 20)",
		ylim = c(0, y_cap)
	)

	lines(all_years, non_horror_plot_series, lwd = 2, col = "steelblue")

	if (length(horror_outlier_idx) > 0) {
		points(all_years[horror_outlier_idx], rep(y_cap, length(horror_outlier_idx)),
			pch = 24, bg = "firebrick", col = "firebrick", cex = 0.8)
	}

	if (length(non_horror_outlier_idx) > 0) {
		points(all_years[non_horror_outlier_idx], rep(y_cap, length(non_horror_outlier_idx)),
			pch = 24, bg = "steelblue", col = "steelblue", cex = 0.8)
	}

	legend("topright",
		legend = c("Horror", "Non-horror", paste0("Capped value (> ", y_cap, ")")),
		col = c("firebrick", "steelblue", "black"),
		lwd = c(2, 2, NA),
		pch = c(NA, NA, 24),
		pt.bg = c(NA, NA, "gray60"),
		bty = "n"
	)

	horror_outlier_note <- if (length(horror_outlier_idx) > 0) {
		paste(
			paste0(
				all_years[horror_outlier_idx],
				" (",
				round(horror_series[horror_outlier_idx], 2),
				"x)"
			),
			collapse = ", "
		)
	} else {
		"none"
	}

	non_horror_outlier_note <- if (length(non_horror_outlier_idx) > 0) {
		paste(
			paste0(
				all_years[non_horror_outlier_idx],
				" (",
				round(non_horror_series[non_horror_outlier_idx], 2),
				"x)"
			),
			collapse = ", "
		)
	} else {
		"none"
	}

	cat(paste0("Outlier years (median ROI > ", y_cap, "):\n"))
	cat("- Horror:", horror_outlier_note, "\n")
	cat("- Non-horror:", non_horror_outlier_note, "\n")
}
```

## Interactive movie table (filter + sort)

```{r}
table_data <- if (exists("movie_profit_enriched")) movie_profit_enriched else movie_profit

interactive_table <- table_data

if (!"roi_worldwide_factor" %in% names(interactive_table)) {
	interactive_table$roi_worldwide_factor <- with(
		interactive_table,
		ifelse(production_budget > 0, worldwide_gross / production_budget, NA_real_)
	)
}

# Keep numeric values numeric so sorting remains correct in DT.
interactive_table$production_budget_k <- interactive_table$production_budget / 1000
interactive_table$domestic_gross_k <- interactive_table$domestic_gross / 1000
interactive_table$worldwide_gross_k <- interactive_table$worldwide_gross / 1000

interactive_table$release_date_fmt <- ifelse(
	is.na(interactive_table$release_date),
	NA_character_,
	format(interactive_table$release_date, "%Y-%m-%d")
)

display_table <- interactive_table[, c(
	"movie",
	"genre",
	"roi_worldwide_factor",
	"production_budget_k",
	"domestic_gross_k",
	"worldwide_gross_k",
	"release_date_fmt",
	"mpaa_rating",
	"distributor"
)]

names(display_table) <- c(
	"Movie",
	"Genre",
	"ROI (x)",
	"Production Cost (x1000 $)",
	"Domestic Gross (x1000 $)",
	"Worldwide Gross (x1000 $)",
	"Release Date",
	"MPAA Rating",
	"Distributor"
)

display_table$`ROI (x)` <- round(display_table$`ROI (x)`, 2)
display_table$`Production Cost (x1000 $)` <- round(display_table$`Production Cost (x1000 $)`, 1)
display_table$`Domestic Gross (x1000 $)` <- round(display_table$`Domestic Gross (x1000 $)`, 1)
display_table$`Worldwide Gross (x1000 $)` <- round(display_table$`Worldwide Gross (x1000 $)`, 1)

DT::datatable(
	display_table,
	rownames = FALSE,
	options = list(
		pageLength = 15,
		lengthMenu = c(10, 15, 25, 50),
		order = list(list(5, "desc")),
		autoWidth = TRUE
	),
	callback = DT::JS(
		"var genreColumn = 1;",
		"var wrapper = $(table.table().container());",
		"var controls = $('<div style=\"margin-bottom: 8px;\"></div>');",
		"var label = $('<label style=\"margin-right: 8px; font-weight: 600;\">Genre:</label>');",
		"var select = $('<select style=\"padding: 2px 6px;\"><option value=\"\">All genres</option></select>');",
		"select.on('change', function() {",
		"  var val = $.fn.dataTable.util.escapeRegex($(this).val());",
		"  table.column(genreColumn).search(val ? '^' + val + '$' : '', true, false).draw();",
		"});",
		"table.column(genreColumn).data().unique().sort().each(function(d) {",
		"  if (d !== null && d !== undefined && d !== '') {",
		"    select.append($('<option></option>').attr('value', d).text(d));",
		"  }",
		"});",
		"controls.append(label).append(select);",
		"wrapper.prepend(controls);"
	),
	caption = "Interactive movie table (filter by Genre in the header row; money columns are in thousands of dollars)"
)
```

Some things we can learn from this table:
- 5/10 movies with ROI > 100x are horror movies (including the top 2).
- Except for some outliers (Exorcist 2 and The Shining), the earlierst horror movies (before 1982) all had a ROI > 10x, while the non-horror movies in that period had lower ROIs on average.
- When sorting after the lowest production budget, we can see that Horror Movies are cheap to produce (20 movies cost <= $1M). For other Genres, the numbers are: Action (4), Adventure (2), Comedy (28), Drama (69).
- When sorting for production budget, we can also see that the cheapest Horror movies (<= $1M) are the movies which end up having a ROI > 100x (5/20 movies), while most movies have a ROI of >= 8x (15/20 movies). At the same time, in other genres the cheapest movies have a way lower ROI on average, with Drama only reaching an ROI >= 5x in half of the movies with a production budget <= $1M.
