---
title: "Movie Data Analysis"
author: "TBD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	comment = NA,
	fig.align = "center",
	fig.width = 10,
	fig.height = 5,
	out.width = "100%"
)

options(repos = c(CRAN = "https://cloud.r-project.org"))

genre_palette <- c(
	Action = "#1f77b4",
	Adventure = "#ff7f0e",
	Comedy = "#2ca02c",
	Drama = "#9467bd",
	Horror = "#d62728"
)

genre_color <- function(genres) {
	colors <- genre_palette[as.character(genres)]
	colors[is.na(colors)] <- "#999999"
	colors
}
```

## Load and verify the data

```{r}
if (!requireNamespace("DT", quietly = TRUE)) {
	stop("Package 'DT' is required. Install it once with: install.packages('DT', repos = 'https://cloud.r-project.org')")
}

movie_profit <- read.csv("movie_profit.csv", stringsAsFactors = FALSE)

overview_table <- data.frame(
	Metric = c("Rows", "Columns", "Numeric columns", "Character columns"),
	Value = c(
		nrow(movie_profit),
		ncol(movie_profit),
		sum(sapply(movie_profit, is.numeric)),
		sum(sapply(movie_profit, is.character))
	)
)

knitr::kable(overview_table, caption = "Dataset overview")
```


## Check variable types and first 5 values

```{r}
variable_overview <- data.frame(
	variable = names(movie_profit),
	type = sapply(movie_profit, function(x) class(x)[1]),
	first_5_values = sapply(movie_profit, function(x) {
		paste(head(x, 5), collapse = ", ")
	}),
	row.names = NULL
)

knitr::kable(variable_overview, caption = "Variable types and first 5 values")
```

## Convert release_date to Date type and check conversion results

```{r}
release_date_raw <- movie_profit$release_date
movie_profit$release_date <- as.Date(release_date_raw, format = "%m/%d/%Y")

conversion_table <- data.frame(
	metric = c("Original type", "New type", "Invalid dates after conversion"),
	value = c(
		class(release_date_raw)[1],
		class(movie_profit$release_date)[1],
		sum(is.na(movie_profit$release_date) & !is.na(release_date_raw) & trimws(release_date_raw) != "")
	)
)

knitr::kable(conversion_table, caption = "release_date conversion summary")
```

## Check for missing values, duplicates, and logical consistency

```{r}
# Missing values (NA or blank strings)
missing_count <- sapply(movie_profit, function(x) {
	if (is.character(x)) {
		sum(is.na(x) | trimws(x) == "")
	} else {
		sum(is.na(x))
	}
})

missing_pct <- round(100 * missing_count / nrow(movie_profit), 2)
missing_summary <- data.frame(
	variable = names(missing_count),
	missing_count = as.integer(missing_count),
	missing_percent = missing_pct,
	row.names = NULL
)

# Duplicate rows
dup_rows <- duplicated(movie_profit)
dup_count <- sum(dup_rows)

# Numeric column checks
numeric_cols <- c("production_budget", "domestic_gross", "worldwide_gross")
numeric_checks <- data.frame()
for (col in numeric_cols) {
	if (col %in% names(movie_profit)) {
		x <- movie_profit[[col]]
		numeric_checks <- rbind(
			numeric_checks,
			data.frame(
				variable = col,
				negative_values = sum(x < 0, na.rm = TRUE),
				zero_values = sum(x == 0, na.rm = TRUE)
			)
		)
	}
}

# Logical consistency check
bad_gross_count <- NA
if (all(c("domestic_gross", "worldwide_gross") %in% names(movie_profit))) {
	bad_gross_rows <- which(movie_profit$worldwide_gross < movie_profit$domestic_gross)
	bad_gross_count <- length(bad_gross_rows)
}

# Date parsing check
bad_date_count <- sum(
	is.na(movie_profit$release_date) &
		!is.na(release_date_raw) &
		trimws(release_date_raw) != ""
)

knitr::kable(missing_summary, caption = "Missing values by variable")

dup_table <- data.frame(
	metric = "Duplicate rows",
	count = dup_count
)
knitr::kable(dup_table, caption = "Duplicate row summary")

knitr::kable(numeric_checks, caption = "Numeric consistency checks")

logic_table <- data.frame(
	check = c("Rows where worldwide_gross < domestic_gross", "Invalid release_date values"),
	count = c(bad_gross_count, bad_date_count)
)
knitr::kable(logic_table, caption = "Logical consistency checks")
```

## Estimate impact of removing rows with missing values or gross = 0

```{r}
# Row-level missing values (NA or blank strings) across all columns
row_has_missing <- apply(movie_profit, 1, function(row) {
	any(is.na(row) | trimws(as.character(row)) == "")
})

# Rows where either gross column is exactly zero
gross_is_zero <-
	movie_profit$domestic_gross == 0 |
	movie_profit$worldwide_gross == 0

overlap_count <- sum(row_has_missing & gross_is_zero)
remove_all_count <- sum(row_has_missing | gross_is_zero)
total_movies <- nrow(movie_profit)
remove_all_pct <- round(100 * remove_all_count / total_movies, 2)

# Same calculation, but ignore missing values in mpaa_rating
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")
row_has_missing_no_rating <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])

remove_no_rating_count <- sum(row_has_missing_no_rating | gross_is_zero)
remove_no_rating_pct <- round(100 * remove_no_rating_count / total_movies, 2)

impact_table <- data.frame(
	metric = c(
		"Rows with at least one missing value",
		"Rows with gross value 0 (domestic or worldwide)",
		"Overlap (missing and gross = 0)",
		"Rows removed (missing or gross = 0)",
		"Rows removed (ignoring missing mpaa_rating)"
	),
	count = c(
		sum(row_has_missing),
		sum(gross_is_zero),
		overlap_count,
		remove_all_count,
		remove_no_rating_count
	),
	percent = c(
		round(100 * sum(row_has_missing) / total_movies, 2),
		round(100 * sum(gross_is_zero) / total_movies, 2),
		round(100 * overlap_count / total_movies, 2),
		remove_all_pct,
		remove_no_rating_pct
	)
)

knitr::kable(impact_table, caption = "Impact of filtering rules")
```

## Remove rows with missing values (except rating) and gross = 0

```{r}
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")

row_has_NA <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])

gross_is_zero <-
	movie_profit$domestic_gross == 0 |
	movie_profit$worldwide_gross == 0

movie_profit_clean <- movie_profit[!(row_has_NA | gross_is_zero), ]

cleaning_table <- data.frame(
	metric = c("Rows before cleaning", "Rows after cleaning", "Rows removed"),
	count = c(nrow(movie_profit), nrow(movie_profit_clean), nrow(movie_profit) - nrow(movie_profit_clean))
)

knitr::kable(cleaning_table, caption = "Cleaning summary")
```

## Create new derived columns
New columns:
- International Gross (numerical)
- ROI domestic (numerical)
- ROI worldwide (numerical)
- Top-100 domestic gross flag (boolean)
- Top-100 international gross flag (boolean)
- Is Horror Movie (logical flag)

```{r}
movie_profit_enriched <- movie_profit_clean

# International gross is the non-domestic part of worldwide gross
movie_profit_enriched$international_gross <- movie_profit_enriched$worldwide_gross - movie_profit_enriched$domestic_gross

# ROI factors (gross divided by production budget)
movie_profit_enriched$roi_domestic_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, domestic_gross / production_budget, NA_real_)
)

movie_profit_enriched$roi_worldwide_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, worldwide_gross / production_budget, NA_real_)
)

# Top-100 flags
rank_domestic <- rank(-movie_profit_enriched$domestic_gross, ties.method = "first")
rank_international <- rank(-movie_profit_enriched$international_gross, ties.method = "first")

movie_profit_enriched$top100_domestic_gross <- rank_domestic <= 100
movie_profit_enriched$top100_international_gross <- rank_international <= 100

# Logical flag for horror movies based on genre text
movie_profit_enriched$is_horror_movie <-
	grepl("horror", movie_profit_enriched$genre, ignore.case = TRUE)

new_columns_preview <- movie_profit_enriched[, c(
	"movie",
	"production_budget",
	"domestic_gross",
	"worldwide_gross",
	"international_gross",
	"roi_domestic_factor",
	"roi_worldwide_factor",
	"top100_domestic_gross",
	"top100_international_gross",
	"is_horror_movie"
)]

knitr::kable(head(new_columns_preview, 5), caption = "Preview of new derived columns")
```

## Basic descriptive statistics

```{r}
analysis_data <- movie_profit_enriched

numeric_vars <- names(analysis_data)[sapply(analysis_data, is.numeric)]
# Exclude row ID from numeric summaries
numeric_vars <- setdiff(numeric_vars, "X")

# Focus categorical summaries on meaningful analysis columns
categorical_vars <- intersect(c("genre", "mpaa_rating", "distributor"), names(analysis_data))

release_dates <- analysis_data$release_date
profile_table <- data.frame(
	metric = c(
		"Rows",
		"Columns",
		"Numeric variables",
		"Categorical variables",
		"Unique movies",
		"Valid release dates",
		"Release year range"
	),
	value = c(
		nrow(analysis_data),
		ncol(analysis_data),
		length(numeric_vars),
		length(categorical_vars),
		length(unique(analysis_data$movie)),
		length(release_dates),
		paste(format(min(release_dates), "%Y"), "to", format(max(release_dates), "%Y"))
	)
)
knitr::kable(profile_table, caption = "Analysis dataset profile")

if (length(numeric_vars) > 0) {
	num_summary_mat <- do.call(cbind, lapply(analysis_data[, numeric_vars, drop = FALSE], summary))
	num_summary <- data.frame(
		statistic = rownames(num_summary_mat),
		num_summary_mat,
		row.names = NULL,
		check.names = FALSE
	)

	for (col in numeric_vars) {
		num_summary[[col]] <- format(
			round(as.numeric(num_summary[[col]]), 2),
			big.mark = ",",
			scientific = FALSE,
			trim = TRUE
		)
	}

	print(knitr::kable(num_summary, caption = "Numeric summary"))

	mean_median_table <- data.frame(
		variable = numeric_vars,
		mean = round(sapply(analysis_data[, numeric_vars, drop = FALSE], mean, na.rm = TRUE), 2),
		median = round(sapply(analysis_data[, numeric_vars, drop = FALSE], median, na.rm = TRUE), 2),
		row.names = NULL
	)
	mean_median_table$mean <- format(mean_median_table$mean, big.mark = ",", scientific = FALSE, trim = TRUE)
	mean_median_table$median <- format(mean_median_table$median, big.mark = ",", scientific = FALSE, trim = TRUE)
	print(knitr::kable(mean_median_table, caption = "Mean and median by numeric variable"))

	sd_table <- data.frame(
		variable = numeric_vars,
		sd = round(sapply(analysis_data[, numeric_vars, drop = FALSE], sd, na.rm = TRUE), 2),
		row.names = NULL
	)
 sd_table$sd <- format(sd_table$sd, big.mark = ",", scientific = FALSE, trim = TRUE)
	print(knitr::kable(sd_table, caption = "Standard deviation by variable"))

	outlier_stats <- do.call(rbind, lapply(numeric_vars, function(v) {
		x <- analysis_data[[v]]
		q1 <- quantile(x, 0.25, na.rm = TRUE)
		q3 <- quantile(x, 0.75, na.rm = TRUE)
		iqr_val <- q3 - q1
		lower <- q1 - 1.5 * iqr_val
		upper <- q3 + 1.5 * iqr_val
		low_outlier <- x < lower
		high_outlier <- x > upper
		is_outlier <- low_outlier | high_outlier

		data.frame(
			variable = v,
			outlier_count = sum(is_outlier, na.rm = TRUE),
			outlier_percent = round(100 * sum(is_outlier, na.rm = TRUE) / sum(!is.na(x)), 2),
			low_outlier_count = sum(low_outlier, na.rm = TRUE),
			high_outlier_count = sum(high_outlier, na.rm = TRUE)
		)
	}))

	print(knitr::kable(outlier_stats, caption = "IQR outlier summary"))
}


if (length(categorical_vars) > 0) {
	cat("\nCategorical variable details:\n")
	cat_table <- do.call(rbind, lapply(categorical_vars, function(v) {
		x <- analysis_data[[v]]
		x_non_missing <- x[!(is.na(x) | trimws(x) == "")]
		tab <- sort(table(x_non_missing), decreasing = TRUE)
		data.frame(
			variable = v,
			unique_levels = length(unique(x_non_missing)),
			most_common = names(tab)[1],
			most_common_count = as.integer(tab[1])
		)
	}))
	print(knitr::kable(cat_table, caption = "Categorical variable summary"))
}
```

## Simple distributions

```{r, include=FALSE}
analysis_data <- if (exists("movie_profit_clean")) movie_profit_clean else movie_profit
```

```{r}
if ("genre" %in% names(analysis_data)) {
	genre_tab <- sort(table(analysis_data$genre), decreasing = TRUE)
	top_n <- min(10, length(genre_tab))
	barplot(
		genre_tab[1:top_n],
		las = 2,
		col = "tan",
		main = "Top Genres",
		ylab = "Number of Movies"
	)
}
```

```{r}
if ("mpaa_rating" %in% names(analysis_data)) {
	rating_tab <- sort(table(analysis_data$mpaa_rating), decreasing = TRUE)
	barplot(
		rating_tab,
		col = "skyblue",
		main = "MPAA Rating Distribution",
		ylab = "Number of Movies"
	)
}
```

```{r}
release_dates <- analysis_data$release_date
release_year <- as.integer(format(release_dates, "%Y"))
year_tab <- table(release_year)
year_tab <- year_tab[order(as.integer(names(year_tab)))]
if (length(year_tab) > 0) {
	barplot(
		as.integer(year_tab),
		names.arg = names(year_tab),
		col = "darkorange",
		border = "white",
		las = 2,
		cex.names = 0.7,
		xlab = "Release Year",
		ylab = "Number of Movies",
		main = "Movies Released Per Year"
	)
}
```

```{r}
if ("distributor" %in% names(analysis_data)) {
	distributor_tab <- sort(table(analysis_data$distributor), decreasing = TRUE)
	top_n <- min(10, length(distributor_tab))
	barplot(
		distributor_tab[1:top_n],
		las = 2,
		cex.names = 0.8,
		col = "cornflowerblue",
		main = "Top 10 Distributors",
		ylab = "Number of Movies"
	)
}
```

## Horror vs non-horror comparison

```{r}
compare_data <- movie_profit_enriched

compare_data$movie_type <- ifelse(compare_data$is_horror_movie, "Horror", "Non-horror")

count_iqr_outliers <- function(x) {
	x <- x[!is.na(x)]
	if (length(x) < 4) {
		return(0)
	}
	q1 <- quantile(x, 0.25, na.rm = TRUE)
	q3 <- quantile(x, 0.75, na.rm = TRUE)
	iqr_val <- q3 - q1
	lower <- q1 - 1.5 * iqr_val
	upper <- q3 + 1.5 * iqr_val
	sum(x < lower | x > upper)
}

summary_table <- data.frame(
	metric = c(
		"movie_count",
		"budget_mean",
		"budget_q1",
		"budget_median",
		"budget_q3",
		"worldwide_mean",
		"worldwide_q1",
		"worldwide_median",
		"worldwide_q3",
		"roi_domestic_mean",
		"roi_domestic_q1",
		"roi_domestic_median",
		"roi_domestic_q3",
		"roi_domestic_outlier_count",
		"roi_worldwide_mean",
		"roi_worldwide_q1",
		"roi_worldwide_median",
		"roi_worldwide_q3",
		"roi_worldwide_outlier_count"
	),
	Horror = c(
		sum(compare_data$is_horror_movie),
		round(mean(compare_data$production_budget[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$production_budget[compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$production_budget[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$production_budget[compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		round(mean(compare_data$worldwide_gross[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$worldwide_gross[compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$worldwide_gross[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$worldwide_gross[compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		round(mean(compare_data$roi_domestic_factor[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_domestic_factor[compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$roi_domestic_factor[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_domestic_factor[compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		count_iqr_outliers(compare_data$roi_domestic_factor[compare_data$is_horror_movie]),
		round(mean(compare_data$roi_worldwide_factor[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_worldwide_factor[compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$roi_worldwide_factor[compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_worldwide_factor[compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		count_iqr_outliers(compare_data$roi_worldwide_factor[compare_data$is_horror_movie])
	),
	`Non-horror` = c(
		sum(!compare_data$is_horror_movie),
		round(mean(compare_data$production_budget[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$production_budget[!compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$production_budget[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$production_budget[!compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		round(mean(compare_data$worldwide_gross[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$worldwide_gross[!compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$worldwide_gross[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$worldwide_gross[!compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		round(mean(compare_data$roi_domestic_factor[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_domestic_factor[!compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$roi_domestic_factor[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_domestic_factor[!compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		count_iqr_outliers(compare_data$roi_domestic_factor[!compare_data$is_horror_movie]),
		round(mean(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie], 0.25, na.rm = TRUE), 2),
		round(median(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie], na.rm = TRUE), 2),
		round(quantile(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie], 0.75, na.rm = TRUE), 2),
		count_iqr_outliers(compare_data$roi_worldwide_factor[!compare_data$is_horror_movie])
	),
	check.names = FALSE
)

numeric_cols <- setdiff(names(summary_table), "metric")
for (col in numeric_cols) {
	summary_table[[col]] <- format(summary_table[[col]], big.mark = ",", scientific = FALSE, trim = TRUE)
}

knitr::kable(summary_table, caption = "Horror vs non-horror basic metrics")
```

## Horror vs non-horror movies over time

```{r}
year_vals <- as.integer(format(compare_data$release_date, "%Y"))
valid_rows <- !is.na(year_vals) & year_vals >= 1975
year_range <- range(year_vals[valid_rows], na.rm = TRUE)

horror_year_tab <- table(year_vals[valid_rows & compare_data$is_horror_movie])
horror_year_tab <- horror_year_tab[order(as.integer(names(horror_year_tab)))]

if (length(horror_year_tab) > 0) {
	plot(
		as.integer(names(horror_year_tab)),
		as.integer(horror_year_tab),
		type = "l",
		lwd = 2,
		col = "firebrick",
		xlim = year_range,
		xlab = "Release Year",
		ylab = "Number of Horror Movies",
		main = "Horror Movies per Year"
	)
}
```

```{r}
year_vals <- as.integer(format(compare_data$release_date, "%Y"))
valid_rows <- !is.na(year_vals) & year_vals >= 1975
year_range <- range(year_vals[valid_rows], na.rm = TRUE)

non_horror_year_tab <- table(year_vals[valid_rows & !compare_data$is_horror_movie])
non_horror_year_tab <- non_horror_year_tab[order(as.integer(names(non_horror_year_tab)))]

if (length(non_horror_year_tab) > 0) {
	plot(
		as.integer(names(non_horror_year_tab)),
		as.integer(non_horror_year_tab),
		type = "l",
		lwd = 2,
		col = "steelblue",
		xlim = year_range,
		xlab = "Release Year",
		ylab = "Number of Non-horror Movies",
		main = "Non-horror Movies per Year"
	)
}
```

## Average ROI by year: horror vs non-horror

```{r}
year_vals <- as.integer(format(compare_data$release_date, "%Y"))
roi_vals <- compare_data$roi_worldwide_factor

valid_rows <- !is.na(year_vals) & !is.na(roi_vals)
all_years <- sort(unique(year_vals[valid_rows]))
all_years <- all_years[all_years >= 1975]

horror_roi_by_year <- tapply(
	roi_vals[valid_rows & compare_data$is_horror_movie],
	year_vals[valid_rows & compare_data$is_horror_movie],
	mean
)

non_horror_roi_by_year <- tapply(
	roi_vals[valid_rows & !compare_data$is_horror_movie],
	year_vals[valid_rows & !compare_data$is_horror_movie],
	mean
)

horror_series <- as.numeric(horror_roi_by_year[as.character(all_years)])
non_horror_series <- as.numeric(non_horror_roi_by_year[as.character(all_years)])

if (length(all_years) > 0) {
	y_cap <- 20
	horror_plot_series <- pmin(horror_series, y_cap)
	non_horror_plot_series <- pmin(non_horror_series, y_cap)

	horror_outlier_idx <- which(!is.na(horror_series) & horror_series > y_cap)
	non_horror_outlier_idx <- which(!is.na(non_horror_series) & non_horror_series > y_cap)

	plot(
		all_years,
		horror_plot_series,
		type = "l",
		lwd = 2,
		col = "firebrick",
		xlab = "Release Year",
		ylab = "Average ROI (Worldwide)",
		main = "Average ROI per Year: Horror vs Non-horror (capped at 20)",
		ylim = c(0, y_cap)
	)

	lines(all_years, non_horror_plot_series, lwd = 2, col = "steelblue")

	if (length(horror_outlier_idx) > 0) {
		points(all_years[horror_outlier_idx], rep(y_cap, length(horror_outlier_idx)),
			pch = 24, bg = "firebrick", col = "firebrick", cex = 0.8)
	}

	if (length(non_horror_outlier_idx) > 0) {
		points(all_years[non_horror_outlier_idx], rep(y_cap, length(non_horror_outlier_idx)),
			pch = 24, bg = "steelblue", col = "steelblue", cex = 0.8)
	}

	legend("topright",
		legend = c("Horror", "Non-horror", paste0("Capped value (> ", y_cap, ")")),
		col = c("firebrick", "steelblue", "black"),
		lwd = c(2, 2, NA),
		pch = c(NA, NA, 24),
		pt.bg = c(NA, NA, "gray60"),
		bty = "n"
	)

	horror_outlier_note <- if (length(horror_outlier_idx) > 0) {
		paste(
			paste0(
				all_years[horror_outlier_idx],
				" (",
				round(horror_series[horror_outlier_idx], 2),
				"x)"
			),
			collapse = ", "
		)
	} else {
		"none"
	}

	non_horror_outlier_note <- if (length(non_horror_outlier_idx) > 0) {
		paste(
			paste0(
				all_years[non_horror_outlier_idx],
				" (",
				round(non_horror_series[non_horror_outlier_idx], 2),
				"x)"
			),
			collapse = ", "
		)
	} else {
		"none"
	}

	cat(paste0("Outlier years (mean ROI > ", y_cap, "):\n"))
	cat("- Horror:", horror_outlier_note, "\n")
	cat("- Non-horror:", non_horror_outlier_note, "\n")
}
```


## Median ROI by year: horror vs non-horror

```{r}
year_vals <- as.integer(format(compare_data$release_date, "%Y"))
roi_vals <- compare_data$roi_worldwide_factor

valid_rows <- !is.na(year_vals) & !is.na(roi_vals)
all_years <- sort(unique(year_vals[valid_rows]))
all_years <- all_years[all_years >= 1975]

horror_roi_by_year <- tapply(
	roi_vals[valid_rows & compare_data$is_horror_movie],
	year_vals[valid_rows & compare_data$is_horror_movie],
	median
)

non_horror_roi_by_year <- tapply(
	roi_vals[valid_rows & !compare_data$is_horror_movie],
	year_vals[valid_rows & !compare_data$is_horror_movie],
	median
)

horror_series <- as.numeric(horror_roi_by_year[as.character(all_years)])
non_horror_series <- as.numeric(non_horror_roi_by_year[as.character(all_years)])

if (length(all_years) > 0) {
	y_cap <- 20
	horror_plot_series <- pmin(horror_series, y_cap)
	non_horror_plot_series <- pmin(non_horror_series, y_cap)

	horror_outlier_idx <- which(!is.na(horror_series) & horror_series > y_cap)
	non_horror_outlier_idx <- which(!is.na(non_horror_series) & non_horror_series > y_cap)

	plot(
		all_years,
		horror_plot_series,
		type = "l",
		lwd = 2,
		col = "firebrick",
		xlab = "Release Year",
		ylab = "Median ROI (Worldwide)",
		main = "Median ROI per Year: Horror vs Non-horror (capped at 20)",
		ylim = c(0, y_cap)
	)

	lines(all_years, non_horror_plot_series, lwd = 2, col = "steelblue")

	if (length(horror_outlier_idx) > 0) {
		points(all_years[horror_outlier_idx], rep(y_cap, length(horror_outlier_idx)),
			pch = 24, bg = "firebrick", col = "firebrick", cex = 0.8)
	}

	if (length(non_horror_outlier_idx) > 0) {
		points(all_years[non_horror_outlier_idx], rep(y_cap, length(non_horror_outlier_idx)),
			pch = 24, bg = "steelblue", col = "steelblue", cex = 0.8)
	}

	legend("topright",
		legend = c("Horror", "Non-horror", paste0("Capped value (> ", y_cap, ")")),
		col = c("firebrick", "steelblue", "black"),
		lwd = c(2, 2, NA),
		pch = c(NA, NA, 24),
		pt.bg = c(NA, NA, "gray60"),
		bty = "n"
	)

	horror_outlier_note <- if (length(horror_outlier_idx) > 0) {
		paste(
			paste0(
				all_years[horror_outlier_idx],
				" (",
				round(horror_series[horror_outlier_idx], 2),
				"x)"
			),
			collapse = ", "
		)
	} else {
		"none"
	}

	non_horror_outlier_note <- if (length(non_horror_outlier_idx) > 0) {
		paste(
			paste0(
				all_years[non_horror_outlier_idx],
				" (",
				round(non_horror_series[non_horror_outlier_idx], 2),
				"x)"
			),
			collapse = ", "
		)
	} else {
		"none"
	}

	cat(paste0("Outlier years (median ROI > ", y_cap, "):\n"))
	cat("- Horror:", horror_outlier_note, "\n")
	cat("- Non-horror:", non_horror_outlier_note, "\n")
}
```

## Interactive movie table (filter + sort)

```{r}
table_data <- if (exists("movie_profit_enriched")) movie_profit_enriched else movie_profit

interactive_table <- table_data

if (!"roi_worldwide_factor" %in% names(interactive_table)) {
	interactive_table$roi_worldwide_factor <- with(
		interactive_table,
		ifelse(production_budget > 0, worldwide_gross / production_budget, NA_real_)
	)
}

# Keep numeric values numeric so sorting remains correct in DT.
interactive_table$production_budget_k <- interactive_table$production_budget / 1000
interactive_table$domestic_gross_k <- interactive_table$domestic_gross / 1000
interactive_table$worldwide_gross_k <- interactive_table$worldwide_gross / 1000

interactive_table$release_date_fmt <- ifelse(
	is.na(interactive_table$release_date),
	NA_character_,
	format(interactive_table$release_date, "%Y-%m-%d")
)

display_table <- interactive_table[, c(
	"movie",
	"genre",
	"roi_worldwide_factor",
	"production_budget_k",
	"domestic_gross_k",
	"worldwide_gross_k",
	"release_date_fmt",
	"mpaa_rating",
	"distributor"
)]

names(display_table) <- c(
	"Movie",
	"Genre",
	"ROI (x)",
	"Production Cost (x1000 $)",
	"Domestic Gross (x1000 $)",
	"Worldwide Gross (x1000 $)",
	"Release Date",
	"MPAA Rating",
	"Distributor"
)

display_table$`ROI (x)` <- round(display_table$`ROI (x)`, 2)
display_table$`Production Cost (x1000 $)` <- round(display_table$`Production Cost (x1000 $)`, 1)
display_table$`Domestic Gross (x1000 $)` <- round(display_table$`Domestic Gross (x1000 $)`, 1)
display_table$`Worldwide Gross (x1000 $)` <- round(display_table$`Worldwide Gross (x1000 $)`, 1)

DT::datatable(
	display_table,
	rownames = FALSE,
	options = list(
		pageLength = 15,
		lengthMenu = c(10, 15, 25, 50),
		order = list(list(5, "desc")),
		autoWidth = TRUE
	),
	callback = DT::JS(
		"var genreColumn = 1;",
		"var wrapper = $(table.table().container());",
		"var controls = $('<div style=\"margin-bottom: 8px;\"></div>');",
		"var label = $('<label style=\"margin-right: 8px; font-weight: 600;\">Genre:</label>');",
		"var select = $('<select style=\"padding: 2px 6px;\"><option value=\"\">All genres</option></select>');",
		"select.on('change', function() {",
		"  var val = $.fn.dataTable.util.escapeRegex($(this).val());",
		"  table.column(genreColumn).search(val ? '^' + val + '$' : '', true, false).draw();",
		"});",
		"table.column(genreColumn).data().unique().sort().each(function(d) {",
		"  if (d !== null && d !== undefined && d !== '') {",
		"    select.append($('<option></option>').attr('value', d).text(d));",
		"  }",
		"});",
		"controls.append(label).append(select);",
		"wrapper.prepend(controls);"
	),
	caption = "Interactive movie table (filter by Genre in the header row; money columns are in thousands of dollars)"
)
```

## Some examples of what we can learn from the interactive table

Horror demonstrates remarkable ROI efficiency in this dataset:

**ROI Leadership:** Horror is significantly overrepresented at the top of the ROI rankings, with 5 of the top 10 films by ROI being horror titles (including #1 and #2)

**Early Advantage:** Except for some outliers (Exorcist 2 and The Shining), the earlierst horror movies (before 1982) all had a ROI > 10x, while the non-horror movies in that period had lower ROIs on average.

**Number of Low-Budget Movies:** When sorting after the lowest production budget, we can see that Horror Movies are cheap to produce (20 movies cost <= $1M). For other Genres, the numbers are: Action (4), Adventure (2), Comedy (28), Drama (69).

**Low Risk, High Reward::** When sorting for production budget, we can also see that the cheapest Horror movies (<= $1M) are the movies which end up having a ROI > 100x (5/20 movies), while most movies have a ROI of >= 8x (15/20 movies). At the same time, in other genres the cheapest movies have a way lower ROI on average, with Drama only reaching an ROI >= 5x in half of the movies with a production budget <= $1M.




## Top 20 movies by ROI

```{r}
roi_plot_data <- compare_data[!is.na(compare_data$roi_worldwide_factor), ]
roi_plot_data <- roi_plot_data[order(-roi_plot_data$roi_worldwide_factor), ]
roi_plot_data <- head(roi_plot_data, 20)

if (nrow(roi_plot_data) > 0) {
	roi_plot_data <- roi_plot_data[order(roi_plot_data$roi_worldwide_factor), ]
	roi_colors <- genre_color(roi_plot_data$genre)

	par(mar = c(5, 12, 4, 2))
	roi_bars <- barplot(
		roi_plot_data$roi_worldwide_factor,
		horiz = TRUE,
		las = 1,
		col = roi_colors,
		border = "white",
		xlab = "ROI (x)",
		main = "Top 20 Movies by Worldwide ROI",
		names.arg = roi_plot_data$movie,
		cex.names = 0.65
	)

	text(
		x = roi_plot_data$roi_worldwide_factor,
		y = roi_bars,
		labels = round(roi_plot_data$roi_worldwide_factor, 1),
		pos = 4,
		cex = 0.75
	)

	legend(
		"bottomright",
		legend = names(genre_palette),
		fill = genre_palette,
		bty = "n",
		cex = 0.8
	)
}
```

## Earliest 50 movies per genre

```{r, fig.width=14, fig.height=12}
genres_to_plot <- intersect(names(genre_palette), unique(compare_data$genre))
earliest_by_genre <- lapply(genres_to_plot, function(genre_name) {
	genre_data <- compare_data[compare_data$genre == genre_name & !is.na(compare_data$release_date) & !is.na(compare_data$roi_worldwide_factor), ]
	genre_data <- genre_data[order(genre_data$release_date), ]
	head(genre_data, 50)
})

names(earliest_by_genre) <- genres_to_plot

max_roi_early <- max(unlist(lapply(earliest_by_genre, function(x) x$roi_worldwide_factor)), na.rm = TRUE)
if (is.finite(max_roi_early)) {
	y_limit_early <- c(0, max_roi_early)
} else {
	y_limit_early <- c(0, 1)
}

old_par <- par(mfrow = c(3, 2), mar = c(4, 4, 3, 1))

for (genre_name in names(earliest_by_genre)) {
	genre_data <- earliest_by_genre[[genre_name]]
	plot(
		genre_data$release_date,
		genre_data$roi_worldwide_factor,
		pch = 16,
		col = genre_palette[[genre_name]],
		xlab = "Release Date",
		ylab = "ROI (x)",
		main = paste(genre_name, "- earliest 50"),
		ylim = y_limit_early
	)
	abline(h = 1, lty = 3, col = "gray70")
}

if (length(genres_to_plot) < 6) {
	for (i in seq_len(6 - length(genres_to_plot))) {
		plot.new()
	}
}

par(old_par)
```

## Low-budget movies by genre

```{r}
budget_threshold <- 1e6
genre_order <- intersect(names(genre_palette), unique(compare_data$genre))

low_budget_count <- sapply(genre_order, function(genre_name) {
	sum(compare_data$genre == genre_name & !is.na(compare_data$production_budget) & compare_data$production_budget <= budget_threshold)
})

genre_total_count <- sapply(genre_order, function(genre_name) {
	sum(compare_data$genre == genre_name)
})

low_budget_pct <- round(100 * low_budget_count / genre_total_count, 1)

low_budget_table <- data.frame(
	genre = genre_order,
	low_budget_count = as.integer(low_budget_count),
	total_genre_movies = as.integer(genre_total_count),
	low_budget_percent = low_budget_pct,
	row.names = NULL
)

if (nrow(low_budget_table) > 0) {
	base_colors <- genre_color(low_budget_table$genre)
	bar_colors_full <- rbind(base_colors, adjustcolor(base_colors, alpha.f = 0.45))
	
	# Scale percentages to match the plot's y-axis range (0 to 10%)
	max_count <- max(low_budget_table$low_budget_count)
	y_max <- max_count * 1.15
	scale_factor <- y_max / 10  # Convert percentage (0-10%) to plot coordinates
	
	bar_matrix <- rbind(
		Count = low_budget_table$low_budget_count,
		Percentage = low_budget_table$low_budget_percent * scale_factor  # Scale percentages
	)
	
	old_par <- par(mar = c(7, 5, 4, 5))
	bar_positions <- barplot(
		bar_matrix,
		beside = TRUE,
		col = bar_colors_full,
		border = "white",
		las = 2,
		cex.names = 0.8,
		ylab = "Number of Movies",
		main = "Low-Budget Movies (≤$1M) by Genre",
		names.arg = low_budget_table$genre,
		ylim = c(0, y_max)
	)

	# Add labels: counts on first bars, percentages on second bars
	text(
		bar_positions[1, ],
		low_budget_table$low_budget_count,
		labels = low_budget_table$low_budget_count,
		pos = 3,
		cex = 0.75,
		font = 2
	)
	
	text(
		bar_positions[2, ],
		low_budget_table$low_budget_percent * scale_factor,
		labels = paste0(low_budget_table$low_budget_percent, "%"),
		pos = 3,
		cex = 0.75,
		font = 2
	)

	# Add secondary y-axis for percentages (0-10%)
	axis(4, at = seq(0, y_max, length.out = 6), labels = seq(0, 10, by = 2), col = "gray50", col.axis = "gray50")
	mtext("Percentage of Total Genre Movies (%)", side = 4, line = 3, col = "gray50", cex = 0.9)

	# Create legend with actual colors
	legend_colors <- c(base_colors[1], adjustcolor(base_colors[1], alpha.f = 0.45))
	legend(
		"topleft",
		legend = c("Low-budget count (≤$1M)", "% of total movies in genre"),
		fill = legend_colors,
		bty = "n",
		cex = 0.9
	)
	par(old_par)
}
```

## Top 250 movies with lowest production budget

```{r}
lowest_budget_data <- compare_data[!is.na(compare_data$production_budget) & !is.na(compare_data$worldwide_gross), ]
lowest_budget_data <- lowest_budget_data[order(lowest_budget_data$production_budget), ]
lowest_budget_data <- head(lowest_budget_data, 250)

if (nrow(lowest_budget_data) > 0) {
	plot_colors <- genre_color(lowest_budget_data$genre)
	x_vals <- lowest_budget_data$production_budget / 1e6
	y_vals <- lowest_budget_data$worldwide_gross / 1e6

	# Light jitter to reduce overlap while preserving the overall pattern.
	x_jitter <- if (diff(range(x_vals, na.rm = TRUE)) > 0) {
		jitter(x_vals, amount = diff(range(x_vals, na.rm = TRUE)) * 0.003)
	} else {
		x_vals
	}

	plot(
		x_jitter,
		y_vals,
		pch = 16,
		col = plot_colors,
		xlab = "Production Budget ($ millions)",
		ylab = "Worldwide Revenue ($ millions)",
		main = "Lowest Budget Movies: Budget vs. Revenue",
		cex = 1.2
	)
	
	legend(
		"topleft",
		legend = names(genre_palette),
		col = genre_palette,
		pch = 16,
		bty = "n",
		cex = 0.85
	)
}
```


## First Result: Cheap Horror Movies Can Be Very Profitable
The data clearly shows that horror movies are often produced with low budgets (20 movies with a production budget of $1M or less) and can achieve very high ROIs (5 of the 20 cheapest horror movies have a ROI > 100x, and 15/20 have a ROI >= 8x). This suggests that low-budget horror films can be a lucrative segment in the movie industry, offering high returns on investment compared to other genres.


## Budget distribution by genre (count of movies per budget band)

```{r}
budget_data <- compare_data[!is.na(compare_data$production_budget) & !is.na(compare_data$genre), ]
genre_order <- intersect(names(genre_palette), unique(budget_data$genre))

budget_breaks <- c(-Inf, 1e6, 5e6, 10e6, 20e6, 50e6, 100e6, Inf)
budget_labels <- c("< $1M", "$1M-$5M", "$5M-$10M", "$10M-$20M", "$20M-$50M", "$50M-$100M", "> $100M")

if (length(genre_order) > 0 && nrow(budget_data) > 0) {
	budget_data$budget_band <- cut(
		budget_data$production_budget,
		breaks = budget_breaks,
		labels = budget_labels,
		right = FALSE
	)

	band_count_table <- table(
		factor(budget_data$budget_band, levels = budget_labels),
		factor(budget_data$genre, levels = genre_order)
	)

	band_colors <- c("#2E86AB", "#A23B72", "#F18F01", "#FF7043", "#4CAF50", "#6C757D", "#9C27B0")
	old_par <- par(mar = c(7, 5, 4, 1))
	bar_positions <- barplot(
		band_count_table,
		beside = TRUE,
		col = band_colors,
		border = "white",
		las = 2,
		cex.names = 0.9,
		ylab = "Movie Count",
		main = "How Production Budgets Are Distributed Across Genres",
		names.arg = genre_order,
		ylim = c(0, max(band_count_table) * 1.15)
	)

	text(
		x = as.vector(bar_positions),
		y = as.vector(band_count_table),
		labels = as.vector(band_count_table),
		pos = 3,
		cex = 0.65
	)

	legend(
		"topright",
		legend = budget_labels,
		fill = band_colors,
		title = "Budget band",
		bty = "n",
		cex = 0.85
	)
	par(old_par)
}
```

## Multiple Linear Regression: Worldwide Gross vs Budget and Genre

```{r}
mlr_data <- compare_data[
	!is.na(compare_data$worldwide_gross) &
	!is.na(compare_data$production_budget) &
	!is.na(compare_data$genre),
	c("worldwide_gross", "production_budget", "genre")
]
mlr_data <- subset(mlr_data, worldwide_gross > 0 & production_budget > 0)

# Set reference genre to the most frequent one for interpretability
most_common_genre <- names(sort(table(mlr_data$genre), decreasing = TRUE))[1]
mlr_data$genre <- relevel(factor(mlr_data$genre), ref = most_common_genre)

mlr_data$log_worldwide <- log(mlr_data$worldwide_gross)
mlr_data$log_budget    <- log(mlr_data$production_budget)

# ── Models ────────────────────────────────────────────────────────────────────
model_main        <- lm(log_worldwide ~ log_budget + genre, data = mlr_data)
model_interaction <- lm(log_worldwide ~ log_budget * genre, data = mlr_data)

# ── 1) R² ─────────────────────────────────────────────────────────────────────
int_summary <- summary(model_interaction)
cat("R-squared:", round(int_summary$r.squared, 3),
    "| Adjusted R-squared:", round(int_summary$adj.r.squared, 3), "\n\n")

# ── 2) Term-level F-tests & interaction comparison ────────────────────────────
cat("Term-level F-tests (full interaction model):\n")
print(drop1(model_interaction, test = "F"))

cat("\nDoes adding interaction improve fit?\n")
print(anova(model_main, model_interaction))

# ── 3) Budget effect + interaction summary ────────────────────────────────────
coef_table <- coef(int_summary)

if ("log_budget" %in% rownames(coef_table)) {
  budget_beta <- coef_table["log_budget", "Estimate"]
  budget_p    <- coef_table["log_budget", "Pr(>|t|)"]
  cat("\nBudget elasticity (reference genre:", most_common_genre, "):",
      "beta =", round(budget_beta, 3),
      ", p-value =", signif(budget_p, 3), "\n")
}

interaction_rows <- grep("^log_budget:genre", rownames(coef_table))
if (length(interaction_rows) > 0) {
  min_p <- min(coef_table[interaction_rows, "Pr(>|t|)"], na.rm = TRUE)
  cat("Smallest interaction-term p-value:", signif(min_p, 3), "\n")
}

# ── 4) Full coefficient table ─────────────────────────────────────────────────
cat("\nFull coefficient table:\n")
printCoefmat(coef_table, digits = 3, signif.stars = TRUE)

# ── 5) Effective budget elasticity per genre ──────────────────────────────────
base_slope    <- coef(model_interaction)["log_budget"]
int_coefs     <- coef(model_interaction)[grep("log_budget:genre", names(coef(model_interaction)))]
genre_names   <- sub("log_budget:genre", "", names(int_coefs))

slopes <- data.frame(
  genre = c(most_common_genre, genre_names),
  elasticity = c(base_slope, base_slope + int_coefs),
  row.names = NULL
)
cat("\nEffective budget elasticity by genre:\n")
print(slopes[order(-slopes$elasticity), ], row.names = FALSE)

# ── 6) Interaction plot ───────────────────────────────────────────────────────
# ── 6) Interaction plot (base R) ──────────────────────────────────────────────
genres  <- levels(mlr_data$genre)
colours <- palette.colors(length(genres), palette = "Okabe-Ito")  # colourblind-safe

# Set up empty plot with correct axis ranges
plot(mlr_data$log_budget, mlr_data$log_worldwide,
    type = "n",
	xlab = "log(Production Budget)",
    ylab = "log(Worldwide Gross)",
    main = "Budget vs. Revenue by Genre (log-log scale)")

# Draw points and regression lines per genre
for (i in seq_along(genres)) {
	g    <- genres[i]
	sub  <- mlr_data[mlr_data$genre == g, ]
	fit  <- lm(log_worldwide ~ log_budget, data = sub)

	points(sub$log_budget, sub$log_worldwide,
			col = adjustcolor(colours[i], alpha.f = 0.25), pch = 16, cex = 0.7)
	abline(fit, col = colours[i], lwd = 2)
}

legend("topleft", legend = genres, col = colours, lwd = 2, bty = "n")

# # ── 7) Regression diagnostics ─────────────────────────────────────────────────
# par(mfrow = c(2, 2))
# plot(model_interaction)
# par(mfrow = c(1, 1))
```

## Second Result: Expensive Horror Movies make less Revenue
The multiple linear regression analaysis shows us, that 43.6% of the variance in revenue (adjusted R² = 0.436) can be explained through production budget and genre, which is reasonable given that many unobserved factors - marketing budget, cast, release timing, critical reception - also influence box office performance.

As one would expect, budget is a strong predictor for revenue, with a significant positive elasticity of around 0.97 for the reference genre (Drama), meaning that a 1% increase in budget is associated with a near-proportional 0.97% increase in worldwide gross, a near-proportional relationship that was highly significant (p < 2×10⁻¹⁶). For Comedy, Adventure and Action we can see elasticity scores of >1, meaning that these movies profit from higher budgets. Notably, Horror is the only Outlier with an elasticity of 0,68.

This shows us why there are significantly more cheap Horror movies than expensive ones, as higher budgets don't result in proportionally higher revenue.