---
title: "Movie Data Analysis"
author: "TBD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	comment = NA,
	fig.align = "center",
	fig.width = 10,
	fig.height = 5,
	out.width = "100%"
)
```

## Load and verify the data

```{r}
movie_profit <- read.csv("movie_profit.csv", stringsAsFactors = FALSE)

overview_table <- data.frame(
	Metric = c("Rows", "Columns", "Numeric columns", "Character columns"),
	Value = c(
		nrow(movie_profit),
		ncol(movie_profit),
		sum(sapply(movie_profit, is.numeric)),
		sum(sapply(movie_profit, is.character))
	)
)

knitr::kable(overview_table, caption = "Dataset overview")
```


## Check variable types and first 5 values

```{r}
variable_overview <- data.frame(
	variable = names(movie_profit),
	type = sapply(movie_profit, function(x) class(x)[1]),
	first_5_values = sapply(movie_profit, function(x) {
		paste(head(x, 5), collapse = ", ")
	}),
	row.names = NULL
)

knitr::kable(variable_overview, caption = "Variable types and first 5 values")
```

## Convert release_date to Date type and check conversion results

```{r}
release_date_raw <- movie_profit$release_date
movie_profit$release_date <- as.Date(release_date_raw, format = "%m/%d/%Y")

conversion_table <- data.frame(
	metric = c("Original type", "New type", "Invalid dates after conversion"),
	value = c(
		class(release_date_raw)[1],
		class(movie_profit$release_date)[1],
		sum(is.na(movie_profit$release_date) & !is.na(release_date_raw) & trimws(release_date_raw) != "")
	)
)

knitr::kable(conversion_table, caption = "release_date conversion summary")
```

## Check for missing values, duplicates, and logical consistency

```{r}
# Missing values (NA or blank strings)
missing_count <- sapply(movie_profit, function(x) {
	if (is.character(x)) {
		sum(is.na(x) | trimws(x) == "")
	} else {
		sum(is.na(x))
	}
})

missing_pct <- round(100 * missing_count / nrow(movie_profit), 2)
missing_summary <- data.frame(
	variable = names(missing_count),
	missing_count = as.integer(missing_count),
	missing_percent = missing_pct,
	row.names = NULL
)

# Duplicate rows
dup_rows <- duplicated(movie_profit)
dup_count <- sum(dup_rows)

# Numeric column checks
numeric_cols <- c("production_budget", "domestic_gross", "worldwide_gross")
numeric_checks <- data.frame()
for (col in numeric_cols) {
	if (col %in% names(movie_profit)) {
		x <- movie_profit[[col]]
		numeric_checks <- rbind(
			numeric_checks,
			data.frame(
				variable = col,
				negative_values = sum(x < 0, na.rm = TRUE),
				zero_values = sum(x == 0, na.rm = TRUE)
			)
		)
	}
}

# Logical consistency check
bad_gross_count <- NA
if (all(c("domestic_gross", "worldwide_gross") %in% names(movie_profit))) {
	bad_gross_rows <- which(movie_profit$worldwide_gross < movie_profit$domestic_gross)
	bad_gross_count <- length(bad_gross_rows)
}

# Date parsing check
bad_date_count <- sum(
	is.na(movie_profit$release_date) &
		!is.na(release_date_raw) &
		trimws(release_date_raw) != ""
)

knitr::kable(missing_summary, caption = "Missing values by variable")

dup_table <- data.frame(
	metric = "Duplicate rows",
	count = dup_count
)
knitr::kable(dup_table, caption = "Duplicate row summary")

knitr::kable(numeric_checks, caption = "Numeric consistency checks")

logic_table <- data.frame(
	check = c("Rows where worldwide_gross < domestic_gross", "Invalid release_date values"),
	count = c(bad_gross_count, bad_date_count)
)
knitr::kable(logic_table, caption = "Logical consistency checks")
```

## Estimate impact of removing rows with missing values or gross = 0

```{r}
# Row-level missing values (NA or blank strings) across all columns
row_has_missing <- apply(movie_profit, 1, function(row) {
	any(is.na(row) | trimws(as.character(row)) == "")
})

# Rows where either gross column is exactly zero
gross_is_zero <-
	movie_profit$domestic_gross == 0 |
	movie_profit$worldwide_gross == 0

overlap_count <- sum(row_has_missing & gross_is_zero)
remove_all_count <- sum(row_has_missing | gross_is_zero)
total_movies <- nrow(movie_profit)
remove_all_pct <- round(100 * remove_all_count / total_movies, 2)

# Same calculation, but ignore missing values in mpaa_rating
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")
row_has_missing_no_rating <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])

remove_no_rating_count <- sum(row_has_missing_no_rating | gross_is_zero)
remove_no_rating_pct <- round(100 * remove_no_rating_count / total_movies, 2)

impact_table <- data.frame(
	metric = c(
		"Rows with at least one missing value",
		"Rows with gross value 0 (domestic or worldwide)",
		"Overlap (missing and gross = 0)",
		"Rows removed (missing or gross = 0)",
		"Rows removed (ignoring missing mpaa_rating)"
	),
	count = c(
		sum(row_has_missing),
		sum(gross_is_zero),
		overlap_count,
		remove_all_count,
		remove_no_rating_count
	),
	percent = c(
		round(100 * sum(row_has_missing) / total_movies, 2),
		round(100 * sum(gross_is_zero) / total_movies, 2),
		round(100 * overlap_count / total_movies, 2),
		remove_all_pct,
		remove_no_rating_pct
	)
)

knitr::kable(impact_table, caption = "Impact of filtering rules")
```

## Remove rows with missing values (except rating) and gross = 0

```{r}
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")

row_has_NA <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])

gross_is_zero <-
	movie_profit$domestic_gross == 0 |
	movie_profit$worldwide_gross == 0

movie_profit_clean <- movie_profit[!(row_has_NA | gross_is_zero), ]

cleaning_table <- data.frame(
	metric = c("Rows before cleaning", "Rows after cleaning", "Rows removed"),
	count = c(nrow(movie_profit), nrow(movie_profit_clean), nrow(movie_profit) - nrow(movie_profit_clean))
)

knitr::kable(cleaning_table, caption = "Cleaning summary")
```

## Create new derived columns
New columns:
- International Gross (numerical)
- ROI domestic (numerical)
- ROI worldwide (numerical)
- Top-100 domestic gross flag (boolean)
- Top-100 international gross flag (boolean)

```{r}
movie_profit_enriched <- movie_profit_clean

# International gross is the non-domestic part of worldwide gross
movie_profit_enriched$international_gross <- movie_profit_enriched$worldwide_gross - movie_profit_enriched$domestic_gross

# ROI factors (gross divided by production budget)
movie_profit_enriched$roi_domestic_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, domestic_gross / production_budget, NA_real_)
)

movie_profit_enriched$roi_worldwide_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, worldwide_gross / production_budget, NA_real_)
)

# Top-100 flags
rank_domestic <- rank(-movie_profit_enriched$domestic_gross, ties.method = "first")
rank_international <- rank(-movie_profit_enriched$international_gross, ties.method = "first")

movie_profit_enriched$top100_domestic_gross <- rank_domestic <= 100
movie_profit_enriched$top100_international_gross <- rank_international <= 100

new_columns_preview <- movie_profit_enriched[, c(
	"movie",
	"production_budget",
	"domestic_gross",
	"worldwide_gross",
	"international_gross",
	"roi_domestic_factor",
	"roi_worldwide_factor",
	"top100_domestic_gross",
	"top100_international_gross"
)]

knitr::kable(head(new_columns_preview, 5), caption = "Preview of new derived columns")
```

## Basic descriptive statistics

```{r}
analysis_data <- movie_profit_enriched

numeric_vars <- names(analysis_data)[sapply(analysis_data, is.numeric)]
# Exclude row ID from numeric summaries
numeric_vars <- setdiff(numeric_vars, "X")

# Focus categorical summaries on meaningful analysis columns
categorical_vars <- intersect(c("genre", "mpaa_rating", "distributor"), names(analysis_data))

release_dates <- analysis_data$release_date
profile_table <- data.frame(
	metric = c(
		"Rows",
		"Columns",
		"Numeric variables",
		"Categorical variables",
		"Unique movies",
		"Valid release dates",
		"Release year range"
	),
	value = c(
		nrow(analysis_data),
		ncol(analysis_data),
		length(numeric_vars),
		length(categorical_vars),
		length(unique(analysis_data$movie)),
		length(release_dates),
		paste(format(min(release_dates), "%Y"), "to", format(max(release_dates), "%Y"))
	)
)
knitr::kable(profile_table, caption = "Analysis dataset profile")

if (length(numeric_vars) > 0) {
	num_summary_mat <- do.call(cbind, lapply(analysis_data[, numeric_vars, drop = FALSE], summary))
	num_summary <- data.frame(
		statistic = rownames(num_summary_mat),
		num_summary_mat,
		row.names = NULL,
		check.names = FALSE
	)

	for (col in numeric_vars) {
		num_summary[[col]] <- format(
			round(as.numeric(num_summary[[col]]), 2),
			big.mark = ",",
			scientific = FALSE,
			trim = TRUE
		)
	}

	print(knitr::kable(num_summary, caption = "Numeric summary"))

	mean_median_table <- data.frame(
		variable = numeric_vars,
		mean = round(sapply(analysis_data[, numeric_vars, drop = FALSE], mean, na.rm = TRUE), 2),
		median = round(sapply(analysis_data[, numeric_vars, drop = FALSE], median, na.rm = TRUE), 2),
		row.names = NULL
	)
	mean_median_table$mean <- format(mean_median_table$mean, big.mark = ",", scientific = FALSE, trim = TRUE)
	mean_median_table$median <- format(mean_median_table$median, big.mark = ",", scientific = FALSE, trim = TRUE)
	print(knitr::kable(mean_median_table, caption = "Mean and median by numeric variable"))

	sd_table <- data.frame(
		variable = numeric_vars,
		sd = round(sapply(analysis_data[, numeric_vars, drop = FALSE], sd, na.rm = TRUE), 2),
		row.names = NULL
	)
	sd_table$sd <- format(sd_table$sd, big.mark = ",", scientific = FALSE, trim = TRUE)
	print(knitr::kable(sd_table, caption = "Standard deviation by variable"))

	outlier_stats <- do.call(rbind, lapply(numeric_vars, function(v) {
		x <- analysis_data[[v]]
		q1 <- quantile(x, 0.25, na.rm = TRUE)
		q3 <- quantile(x, 0.75, na.rm = TRUE)
		iqr_val <- q3 - q1
		lower <- q1 - 1.5 * iqr_val
		upper <- q3 + 1.5 * iqr_val
		low_outlier <- x < lower
		high_outlier <- x > upper
		is_outlier <- low_outlier | high_outlier

		data.frame(
			variable = v,
			outlier_count = sum(is_outlier, na.rm = TRUE),
			outlier_percent = round(100 * sum(is_outlier, na.rm = TRUE) / sum(!is.na(x)), 2),
			low_outlier_count = sum(low_outlier, na.rm = TRUE),
			high_outlier_count = sum(high_outlier, na.rm = TRUE)
		)
	}))

	print(knitr::kable(outlier_stats, caption = "IQR outlier summary"))
}


if (length(categorical_vars) > 0) {
	cat("\nCategorical variable details:\n")
	cat_table <- do.call(rbind, lapply(categorical_vars, function(v) {
		x <- analysis_data[[v]]
		x_non_missing <- x[!(is.na(x) | trimws(x) == "")]
		tab <- sort(table(x_non_missing), decreasing = TRUE)
		data.frame(
			variable = v,
			unique_levels = length(unique(x_non_missing)),
			most_common = names(tab)[1],
			most_common_count = as.integer(tab[1])
		)
	}))
	print(knitr::kable(cat_table, caption = "Categorical variable summary"))
}
```

## Simple distributions

```{r, include=FALSE}
analysis_data <- if (exists("movie_profit_clean")) movie_profit_clean else movie_profit
```

```{r}
if ("production_budget" %in% names(analysis_data)) {
	hist(analysis_data$production_budget,
		main = "Production Budget Distribution",
		xlab = "Production Budget",
		col = "lightblue",
		border = "white"
	)
}
```

```{r}
if ("domestic_gross" %in% names(analysis_data)) {
	hist(analysis_data$domestic_gross,
		main = "Domestic Gross Distribution",
		xlab = "Domestic Gross",
		col = "lightgreen",
		border = "white"
	)
}
```

```{r}
if ("worldwide_gross" %in% names(analysis_data)) {
	hist(analysis_data$worldwide_gross,
		main = "Worldwide Gross Distribution",
		xlab = "Worldwide Gross",
		col = "lightpink",
		border = "white"
	)
}
```

```{r}
if (all(c("production_budget", "domestic_gross", "worldwide_gross") %in% names(analysis_data))) {
	boxplot(
		analysis_data$production_budget,
		analysis_data$domestic_gross,
		analysis_data$worldwide_gross,
		names = c("Budget", "Domestic", "Worldwide"),
		main = "Boxplots for Key Numeric Variables",
		col = c("lightblue", "lightgreen", "lightpink")
	)
}
```

```{r}
if ("genre" %in% names(analysis_data)) {
	genre_tab <- sort(table(analysis_data$genre), decreasing = TRUE)
	top_n <- min(10, length(genre_tab))
	barplot(
		genre_tab[1:top_n],
		las = 2,
		col = "tan",
		main = "Top Genres",
		ylab = "Number of Movies"
	)
}
```

```{r}
if ("mpaa_rating" %in% names(analysis_data)) {
	rating_tab <- sort(table(analysis_data$mpaa_rating), decreasing = TRUE)
	barplot(
		rating_tab,
		col = "skyblue",
		main = "MPAA Rating Distribution",
		ylab = "Number of Movies"
	)
}
```

```{r}
release_dates <- analysis_data$release_date
release_year <- as.integer(format(release_dates, "%Y"))
year_tab <- table(release_year)
year_tab <- year_tab[order(as.integer(names(year_tab)))]
if (length(year_tab) > 0) {
	barplot(
		as.integer(year_tab),
		names.arg = names(year_tab),
		col = "darkorange",
		border = "white",
		las = 2,
		cex.names = 0.7,
		xlab = "Release Year",
		ylab = "Number of Movies",
		main = "Movies Released Per Year"
	)
}
```

```{r}
if ("distributor" %in% names(analysis_data)) {
	distributor_tab <- sort(table(analysis_data$distributor), decreasing = TRUE)
	top_n <- min(10, length(distributor_tab))
	barplot(
		distributor_tab[1:top_n],
		las = 2,
		cex.names = 0.8,
		col = "cornflowerblue",
		main = "Top 10 Distributors",
		ylab = "Number of Movies"
	)
}
```
