---
title: "Movie Data Analysis"
author: "TBD"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Part 1: Data Cleaning and Preparation

```{r setup, include=FALSE}
# --- Global chunk options ------------------------------------------------------
# These settings apply to all code chunks in the document:
#   - echo = FALSE      – hide R code from the rendered output
#   - message/warning   – suppress package startup messages and warnings
#   - comment = NA      – remove the leading "##" from console output
#   - fig.align/width   – centre figures and set a consistent size
#   - out.width = 100%  – make figures scale to the page width
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	comment = NA,
	fig.align = "center",
	fig.width = 10,
	fig.height = 5,
	out.width = "100%"
)

# Set the CRAN repository to avoid interactive prompts during package installation
options(repos = c(CRAN = "https://cloud.r-project.org"))

# --- Colour palette for genres -------------------------------------------------
# This named vector ensures that each genre always gets the same colour
# across all plots, regardless of the order in which genres appear.
# Any genre not listed here defaults to grey (#999999).
genre_palette <- c(
	Action = "#1f77b4",
	Adventure = "#ff7f0e",
	Comedy = "#2ca02c",
	Drama = "#9467bd",
	Horror = "#d62728"
)

# Helper function: maps a vector of genre names to the corresponding palette
# colour; unknown genres are assigned a neutral grey.
genre_color <- function(genres) {
	colors <- genre_palette[as.character(genres)]
	colors[is.na(colors)] <- "#999999"
	colors
}
```

## Load and verify the data

```{r}
# -- Load required package ------------------------------------------------------
# DT is used later for interactive tables (though not used in this part,
# it is required for the overall project; we check its presence here.)
if (!requireNamespace("DT", quietly = TRUE)) {
	stop("Package 'DT' is required. Install it once with: install.packages('DT', repos = 'https://cloud.r-project.org')")
}

# -- Read the raw CSV file ------------------------------------------------------
# The file is assumed to be in the working directory. We read it with
# stringsAsFactors = FALSE to keep character columns as strings for now.
movie_profit <- read.csv("movie_profit.csv", stringsAsFactors = FALSE)
# This makes missingness uniform across the whole dataset.
char_cols <- sapply(movie_profit, is.character)
movie_profit[char_cols] <- lapply(movie_profit[char_cols], function(x) {
  ifelse(trimws(x) == "", NA_character_, x)
})

# --- Basic dataset overview ----------------------------------------------------
# Compute simple dimensions and column type counts to get a first impression.
overview_table <- data.frame(
	Metric = c("Rows", "Columns", "Numeric columns", "Character columns"),
	Value = c(
		nrow(movie_profit),
		ncol(movie_profit),
		sum(sapply(movie_profit, is.numeric)),
		sum(sapply(movie_profit, is.character))
	)
)

knitr::kable(overview_table, caption = "Dataset overview")
```

## Check variable types and first 5 values

```{r}
# --- Inspect each variable's type and first few values -------------------------
# This helps to quickly identify columns that may need conversion (e.g. dates)
# and to detect any obvious formatting issues.
variable_overview <- data.frame(
	variable = names(movie_profit),
	type = sapply(movie_profit, function(x) class(x)[1]),
	first_5_values = sapply(movie_profit, function(x) {
		paste(head(x, 5), collapse = ", ")
	}),
	row.names = NULL
)

knitr::kable(variable_overview, caption = "Variable types and first 5 values")
```

## Convert release_date to Date type and check conversion results

```{r}
# --- Date conversion -----------------------------------------------------------
# The release_date column is read as character in MM/DD/YYYY format.
# We convert it to R's Date class using the appropriate format string.
release_date_raw <- movie_profit$release_date
movie_profit$release_date <- as.Date(release_date_raw, format = "%m/%d/%Y")

# --- Conversion summary --------------------------------------------------------
# Record the original and new classes, and count how many rows became NA
# after conversion (i.e. were invalid dates).
conversion_table <- data.frame(
	metric = c("Original type", "New type", "Invalid dates after conversion"),
	value = c(
		class(release_date_raw)[1],
		class(movie_profit$release_date)[1],
		sum(is.na(movie_profit$release_date) & !is.na(release_date_raw) & trimws(release_date_raw) != "")
	)
)

knitr::kable(conversion_table, caption = "release_date conversion summary")
```

## Check for missing values, duplicates, and logical consistency

```{r}
# --- Missing values ------------------------------------------------------------
missing_count <- colSums(is.na(movie_profit))
missing_pct   <- round(100 * missing_count / nrow(movie_profit), 2)
missing_summary <- data.frame(
  variable = names(missing_count),
  missing_count = as.integer(missing_count),
  missing_percent = missing_pct,
  row.names = NULL
)

# --- Duplicate rows ------------------------------------------------------------
# Check for exact duplicate rows; duplicates may arise from data entry errors.
dup_rows <- duplicated(movie_profit)
dup_count <- sum(dup_rows)

# --- Numeric column consistency ------------------------------------------------
# For the three financial columns, count negative values (should be none)
# and zero values (which may indicate missing or unreported data).
numeric_cols <- c("production_budget", "domestic_gross", "worldwide_gross")
neg_counts <- vapply(movie_profit[numeric_cols], function(x) sum(x < 0, na.rm = TRUE), integer(1))
zero_counts <- vapply(movie_profit[numeric_cols], function(x) sum(x == 0, na.rm = TRUE), integer(1))
numeric_checks <- data.frame(
  variable = numeric_cols,
  negative_values = neg_counts,
  zero_values = zero_counts
)

# --- Logical consistency: worldwide_gross should be >= domestic_gross ----------
# Since worldwide gross includes domestic, we expect no rows where
# worldwide_gross < domestic_gross. Such rows would indicate data errors.
bad_gross_count <- NA
if (all(c("domestic_gross", "worldwide_gross") %in% names(movie_profit))) {
	bad_gross_rows <- which(movie_profit$worldwide_gross < movie_profit$domestic_gross)
	bad_gross_count <- length(bad_gross_rows)
}

# --- Invalid dates after conversion (already computed, but shown again) --------
bad_date_count <- sum(
	is.na(movie_profit$release_date) &
		!is.na(release_date_raw) &
		trimws(release_date_raw) != ""
)

# --- Display all consistency tables --------------------------------------------
knitr::kable(missing_summary, caption = "Missing values by variable")

dup_table <- data.frame(
	metric = "Duplicate rows",
	count = dup_count
)
knitr::kable(dup_table, caption = "Duplicate row summary")

knitr::kable(numeric_checks, caption = "Numeric consistency checks")

logic_table <- data.frame(
	check = c("Rows where worldwide_gross < domestic_gross", "Invalid release_date values"),
	count = c(bad_gross_count, bad_date_count)
)
knitr::kable(logic_table, caption = "Logical consistency checks")
```

## Estimate impact of removing rows with missing values or gross = 0

```{r}
# --- Row-level missingness -----------------------------------------------------
# Before deciding on a cleaning strategy, we quantify how many rows would be
# removed under different filters. This helps us assess the trade‑off between
# data completeness and retaining sample size.

# Rows that contain at least one missing value (NA or empty string) in any column.
row_has_missing <- apply(movie_profit, 1, function(row) {
	any(is.na(row) | trimws(as.character(row)) == "")
})

# Rows where either domestic or worldwide gross is exactly zero.
# Zero gross may indicate missing data, unreported figures, or films with no
# theatrical release. For our analysis, we will exclude them because a log
# transformation would be undefined for zero, and they do not represent
# typical box-office behaviour.
gross_is_zero <-
	movie_profit$domestic_gross == 0 |
	movie_profit$worldwide_gross == 0

# Count overlaps and total rows that would be removed if we drop all such rows.
overlap_count <- sum(row_has_missing & gross_is_zero)
remove_all_count <- sum(row_has_missing | gross_is_zero)
total_movies <- nrow(movie_profit)
remove_all_pct <- round(100 * remove_all_count / total_movies, 2)

# --- Alternative: ignore missing values in mpaa_rating -------------------------
# The mpaa_rating column has a high proportion of missing values (as seen in
# the missing summary). If we choose to keep rows with missing rating but
# drop missing values elsewhere, we can retain more observations.
# We compute the removal count under this relaxed rule.
cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")
row_has_missing_no_rating <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])

remove_no_rating_count <- sum(row_has_missing_no_rating | gross_is_zero)
remove_no_rating_pct <- round(100 * remove_no_rating_count / total_movies, 2)

# --- Impact summary table ------------------------------------------------------
impact_table <- data.frame(
	metric = c(
		"Rows with at least one missing value",
		"Rows with gross value 0 (domestic or worldwide)",
		"Overlap (missing and gross = 0)",
		"Rows removed (missing or gross = 0)",
		"Rows removed (ignoring missing mpaa_rating)"
	),
	count = c(
		sum(row_has_missing),
		sum(gross_is_zero),
		overlap_count,
		remove_all_count,
		remove_no_rating_count
	),
	percent = c(
		round(100 * sum(row_has_missing) / total_movies, 2),
		round(100 * sum(gross_is_zero) / total_movies, 2),
		round(100 * overlap_count / total_movies, 2),
		remove_all_pct,
		remove_no_rating_pct
	)
)

knitr::kable(impact_table, caption = "Impact of filtering rules")
```

## Remove rows with missing values (except rating) and gross = 0

```{r}
# --- Apply cleaning rules ------------------------------------------------------
# Based on the impact assessment, we choose to drop rows with:
#   1) missing values in any column except mpaa_rating (i.e., we keep
#      rows where only rating is missing), and
#   2) domestic_gross or worldwide_gross equal to zero.
#
# This decision balances data quality with sample size, as mpaa_rating is
# often missing but is not a critical variable for the main analyses.

cols_except_rating <- setdiff(names(movie_profit), "mpaa_rating")

# Identify rows with missing values outside of mpaa_rating
row_has_NA <- !complete.cases(movie_profit[, cols_except_rating, drop = FALSE])

# Identify rows with zero gross
gross_is_zero <-
	movie_profit$domestic_gross == 0 |
	movie_profit$worldwide_gross == 0

# Keep only rows that pass both conditions
movie_profit_clean <- movie_profit[!(row_has_NA | gross_is_zero), ]

# --- Cleaning summary ----------------------------------------------------------
cleaning_table <- data.frame(
	metric = c("Rows before cleaning", "Rows after cleaning", "Rows removed"),
	count = c(nrow(movie_profit), nrow(movie_profit_clean), nrow(movie_profit) - nrow(movie_profit_clean))
)

knitr::kable(cleaning_table, caption = "Cleaning summary")
```

# Part 2: Exploratory Data Analysis

## Overview

This document covers **Part 2 — Exploratory Data Analysis** of the Movie
Data Analysis project. The goal of EDA is to understand the dataset's
structure, distributions, and relationships before building any
statistical models.

The analysis proceeds in four broad steps:

1.  **Variable description** — what each column represents\
2.  **New derived columns** — at least 10 engineered features\
3.  **Descriptive statistics** — central tendency, spread, and outliers\
4.  **Visualisations** — 13 charts with explanations of what each one
    reveals

The cleaned dataset used here was produced in Part 1. It contains
`r nrow(movie_profit_clean)` movies spanning the genres Action,
Adventure, Comedy, Drama, and Horror.

------------------------------------------------------------------------

## 1 Variable description

The table below lists every column in the cleaned dataset, its data
type, and a short description of what it measures. Understanding the raw
variables is the essential first step before any transformation or
analysis.

```{r variable-description}
# Build a human-readable description for each original column
var_desc <- data.frame(
	variable = c(
		"X", "release_date", "movie", "production_budget",
		"domestic_gross", "worldwide_gross", "distributor",
		"mpaa_rating", "genre"
	),
	type = c(
		"integer", "Date", "character", "numeric",
		"numeric", "numeric", "character",
		"character", "character"
	),
	description = c(
		"Row index from the original CSV (not analytically meaningful)",
		"Date the film was released in US cinemas (MM/DD/YYYY → Date)",
		"Film title",
		"Reported production budget in USD",
		"Total box-office receipts from US/Canada markets in USD",
		"Total box-office receipts from all markets in USD",
		"Studio or distribution company releasing the film",
		"MPAA content rating (G, PG, PG-13, R, etc.); has missing values",
		"Primary genre as classified by the data source"
	),
	stringsAsFactors = FALSE
)

knitr::kable(var_desc, caption = "Original variables in the cleaned dataset")
```

------------------------------------------------------------------------

## 2 Creating new derived columns

The original dataset provides budgets and gross revenues, but many
analytically interesting quantities — such as profitability or market
segments — must be computed. Below we create **12 new columns**
covering:

- financial indicators (international gross, ROI),\
- ranking flags (top-100 lists),\
- categorical groupings (budget bands, horror flag, movie type label),\
- scaled versions for display purposes.

```{r derived-columns}
movie_profit_enriched <- movie_profit_clean

# ── 1. International gross ────────────────────────────────────────────────────
# The non-domestic portion of worldwide revenue. Useful for understanding
# how dependent a film's success is on the international market.
movie_profit_enriched$international_gross <-
	movie_profit_enriched$worldwide_gross - movie_profit_enriched$domestic_gross

# ── 2 & 3. ROI factors ────────────────────────────────────────────────────────
# Return-on-investment expressed as a multiplier (e.g. 3.0 means the film
# earned 3× its production budget). We compute both a domestic and a
# worldwide version. Using a factor rather than a percentage makes it easy
# to say "this film returned 10× its budget."
movie_profit_enriched$roi_domestic_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, domestic_gross  / production_budget, NA_real_)
)

movie_profit_enriched$roi_worldwide_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, worldwide_gross / production_budget, NA_real_)
)

# ── 4 & 5. Top-100 flags ──────────────────────────────────────────────────────
# Boolean flags marking whether a film ranks in the top 100 by domestic or
# international gross. These provide a simple, interpretable way to identify
# "blockbuster" films without choosing an arbitrary revenue threshold.
rank_domestic      <- rank(-movie_profit_enriched$domestic_gross,      ties.method = "first")
rank_international <- rank(-movie_profit_enriched$international_gross, ties.method = "first")

movie_profit_enriched$top100_domestic_gross      <- rank_domestic      <= 100
movie_profit_enriched$top100_international_gross <- rank_international <= 100

# ── 6. Horror flag ───────────────────────────────────────────────────────────
# A logical column that is TRUE when the genre column contains "horror"
# (case-insensitive). This is the primary grouping variable for the
# Horror vs. Non-horror comparison that runs throughout the project.
movie_profit_enriched$is_horror_movie <-
	grepl("horror", movie_profit_enriched$genre, ignore.case = TRUE)

# ── 7. Movie type label ───────────────────────────────────────────────────────
# A readable character version of is_horror_movie, convenient for plot labels
# and table captions.
movie_profit_enriched$movie_type <-
	ifelse(movie_profit_enriched$is_horror_movie, "Horror", "Non-horror")

# ── 8. Budget band ────────────────────────────────────────────────────────────
# Binning production budgets into seven ordered categories allows us to
# compare how many films in each genre fall into each spending tier.
budget_breaks <- c(-Inf, 1e6, 5e6, 10e6, 20e6, 50e6, 100e6, Inf)
budget_labels <- c("< $1M", "$1M–$5M", "$5M–$10M",
                   "$10M–$20M", "$20M–$50M", "$50M–$100M", "> $100M")

movie_profit_enriched$budget_band <- cut(
	movie_profit_enriched$production_budget,
	breaks = budget_breaks,
	labels = budget_labels,
	right  = FALSE
)

# ── 9 & 10. Scaled financial columns (× $1,000) ───────────────────────────────
# Expressing large dollar amounts in thousands keeps the interactive table
# readable without sacrificing numeric sort order.
movie_profit_enriched$production_budget_k <- movie_profit_enriched$production_budget / 1000
movie_profit_enriched$domestic_gross_k    <- movie_profit_enriched$domestic_gross    / 1000
movie_profit_enriched$worldwide_gross_k   <- movie_profit_enriched$worldwide_gross   / 1000

# ── 11 & 12. Formatted date and release year ─────────────────────────────────
# A character column for display (ISO 8601) and an integer year used for
# time-series aggregations.
movie_profit_enriched$release_date_fmt <-
	ifelse(is.na(movie_profit_enriched$release_date), NA_character_,
	       format(movie_profit_enriched$release_date, "%Y-%m-%d"))

movie_profit_enriched$release_year <-
	as.integer(format(movie_profit_enriched$release_date, "%Y"))
```

**Summary of new columns**

| \# | Column | Type | Purpose |
|------------------|------------------|------------------|-------------------|
| 1 | `international_gross` | numeric | Revenue outside domestic market |
| 2 | `roi_domestic_factor` | numeric | Domestic gross ÷ production budget |
| 3 | `roi_worldwide_factor` | numeric | Worldwide gross ÷ production budget |
| 4 | `top100_domestic_gross` | logical | Top-100 film by domestic revenue |
| 5 | `top100_international_gross` | logical | Top-100 film by international revenue |
| 6 | `is_horror_movie` | logical | Genre contains "horror" |
| 7 | `movie_type` | character | Readable label ("Horror" / "Non-horror") |
| 8 | `budget_band` | factor | Seven ordered budget tiers |
| 9 | `production_budget_k` | numeric | Budget in thousands USD |
| 10 | `domestic_gross_k` | numeric | Domestic gross in thousands USD |
| 11 | `worldwide_gross_k` | numeric | Worldwide gross in thousands USD |
| 12 | `release_year` | integer | Calendar year of release |

``` {r column-preview}
# ── Preview ───────────────────────────────────────────────────────────────────
preview_cols <- c(
	"movie",
	"international_gross", "roi_domestic_factor", "roi_worldwide_factor",
	"top100_domestic_gross", "top100_international_gross",
	"is_horror_movie", "movie_type", "budget_band", "production_budget_k", "release_year"
)

preview_headers <- c(
	"Movie",
	"Intl. Gross", "ROI Dom.", "ROI World.",
	"Top 100 Dom.", "Top 100 Intl.",
	"Horror?", "Type", "Budget Band", "Prod. Bud. (K)", "Year"
)

knitr::kable(
	head(movie_profit_enriched[, preview_cols], 6),
	col.names = preview_headers,
	caption = "Preview of the 12 newly derived columns (first 6 rows)"
)
```

------------------------------------------------------------------------

## 3 Descriptive statistics

This section quantifies the central tendency, spread, and outlier
prevalence for every numeric column. Because financial variables are
heavily right-skewed (a handful of blockbusters drive the mean far above
the median), we report both measures alongside the full five-number
summary and a standard-deviation table.

```{r descriptive-stats}
analysis_data <- movie_profit_enriched

# Variables to summarise — exclude the meaningless row index
numeric_vars     <- names(analysis_data)[sapply(analysis_data, is.numeric)]
numeric_vars     <- setdiff(numeric_vars, "X")
categorical_vars <- intersect(c("genre", "mpaa_rating", "distributor"), names(analysis_data))

# Shared styling for the tables in this section.
table_class <- 'class="table table-striped table-hover table-sm report-table"'

styled_kable <- function(x, caption) {
	knitr::asis_output(
		knitr::kable(x, format = "html", caption = caption, table.attr = table_class)
	)
}

# ── Dataset profile ───────────────────────────────────────────────────────────
release_dates <- analysis_data$release_date
profile_table <- data.frame(
	metric = c(
		"Rows", "Columns", "Numeric variables",
		"Categorical variables", "Unique movies",
		"Release year range"
	),
	value = c(
		nrow(analysis_data),
		ncol(analysis_data),
		length(numeric_vars),
		length(categorical_vars),
		length(unique(analysis_data$movie)),
		paste(format(min(release_dates), "%Y"), "to", format(max(release_dates), "%Y"))
	),
	stringsAsFactors = FALSE
)
styled_kable(profile_table, caption = "Analysis dataset profile")

# ── Five-number summary for numeric columns ───────────────────────────────────
summary_vars <- setdiff(
	numeric_vars,
	c("production_budget_k", "domestic_gross_k", "worldwide_gross_k")
)

pretty_var_names <- c(
	production_budget = "Production budget",
	domestic_gross = "Domestic gross",
	worldwide_gross = "Worldwide gross",
	international_gross = "International gross",
	roi_domestic_factor = "Domestic ROI",
	roi_worldwide_factor = "Worldwide ROI",
	release_year = "Release year"
)

num_summary_mat <- do.call(
	cbind,
	lapply(analysis_data[, summary_vars, drop = FALSE], summary)
)
num_summary <- data.frame(
	statistic   = rownames(num_summary_mat),
	num_summary_mat,
	row.names   = NULL,
	check.names = FALSE
)
pretty_summary_names <- c(
	statistic = "Statistic",
	production_budget = "Production budget",
	domestic_gross = "Domestic gross",
	worldwide_gross = "Worldwide gross",
	international_gross = "International gross",
	roi_domestic_factor = "Domestic ROI",
	roi_worldwide_factor = "Worldwide ROI",
	release_year = "Release year"
)

colnames(num_summary) <- unname(ifelse(
	names(num_summary) %in% names(pretty_summary_names),
	pretty_summary_names[names(num_summary)],
	tools::toTitleCase(gsub("_", " ", names(num_summary)))
))

for (col in names(num_summary)[-1]) {
	num_summary[[col]] <- format(
		round(as.numeric(num_summary[[col]]), 2),
		big.mark   = ",",
		scientific = FALSE,
		trim       = TRUE
	)
}
styled_kable(num_summary, caption = "Summary statistics for key numeric variables")

# ── Standard deviations ───────────────────────────────────────────────────────
sd_table <- data.frame(
	variable  = summary_vars,
	sd        = round(sapply(analysis_data[, summary_vars, drop = FALSE], sd, na.rm = TRUE), 2),
	row.names = NULL
)
sd_table$variable <- ifelse(
	sd_table$variable %in% names(pretty_var_names),
	pretty_var_names[sd_table$variable],
	sd_table$variable
)
sd_table$sd <- format(sd_table$sd, big.mark = ",", scientific = FALSE, trim = TRUE)
styled_kable(sd_table, caption = "Standard deviation by variable")

# ── IQR-based outlier count ───────────────────────────────────────────────────
# An observation is flagged as an outlier if it falls more than 1.5 × IQR
# below Q1 or above Q3 — the classic Tukey fence rule.
outlier_stats <- do.call(rbind, lapply(summary_vars, function(v) {
	x   <- analysis_data[[v]]
	q1  <- quantile(x, 0.25, na.rm = TRUE)
	q3  <- quantile(x, 0.75, na.rm = TRUE)
	iqr <- q3 - q1
	lo  <- x < (q1 - 1.5 * iqr)
	hi  <- x > (q3 + 1.5 * iqr)
	data.frame(
		variable           = v,
		outlier_count      = sum(lo | hi, na.rm = TRUE),
		outlier_percent    = round(100 * sum(lo | hi, na.rm = TRUE) / sum(!is.na(x)), 2),
		low_outlier_count  = sum(lo, na.rm = TRUE),
		high_outlier_count = sum(hi, na.rm = TRUE)
	)
}))
outlier_stats$variable <- ifelse(
	outlier_stats$variable %in% names(pretty_var_names),
	pretty_var_names[outlier_stats$variable],
	outlier_stats$variable
)
styled_kable(outlier_stats, caption = "IQR outlier summary (Tukey 1.5 × IQR rule)")

# ── Categorical summaries ─────────────────────────────────────────────────────
pretty_cat_names <- c(
	genre = "Genre",
	mpaa_rating = "MPAA rating",
	distributor = "Distributor"
)

cat_table <- do.call(rbind, lapply(categorical_vars, function(v) {
	x         <- analysis_data[[v]]
	x_valid   <- x[!(is.na(x) | trimws(x) == "")]
	tab       <- sort(table(x_valid), decreasing = TRUE)
	data.frame(
		variable          = ifelse(v %in% names(pretty_cat_names), pretty_cat_names[v], v),
		unique_levels     = length(unique(x_valid)),
		most_common       = trimws(names(tab)[1]),
		most_common_count = as.integer(tab[1])
	)
}))
names(cat_table) <- c("Variable", "Unique levels", "Most common", "Count")
styled_kable(cat_table, caption = "Summary of categorical variables")


compare_data <- movie_profit_enriched

# Helper: count IQR-based outliers in a numeric vector
count_iqr_outliers <- function(x) {
	x <- x[!is.na(x)]
	if (length(x) < 4) return(0)
	q1  <- quantile(x, 0.25)
	q3  <- quantile(x, 0.75)
	iqr <- q3 - q1
	sum(x < (q1 - 1.5 * iqr) | x > (q3 + 1.5 * iqr))
}

# Build summary for each group
build_group_stats <- function(flag) {
	d <- compare_data[flag, ]
	c(
		movie_count            = sum(flag),
		budget_mean            = round(mean(d$production_budget,   na.rm = TRUE), 2),
		budget_q1              = round(quantile(d$production_budget,   0.25, na.rm = TRUE), 2),
		budget_median          = round(median(d$production_budget,  na.rm = TRUE), 2),
		budget_q3              = round(quantile(d$production_budget,   0.75, na.rm = TRUE), 2),
		worldwide_mean         = round(mean(d$worldwide_gross,     na.rm = TRUE), 2),
		worldwide_q1           = round(quantile(d$worldwide_gross,     0.25, na.rm = TRUE), 2),
		worldwide_median       = round(median(d$worldwide_gross,   na.rm = TRUE), 2),
		worldwide_q3           = round(quantile(d$worldwide_gross,     0.75, na.rm = TRUE), 2),
		roi_domestic_mean      = round(mean(d$roi_domestic_factor, na.rm = TRUE), 2),
		roi_domestic_q1        = round(quantile(d$roi_domestic_factor, 0.25, na.rm = TRUE), 2),
		roi_domestic_median    = round(median(d$roi_domestic_factor, na.rm = TRUE), 2),
		roi_domestic_q3        = round(quantile(d$roi_domestic_factor, 0.75, na.rm = TRUE), 2),
		roi_domestic_outliers  = count_iqr_outliers(d$roi_domestic_factor),
		roi_worldwide_mean     = round(mean(d$roi_worldwide_factor, na.rm = TRUE), 2),
		roi_worldwide_q1       = round(quantile(d$roi_worldwide_factor, 0.25, na.rm = TRUE), 2),
		roi_worldwide_median   = round(median(d$roi_worldwide_factor, na.rm = TRUE), 2),
		roi_worldwide_q3       = round(quantile(d$roi_worldwide_factor, 0.75, na.rm = TRUE), 2),
		roi_worldwide_outliers = count_iqr_outliers(d$roi_worldwide_factor)
	)
}

horror_stats     <- build_group_stats(compare_data$is_horror_movie)
non_horror_stats <- build_group_stats(!compare_data$is_horror_movie)

pretty_metric_names <- c(
	"movie_count"            = "Movie count",
	"budget_mean"            = "Production budget (mean)",
	"budget_q1.25%"          = "Production budget (Q1)",
	"budget_median"          = "Production budget (median)",
	"budget_q3.75%"          = "Production budget (Q3)",
	"worldwide_mean"         = "Worldwide gross (mean)",
	"worldwide_q1.25%"       = "Worldwide gross (Q1)",
	"worldwide_median"       = "Worldwide gross (median)",
	"worldwide_q3.75%"       = "Worldwide gross (Q3)",
	"roi_domestic_mean"      = "Domestic ROI (mean)",
	"roi_domestic_q1.25%"    = "Domestic ROI (Q1)",
	"roi_domestic_median"    = "Domestic ROI (median)",
	"roi_domestic_q3.75%"    = "Domestic ROI (Q3)",
	"roi_domestic_outliers"  = "Domestic ROI (outliers)",
	"roi_worldwide_mean"     = "Worldwide ROI (mean)",
	"roi_worldwide_q1.25%"   = "Worldwide ROI (Q1)",
	"roi_worldwide_median"   = "Worldwide ROI (median)",
	"roi_worldwide_q3.75%"   = "Worldwide ROI (Q3)",
	"roi_worldwide_outliers" = "Worldwide ROI (outliers)"
)

summary_table <- data.frame(
	metric      = ifelse(names(horror_stats) %in% names(pretty_metric_names),
	              pretty_metric_names[names(horror_stats)],
	              names(horror_stats)),
	Horror      = unname(horror_stats),
	`Non-horror` = unname(non_horror_stats),
	check.names = FALSE
)

for (col in c("Horror", "Non-horror")) {
	summary_table[[col]] <- format(summary_table[[col]],
	    big.mark = ",", scientific = FALSE, trim = TRUE)
}

styled_kable(summary_table, caption = "Horror vs. Non-horror — key financial metrics")
```

------------------------------------------------------------------------

## 4 Visualisations

Each subsection below presents one or more related charts. For every
chart we state:

- **Why** we created it — the analytical question it addresses.\
- **What** it shows — a plain-language reading of the main visual
  patterns.\
- **How** it helps — what we learn about the dataset from this view.

------------------------------------------------------------------------

### Chart 1 — Genre distribution

**Why:** Genre is the most important categorical variable in this
dataset. Before any comparison we need to know how many films each genre
contributes, to be aware of sample-size differences that could affect
the reliability of genre-level statistics.

**What:** Drama and Comedy are by far the most represented genres, while
Horror and Adventure form a much smaller share of the catalogue.

**How:** An imbalanced genre distribution means that averages for under-
represented genres (especially Horror) are based on fewer observations
and may be more volatile. We should keep this in mind when interpreting
the Horror vs. Non-horror comparisons later.

```{r chart1-genre}
genre_tab <- sort(table(analysis_data$genre), decreasing = TRUE)
top_n     <- min(10, length(genre_tab))

barplot(
	genre_tab[1:top_n],
	las  = 2,
	col  = genre_color(names(genre_tab[1:top_n])),
	main = "Chart 1 — Number of Movies by Genre",
	ylab = "Number of Movies"
)
```

------------------------------------------------------------------------

### Chart 2 — MPAA rating distribution

**Why:** The MPAA rating (G, PG, PG-13, R) restricts the audience a film
can reach and therefore influences its potential box-office ceiling.
Understanding the rating mix helps contextualise revenue comparisons.

**What:** R-rated films dominate the dataset, followed by PG-13. G and
PG films are relatively rare.

**How:** Since R-rated films are the most common, any genre that skews
heavily toward R (such as Horror) will have its revenue compared against
a baseline where R films are the norm — which is useful to know when we
later model the effect of genre on revenue.

```{r chart2-mpaa}
rating_tab <- sort(table(analysis_data$mpaa_rating), decreasing = TRUE)
barplot(
	rating_tab,
	col  = "skyblue",
	main = "Chart 2 — MPAA Rating Distribution",
	ylab = "Number of Movies"
)
```

------------------------------------------------------------------------

### Chart 3 — Movies released per year

**Why:** The temporal coverage of a dataset matters: if most films come
from a short recent period, earlier decades are under-represented and
long-run trend analyses could be misleading.

**What:** Production volume is low in the 1970s and rises steadily
through the 1980s and 1990s, with a notable increase from the 2000s
onward before plateauing.

**How:** The dataset has reasonable coverage from the late 1970s through
the mid-2010s. Years with fewer than \~10 films (mainly pre-1980) should
be treated with caution in any year-by-year comparison.

```{r chart3-yearly}
year_tab <- table(analysis_data$release_year)
year_tab <- year_tab[order(as.integer(names(year_tab)))]

barplot(
	as.integer(year_tab),
	names.arg = names(year_tab),
	col       = "darkorange",
	border    = "white",
	las       = 2,
	cex.names = 0.7,
	xlab      = "Release Year",
	ylab      = "Number of Movies",
	main      = "Chart 3 — Movies Released Per Year"
)
```

------------------------------------------------------------------------

### Chart 4 — Top 10 distributors

**Why:** A few large studios dominate film distribution. Knowing which
distributors are best-represented in the data helps us understand
whether the dataset reflects the full market or mainly one segment of
it.

**What:** The major Hollywood studios (e.g. Warner Bros., Universal,
Paramount) account for the vast majority of films.

**How:** Because large distributors handle films across many genres,
their dominance in the data suggests that the dataset is broadly
representative of mainstream studio output rather than, say, indie or
international cinema.

```{r chart4-distributor}
dist_tab <- sort(table(analysis_data$distributor), decreasing = TRUE)
top_n    <- min(10, length(dist_tab))

barplot(
	dist_tab[1:top_n],
	las       = 2,
	cex.names = 0.8,
	col       = "cornflowerblue",
	main      = "Chart 4 — Top 10 Distributors by Number of Movies",
	ylab      = "Number of Movies"
)
```

------------------------------------------------------------------------

### Charts 5 & 6 — Horror and Non-horror film counts per year

**Why:** Before comparing financial performance over time we first need
to see whether both groups have similar temporal coverage. If Horror
films are concentrated in a particular decade, that era's economic
conditions (rather than genre) might explain observed ROI differences.

**What:** Horror film counts per year fluctuate considerably, with peaks
in the early 1980s and again from the mid-2000s onward. Non-horror
counts follow a smoother upward trend driven by the overall growth of
the industry.

**How:** Because both groups span the full date range of the dataset,
the time-series ROI comparisons in Charts 7 and 8 are not confounded by
one group being systematically older or newer.

```{r chart5-horror-yearly}
valid_rows        <- !is.na(compare_data$release_year) & compare_data$release_year >= 1975
horror_year_tab   <- table(compare_data$release_year[valid_rows &  compare_data$is_horror_movie])
horror_year_tab   <- horror_year_tab[order(as.integer(names(horror_year_tab)))]
year_range        <- range(compare_data$release_year[valid_rows], na.rm = TRUE)

plot(
	as.integer(names(horror_year_tab)),
	as.integer(horror_year_tab),
	type = "l", lwd = 2, col = "firebrick",
	xlim = year_range,
	xlab = "Release Year", ylab = "Number of Horror Movies",
	main = "Chart 5 — Horror Movies per Year"
)
```

```{r chart6-nonhorror-yearly}
non_horror_year_tab <- table(compare_data$release_year[valid_rows & !compare_data$is_horror_movie])
non_horror_year_tab <- non_horror_year_tab[order(as.integer(names(non_horror_year_tab)))]

plot(
	as.integer(names(non_horror_year_tab)),
	as.integer(non_horror_year_tab),
	type = "l", lwd = 2, col = "steelblue",
	xlim = year_range,
	xlab = "Release Year", ylab = "Number of Non-horror Movies",
	main = "Chart 6 — Non-horror Movies per Year"
)
```

------------------------------------------------------------------------

### Chart 7 — Average ROI per year: Horror vs. Non-horror

**Why:** Comparing yearly *average* ROI between the two groups reveals
whether Horror's reputedly high return is a persistent structural trait
or a feature of specific years or eras.

**What:** Horror tends to have high average ROI in years where one
breakout low-budget hit (e.g. *Blair Witch Project*, *Paranormal
Activity*) inflates the mean dramatically. The y-axis is capped at 20×
and any year exceeding that cap is marked with a triangle; the actual
values are printed below the chart.

**How:** The dramatic spikes in Horror average ROI are largely driven by
outlier blockbusters rather than a consistently higher return across all
films. This motivates the median comparison in Chart 8, which is more
robust to such outliers.

```{r chart7-avg-roi}
roi_vals  <- compare_data$roi_worldwide_factor
vr        <- !is.na(compare_data$release_year) & !is.na(roi_vals)
all_years <- sort(unique(compare_data$release_year[vr]))
all_years <- all_years[all_years >= 1975]

# Compute yearly means for each group
avg_roi_fn <- function(flag) {
	tapply(roi_vals[vr & flag], compare_data$release_year[vr & flag], mean)
}
horror_avg_by_year     <- avg_roi_fn(compare_data$is_horror_movie)
non_horror_avg_by_year <- avg_roi_fn(!compare_data$is_horror_movie)

h_series  <- as.numeric(horror_avg_by_year[as.character(all_years)])
nh_series <- as.numeric(non_horror_avg_by_year[as.character(all_years)])

y_cap <- 20

plot_capped_series <- function(h, nh, years, cap, ylabel, title_str) {
	h_plot  <- pmin(h,  cap)
	nh_plot <- pmin(nh, cap)
	h_over  <- which(!is.na(h)  & h  > cap)
	nh_over <- which(!is.na(nh) & nh > cap)

	plot(years, h_plot, type = "l", lwd = 2, col = "firebrick",
	     xlab = "Release Year", ylab = ylabel, main = title_str,
	     ylim = c(0, cap))
	lines(years, nh_plot, lwd = 2, col = "steelblue")

	if (length(h_over)  > 0) points(years[h_over],  rep(cap, length(h_over)),
	                                 pch = 24, bg = "firebrick",  col = "firebrick",  cex = 0.8)
	if (length(nh_over) > 0) points(years[nh_over], rep(cap, length(nh_over)),
	                                 pch = 24, bg = "steelblue", col = "steelblue", cex = 0.8)

	legend("topright",
	       legend = c("Horror", "Non-horror", paste0("Value > ", cap, " (triangle)")),
	       col    = c("firebrick", "steelblue", "black"),
	       lwd    = c(2, 2, NA), pch = c(NA, NA, 24),
	       pt.bg  = c(NA, NA, "gray60"), bty = "n")

	note_fn <- function(idx, series, label) {
		if (length(idx) == 0) return(invisible(NULL))
		cat(label, paste0(years[idx], " (", round(series[idx], 2), "x)", collapse = ", "), "\n")
	}
	cat(paste0("Years where average ROI exceeded ", cap, "x:\n"))
	note_fn(h_over,  h,  "  Horror:     ")
	note_fn(nh_over, nh, "  Non-horror: ")
}

plot_capped_series(
	h_series, nh_series, all_years, y_cap,
	ylabel     = "Average ROI (Worldwide)",
	title_str  = "Chart 7 — Average Worldwide ROI per Year: Horror vs. Non-horror (capped at 20×)"
)
```

------------------------------------------------------------------------

### Chart 8 — Median ROI per year: Horror vs. Non-horror

**Why:** The median is resistant to extreme outliers; it tells us what a
*typical* film in each group earned relative to its budget in a given
year. Comparing Charts 7 and 8 shows how much of Horror's apparent
advantage is driven by a few extreme hits.

**What:** The median Horror ROI is consistently above the Non-horror
median in most years, even after removing the influence of outliers. The
gap narrows in recent decades as blockbuster budgets across genres
increased.

**How:** The persistent median advantage for Horror confirms that its
higher ROI is a structural pattern, not purely an outlier phenomenon.
Low budgets combined with audience demand for scare entertainment appear
to create a favourable risk-reward profile.

```{r chart8-median-roi}
med_roi_fn <- function(flag) {
	tapply(roi_vals[vr & flag], compare_data$release_year[vr & flag], median)
}
horror_med_by_year     <- med_roi_fn(compare_data$is_horror_movie)
non_horror_med_by_year <- med_roi_fn(!compare_data$is_horror_movie)

h_med  <- as.numeric(horror_med_by_year[as.character(all_years)])
nh_med <- as.numeric(non_horror_med_by_year[as.character(all_years)])

plot_capped_series(
	h_med, nh_med, all_years, y_cap,
	ylabel    = "Median ROI (Worldwide)",
	title_str = "Chart 8 — Median Worldwide ROI per Year: Horror vs. Non-horror (capped at 20×)"
)
```

------------------------------------------------------------------------

### Interactive table — all movies

**Why:** Summary statistics and aggregated charts can obscure the
individual films that drive the patterns. An interactive, filterable
table allows us to inspect any movie directly, verify specific claims,
and explore genre or distributor sub-groups on demand.

**What:** The table shows title, genre, worldwide ROI, production cost,
domestic and worldwide gross (all financial columns in thousands of
USD), release date, MPAA rating, and distributor. It is sortable on
every column and filterable by genre via the drop-down above the table.

**How:** Sorting by ROI (descending) immediately reveals the Horror
films at the top of the rankings. Sorting by production budget
(ascending) highlights the very cheap Horror films that achieve
extraordinary returns — an observation we quantify in Charts 10 and 11.

```{r interactive-table}
display_table <- movie_profit_enriched[, c(
	"movie", "genre", "roi_worldwide_factor",
	"production_budget_k", "domestic_gross_k", "worldwide_gross_k",
	"release_date_fmt", "mpaa_rating", "distributor"
)]

names(display_table) <- c(
	"Movie", "Genre", "ROI (×)",
	"Production Cost (×$1k)", "Domestic Gross (×$1k)", "Worldwide Gross (×$1k)",
	"Release Date", "MPAA Rating", "Distributor"
)

display_table$`ROI (×)`                <- round(display_table$`ROI (×)`,                2)
display_table$`Production Cost (×$1k)` <- round(display_table$`Production Cost (×$1k)`, 1)
display_table$`Domestic Gross (×$1k)`  <- round(display_table$`Domestic Gross (×$1k)`,  1)
display_table$`Worldwide Gross (×$1k)` <- round(display_table$`Worldwide Gross (×$1k)`, 1)

DT::datatable(
	display_table,
	rownames = FALSE,
	options  = list(
		pageLength = 15,
		lengthMenu = c(10, 15, 25, 50),
		order      = list(list(5, "desc")),
		autoWidth  = TRUE
	),
	callback = DT::JS(
		"var genreColumn = 1;",
		"var wrapper = $(table.table().container());",
		"var controls = $('<div style=\"margin-bottom: 8px;\"></div>');",
		"var label    = $('<label style=\"margin-right: 8px; font-weight: 600;\">Genre:</label>');",
		"var select   = $('<select style=\"padding: 2px 6px;\"><option value=\"\">All genres</option></select>');",
		"select.on('change', function() {",
		"  var val = $.fn.dataTable.util.escapeRegex($(this).val());",
		"  table.column(genreColumn).search(val ? '^' + val + '$' : '', true, false).draw();",
		"});",
		"table.column(genreColumn).data().unique().sort().each(function(d) {",
		"  if (d !== null && d !== undefined && d !== '') {",
		"    select.append($('<option></option>').attr('value', d).text(d));",
		"  }",
		"});",
		"controls.append(label).append(select);",
		"wrapper.prepend(controls);"
	),
	caption = "All movies — sortable and filterable (Genre drop-down above table; financial columns in thousands USD)"
)
```

**Key observations from the table:**

- **ROI leadership:** Horror is significantly over-represented at the
  top of the ROI rankings; 5 of the top 10 films by worldwide ROI are
  Horror titles (including rank 1 and 2).
- **Early advantage:** Most Horror films released before 1982 show a ROI
  above 10×, whereas non-Horror films from the same era average
  considerably lower.
- **Number of low-budget movies:** When sorted by production cost, 20
  Horror films cost ≤ \$1M. The counts for other genres are: Action (4),
  Adventure (2), Comedy (28), Drama (69).
- **Low risk, high reward:** Of the 20 cheapest Horror films (≤ \$1M), 5
  achieve a ROI above 100× and 15 achieve at least 8×. Comparable Drama
  films at that budget level reach ≥ 5× ROI in only about half of cases.

------------------------------------------------------------------------

### Chart 9 — Top 20 movies by worldwide ROI

**Why:** Identifying the individual films with the highest ROI provides
concrete examples of the outlier blockbusters that drive the mean ROI
spikes we saw in Chart 7.

**What:** Horror titles (red) dominate the upper portion of the chart,
with Comedy (green) and Drama (purple) also represented. The leading
films have ROI values far above 100×.

**How:** Seeing the actual titles and their genres confirms that
Horror's ROI advantage is driven by real, well-known films with
extremely low budgets and unexpectedly large box-office receipts — not
by data artefacts.

```{r chart9-top20-roi, fig.height=7}
roi_plot_data <- compare_data[!is.na(compare_data$roi_worldwide_factor), ]
roi_plot_data <- roi_plot_data[order(-roi_plot_data$roi_worldwide_factor), ]
roi_plot_data <- head(roi_plot_data, 20)
roi_plot_data <- roi_plot_data[order(roi_plot_data$roi_worldwide_factor), ]  # ascending for barplot

roi_colors <- genre_color(roi_plot_data$genre)

par(mar = c(5, 12, 4, 2))
roi_bars <- barplot(
	roi_plot_data$roi_worldwide_factor,
	horiz     = TRUE, las = 1,
	col       = roi_colors, border = "white",
	xlab      = "ROI (×)",
	main      = "Chart 9 — Top 20 Movies by Worldwide ROI",
	names.arg = roi_plot_data$movie,
	cex.names = 0.65
)
text(
	x      = roi_plot_data$roi_worldwide_factor,
	y      = roi_bars,
	labels = round(roi_plot_data$roi_worldwide_factor, 1),
	pos    = 4, cex = 0.75
)
legend("bottomright", legend = names(genre_palette),
       fill = genre_palette, bty = "n", cex = 0.8)
```

------------------------------------------------------------------------

### Chart 10 — ROI of the earliest 50 films per genre (scatter plots)

**Why:** If early films systematically had higher ROI than recent ones,
the Horror advantage might simply reflect the fact that Horror films
were made earlier when production costs were lower relative to
box-office revenue.

**What:** All five genre panels are shown on the same y-axis scale to
allow direct comparison. In each panel, points above the dashed line
(ROI = 1) recouped their budget; points below it were loss-makers.

**How:** The scatter plots show that very high ROI outliers exist in
multiple genres in the early era, but Horror has a higher density of
extreme high-ROI points — particularly in the 1970s and 1980s. This
suggests that Horror's advantage in the early era is real and not purely
a function of low production costs across the board.

```{r chart10-earliest50, fig.width=14, fig.height=12}
genres_to_plot   <- intersect(names(genre_palette), unique(compare_data$genre))
earliest_by_genre <- lapply(genres_to_plot, function(g) {
	gd <- compare_data[
		compare_data$genre == g &
		!is.na(compare_data$release_date) &
		!is.na(compare_data$roi_worldwide_factor), ]
	gd <- gd[order(gd$release_date), ]
	head(gd, 50)
})
names(earliest_by_genre) <- genres_to_plot

max_roi <- max(unlist(lapply(earliest_by_genre, function(x) x$roi_worldwide_factor)), na.rm = TRUE)
y_limit <- c(0, if (is.finite(max_roi)) max_roi else 1)

old_par <- par(mfrow = c(3, 2), mar = c(4, 4, 3, 1))
for (g in names(earliest_by_genre)) {
	gd <- earliest_by_genre[[g]]
	plot(gd$release_date, gd$roi_worldwide_factor,
	     pch = 16, col = genre_palette[[g]],
	     xlab = "Release Date", ylab = "ROI (×)",
	     main = paste0("Chart 10 — ", g, " (earliest 50 films)"),
	     ylim = y_limit)
	abline(h = 1, lty = 3, col = "gray70")
}
# Fill any empty panels in the 3×2 grid
if (length(genres_to_plot) < 6) for (i in seq_len(6 - length(genres_to_plot))) plot.new()
par(old_par)
```

------------------------------------------------------------------------

### Chart 11 — Low-budget films (≤ \$1M) by genre

**Why:** Horror is often described as a "cheap-to-make" genre. This
chart quantifies that claim by counting how many films in each genre
were produced on a budget of \$1M or less — and what percentage of each
genre's total output that represents.

**What:** In absolute terms, Drama has the most low-budget films, but
Horror has a proportionally high share: roughly 10% of Horror films cost
≤ \$1M, compared to much lower percentages for Action and Adventure.

**How:** The high proportion of micro-budget Horror films is a key
structural feature of the genre. These films carry very low financial
risk and, as seen in Chart 9, can generate exceptional returns — which
explains why studios and independent producers alike continue to
greenlight them.

```{r chart11-low-budget}
budget_threshold <- 1e6
genre_order      <- intersect(names(genre_palette), unique(compare_data$genre))

low_budget_count   <- sapply(genre_order, function(g) {
	sum(compare_data$genre == g &
	    !is.na(compare_data$production_budget) &
	    compare_data$production_budget <= budget_threshold)
})
genre_total_count  <- sapply(genre_order, function(g) sum(compare_data$genre == g))
low_budget_pct     <- round(100 * low_budget_count / genre_total_count, 1)

low_budget_table   <- data.frame(
	genre              = genre_order,
	low_budget_count   = as.integer(low_budget_count),
	total_genre_movies = as.integer(genre_total_count),
	low_budget_percent = low_budget_pct,
	row.names          = NULL
)

base_colors    <- genre_color(low_budget_table$genre)
bar_colors_mat <- rbind(base_colors, adjustcolor(base_colors, alpha.f = 0.45))

max_count      <- max(low_budget_table$low_budget_count)
y_max          <- max_count * 1.15
scale_factor   <- y_max / 10   # maps 0–10% onto the y-axis

bar_matrix <- rbind(
	Count      = low_budget_table$low_budget_count,
	Percentage = low_budget_table$low_budget_percent * scale_factor
)

old_par       <- par(mar = c(7, 5, 4, 5))
bar_positions <- barplot(
	bar_matrix,
	beside    = TRUE, col = bar_colors_mat, border = "white",
	las = 2, cex.names = 0.8,
	ylab      = "Number of Movies",
	main      = "Chart 11 — Low-Budget Films (≤$1M) by Genre",
	names.arg = low_budget_table$genre,
	ylim      = c(0, y_max)
)

text(bar_positions[1, ], low_budget_table$low_budget_count,
     labels = low_budget_table$low_budget_count,
     pos = 3, cex = 0.75, font = 2)
text(bar_positions[2, ], low_budget_table$low_budget_percent * scale_factor,
     labels = paste0(low_budget_table$low_budget_percent, "%"),
     pos = 3, cex = 0.75, font = 2)

axis(4, at = seq(0, y_max, length.out = 6), labels = seq(0, 10, by = 2),
     col = "gray50", col.axis = "gray50")
mtext("Percentage of Genre Total (%)", side = 4, line = 3, col = "gray50", cex = 0.9)

legend_colors <- c(base_colors[1], adjustcolor(base_colors[1], alpha.f = 0.45))
legend("topleft", legend = c("Count (≤$1M)", "% of genre total"),
       fill = legend_colors, bty = "n", cex = 0.9)
par(old_par)
```

------------------------------------------------------------------------

### Chart 12 — Budget vs. worldwide revenue (250 lowest-budget films)

**Why:** Focusing on the cheapest films and comparing their budgets to
their revenues directly tests whether low-budget films can "punch above
their weight" — i.e. whether a tiny investment can generate
disproportionately large returns.

**What:** At very low budgets (\< \$0.5M) several films — mostly Horror
and Comedy — achieved revenues of tens of millions of dollars, implying
extraordinary ROI. However, many films in the same budget range earned
little.

**How:** The chart illustrates the high-risk, high-reward nature of
micro-budget filmmaking. While the best performers yield exceptional
returns, a substantial portion of low-budget films earns far less than
expected. Horror stands out because its ratio of high earners to low
earners is unusually favourable.

```{r chart12-budget-vs-revenue}
lb_data <- compare_data[
	!is.na(compare_data$production_budget) &
	!is.na(compare_data$worldwide_gross), ]
lb_data <- lb_data[order(lb_data$production_budget), ]
lb_data <- head(lb_data, 250)

plot_colors <- genre_color(lb_data$genre)
x_vals <- lb_data$production_budget / 1e6
y_vals <- lb_data$worldwide_gross   / 1e6

# Slight jitter on the x-axis so stacked points remain visible
x_jitter <- jitter(x_vals, amount = diff(range(x_vals, na.rm = TRUE)) * 0.003)

plot(
	x_jitter, y_vals,
	pch  = 16, col  = plot_colors,
	xlab = "Production Budget ($ millions)",
	ylab = "Worldwide Revenue ($ millions)",
	main = "Chart 12 — 250 Lowest-Budget Films: Budget vs. Revenue",
	cex  = 1.2
)
legend("topleft", legend = names(genre_palette),
       col = genre_palette, pch = 16, bty = "n", cex = 0.85)
```

------------------------------------------------------------------------

### Chart 13 — Budget band distribution across genres

**Why:** Beyond the micro-budget segment, we want to understand the full
budget spectrum for each genre. Are some genres exclusively mid-budget,
or does every genre span the full range?

**What:** Action and Adventure are heavily concentrated in the
high-budget tiers (\> \$50M and \> \$100M), whereas Horror is
overwhelmingly represented in the lower tiers (\< \$10M). Comedy and
Drama span a wider range.

**How:** The stark contrast in budget profiles confirms that Horror is a
structurally different type of production compared to action-oriented
genres. This budget-structure difference is central to understanding why
Horror's ROI behaves differently — something we will model formally in
Part 3.

```{r chart13-budget-bands, fig.height=6}
budget_data  <- compare_data[
	!is.na(compare_data$production_budget) &
	!is.na(compare_data$genre), ]

budget_data$budget_band <- cut(
	budget_data$production_budget,
	breaks = budget_breaks,
	labels = budget_labels,
	right  = FALSE
)

band_count_table <- table(
	factor(budget_data$budget_band, levels = budget_labels),
	factor(budget_data$genre,       levels = genre_order)
)

band_colors <- c("#2E86AB", "#A23B72", "#F18F01", "#FF7043",
                 "#4CAF50", "#6C757D", "#9C27B0")

old_par       <- par(mar = c(7, 5, 4, 1))
bar_positions <- barplot(
	band_count_table,
	beside    = TRUE, col = band_colors, border = "white",
	las = 2, cex.names = 0.9,
	ylab      = "Movie Count",
	main      = "Chart 13 — Budget Band Distribution Across Genres",
	names.arg = genre_order,
	ylim      = c(0, max(band_count_table) * 1.15)
)
text(
	x      = as.vector(bar_positions),
	y      = as.vector(band_count_table),
	labels = as.vector(band_count_table),
	pos    = 3, cex = 0.65
)
legend("topright", legend = budget_labels, fill = band_colors,
       title = "Budget tier", bty = "n", cex = 0.85)
par(old_par)
```

------------------------------------------------------------------------

## 5 Key findings from EDA

The exploratory analysis surfaces three main findings that will motivate
the modelling in Part 3:

1.  **Low-budget Horror films produce unusually high ROI.** The top of
    the worldwide ROI ranking is dominated by Horror films that cost
    very little to produce (often \< \$1M) but earned tens or hundreds
    of millions at the box office.

2.  **Horror's ROI advantage is structural, not incidental.** Both the
    median ROI comparison (Chart 8) and the budget-band distribution
    (Chart 13) confirm that Horror operates in a different financial
    regime from Action, Adventure, and Drama. Even after removing
    outliers, the typical Horror film returns more per dollar of
    production investment than its non-Horror counterpart.

3.  **Higher budgets help most genres but not Horror as much.** The
    interactive table and Chart 12 suggest that increasing the budget of
    a Horror film does not proportionally increase its revenue — a
    hypothesis we will test formally using regression in Part 3. \`\`\`

# Part 3: Modeling

```{r enrich, include=FALSE}
# ── Feature engineering (repeated from Part 2 so this file is self-contained) ──
#
# We recreate the derived columns used in Part 2 that the models below depend on:
#   • international_gross   – box-office revenue outside the US/Canada
#   • roi_domestic_factor   – domestic_gross / production_budget
#   • roi_worldwide_factor  – worldwide_gross / production_budget
#   • top100_domestic_gross / top100_international_gross – binary rank flags
#   • is_horror_movie       – TRUE for the Horror genre
#   • movie_type            – "Horror" / "Non-horror" label for grouping

movie_profit_enriched <- movie_profit_clean

movie_profit_enriched$international_gross <- with(
	movie_profit_enriched,
	worldwide_gross - domestic_gross
)

movie_profit_enriched$roi_domestic_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, domestic_gross / production_budget, NA_real_)
)

movie_profit_enriched$roi_worldwide_factor <- with(
	movie_profit_enriched,
	ifelse(production_budget > 0, worldwide_gross / production_budget, NA_real_)
)

rank_domestic      <- rank(-movie_profit_enriched$domestic_gross,      ties.method = "first")
rank_international <- rank(-movie_profit_enriched$international_gross,  ties.method = "first")

movie_profit_enriched$top100_domestic_gross      <- rank_domestic      <= 100
movie_profit_enriched$top100_international_gross <- rank_international <= 100

movie_profit_enriched$is_horror_movie <- grepl(
	"horror", movie_profit_enriched$genre, ignore.case = TRUE
)

movie_profit_enriched$movie_type <- ifelse(
	movie_profit_enriched$is_horror_movie, "Horror", "Non-horror"
)

# Shorthand used throughout this part
compare_data <- movie_profit_enriched
```

------------------------------------------------------------------------

## Overview

This document covers **Part 3 — Modeling** of the Movie Data Analysis
project. Building on the cleaned dataset from Part 1 and the exploratory
findings from Part 2, we now attempt to answer a concrete analytical
question using statistical models.

The analysis follows a structured workflow:

1.  **Research question** — formulate what we want to explain or
    predict\
2.  **Variable selection** — choose the response and predictor
    variables, with justification\
3.  **Model 1** — a main-effects multiple linear regression\
4.  **Model 2** — an interaction model that extends Model 1\
5.  **Model comparison** — goodness of fit, interpretability, and
    practical conclusions

------------------------------------------------------------------------

## 1 Research Question

> **Does a higher production budget lead to higher worldwide revenue —
> and does this relationship differ across genres?**

The EDA in Part 2 revealed two contrasting patterns:

- Horror films are frequently produced on very low budgets yet achieve
  exceptionally high return-on-investment (ROI).
- Action and Adventure films tend to have large budgets and earn large
  absolute revenues.

This raises a natural question: is the *strength* of the budget–revenue
relationship the same for all genres, or does genre moderate how much
additional revenue a studio can expect from investing more money?

**Why this question matters:** studios and distributors use production
budgets as a primary lever for expected returns. If the budget
elasticity of revenue is genre-dependent, genre-specific investment
strategies are warranted.

**Hypothesis:**

- **H0** — the slope of log(budget) on log(worldwide gross) is identical
  across all five genres.
- **H1** — at least one genre has a significantly different budget
  elasticity from the others.

------------------------------------------------------------------------

## 2 Variable Selection

**Response variable:** `log(worldwide_gross)`

We model the natural logarithm of worldwide box-office revenue rather
than the raw dollar amount for two reasons:

1.  Revenue is right-skewed (a few blockbusters earn orders of magnitude
    more than the median film); the log transformation brings the
    distribution closer to normality and stabilises variance.
2.  On a log–log scale the regression coefficient on log(budget) is
    directly interpretable as an *elasticity* — a 1 % increase in budget
    is associated with a β % change in revenue.

**Predictors:**

| Predictor | Type | Rationale |
|------------------------|------------------------|------------------------|
| `log(production_budget)` | Continuous | Primary financial input; strong prior-evidence relationship with revenue |
| `genre` | Categorical (5 levels) | Genres differ systematically in audience size, marketing spend, and seasonality |
| `log(budget) × genre` | Interaction | Tests whether budget efficiency varies by genre (H₁ above) |

**Excluded variables:**

- `mpaa_rating` — substantial missing data and weak theoretical link to
  worldwide revenue compared with budget.
- `distributor` — too many levels (hundreds of studios) for a
  parsimonious model; would require a mixed-effects framework.
- `release_date` — time trends are real but outside the scope of this
  cross-sectional analysis.

------------------------------------------------------------------------

## 3 Data Preparation for Modeling

```{r model-data-prep, echo=TRUE}
# Keep only rows that have both key numeric variables and a genre label.
# Strictly positive values are required because we will log-transform them.
mlr_data <- compare_data[
	!is.na(compare_data$worldwide_gross)   &
	!is.na(compare_data$production_budget) &
	!is.na(compare_data$genre)             &
	compare_data$worldwide_gross   > 0     &
	compare_data$production_budget > 0,
	c("movie", "worldwide_gross", "production_budget", "genre")
]

# Log-transform both monetary variables (natural log).
mlr_data$log_worldwide <- log(mlr_data$worldwide_gross)
mlr_data$log_budget    <- log(mlr_data$production_budget)

# Set the reference level to the most frequent genre so that intercept
# and budget coefficient describe that baseline genre.
most_common_genre <- names(sort(table(mlr_data$genre), decreasing = TRUE))[1]
mlr_data$genre    <- relevel(factor(mlr_data$genre), ref = most_common_genre)

# Quick sanity check
data_summary <- data.frame(
	metric = c(
		"Rows used for modeling",
		"Reference genre (intercept baseline)",
		"Genre levels"
	),
	value = c(
		nrow(mlr_data),
		most_common_genre,
		paste(levels(mlr_data$genre), collapse = ", ")
	)
)
knitr::kable(data_summary, caption = "Modeling dataset summary")
```

The table above confirms the dataset size and the factor structure
before any model is fitted. Using the most frequent genre as the
reference level maximises the number of observations behind the baseline
coefficient, giving it the smallest standard error.

------------------------------------------------------------------------

## 4 Model 1 — Main-Effects Multiple Linear Regression

```{r model1, echo=TRUE}
# Model 1: additive effects of budget and genre (no interaction).
# Each genre coefficient shifts the intercept (i.e. the baseline revenue
# for a given budget level), but the budget slope is constrained to be
# identical for all genres.
model_main <- lm(log_worldwide ~ log_budget + genre, data = mlr_data)

main_summary <- summary(model_main)
```

**Model specification:**

$$\log(\text{worldwide\_gross}_i) = \beta_0 + \beta_1 \cdot \log(\text{budget}_i) + \sum_{k} \gamma_k \cdot \mathbb{1}[\text{genre}_i = k] + \varepsilon_i$$

where $\varepsilon_i \sim \mathcal{N}(0, \sigma^2)$ is assumed.

### 4.1 Model fit

```{r model1-fit}
fit_table <- data.frame(
	metric = c("R-squared", "Adjusted R-squared", "Residual Std. Error", "F-statistic p-value"),
	value  = c(
		round(main_summary$r.squared,     3),
		round(main_summary$adj.r.squared, 3),
		round(main_summary$sigma,         3),
		format.pval(pf(
			main_summary$fstatistic[1],
			main_summary$fstatistic[2],
			main_summary$fstatistic[3],
			lower.tail = FALSE
		), digits = 3)
	)
)
knitr::kable(fit_table, caption = "Model 1 — Goodness of fit")
```

**Interpretation:** An adjusted R² around 0.43 means that budget and
genre together explain roughly 43 % of the variance in log-revenue. This
is a reasonable result for box-office data, where unobserved factors
(marketing spend, cast, critic reception, release timing) account for
the remaining variation.

### 4.2 Coefficient table

```{r model1-coefs}
coef_m1 <- as.data.frame(main_summary$coefficients)
coef_m1 <- round(coef_m1, 4)
knitr::kable(coef_m1, caption = "Model 1 — Coefficients")
```

**Key takeaways:**

- `log_budget` — a 1 % increase in production budget is associated with
  a \~β % increase in worldwide gross (read the Estimate). The
  relationship is highly significant.
- Genre intercept shifts — each genre dummy shows how much higher or
  lower the average log-revenue is relative to the reference genre,
  holding budget constant.

------------------------------------------------------------------------

## 5 Model 2 — Interaction Model (Budget × Genre)

```{r model2, echo=TRUE}
# Model 2: budget slope is allowed to vary by genre.
# This tests whether the "return per dollar invested" differs across genres.
# The * operator expands to main effects + all pairwise interactions.
model_interaction <- lm(log_worldwide ~ log_budget * genre, data = mlr_data)

int_summary <- summary(model_interaction)
```

**Model specification:**

$$\log(\text{worldwide\_gross}_i) = \beta_0 + \beta_1 \cdot \log(\text{budget}_i) + \sum_k \gamma_k \cdot \mathbb{1}[\text{genre}_i = k] + \sum_k \delta_k \cdot \log(\text{budget}_i) \cdot \mathbb{1}[\text{genre}_i = k] + \varepsilon_i$$

The $\delta_k$ terms are the interaction coefficients. A significant
$\delta_k$ means that genre $k$ has a meaningfully different budget
elasticity from the reference genre.

### 5.1 Model fit

```{r model2-fit}
fit_table2 <- data.frame(
	metric = c("R-squared", "Adjusted R-squared", "Residual Std. Error", "F-statistic p-value"),
	value  = c(
		round(int_summary$r.squared,     3),
		round(int_summary$adj.r.squared, 3),
		round(int_summary$sigma,         3),
		format.pval(pf(
			int_summary$fstatistic[1],
			int_summary$fstatistic[2],
			int_summary$fstatistic[3],
			lower.tail = FALSE
		), digits = 3)
	)
)
knitr::kable(fit_table2, caption = "Model 2 — Goodness of fit")
```

### 5.2 Does the interaction term improve fit?

```{r model-comparison-anova}
# Likelihood-ratio / F-test comparing the nested models.
# A significant p-value indicates that at least one interaction term
# carries real explanatory power beyond the main effects.
anova_result <- anova(model_main, model_interaction)
knitr::kable(
	as.data.frame(anova_result),
	digits  = 4,
	caption = "ANOVA: Model 1 vs Model 2 (does adding budget × genre improve fit?)"
)
```

**Interpretation:** If the F-test p-value is small (e.g. \< 0.05), the
interaction model fits significantly better than the additive model,
supporting H₁ — budget efficiency is genre-dependent.

### 5.3 Term-level F-tests (interaction model)

```{r model2-drop1}
# drop1() removes each term in turn and reports the F-test,
# giving a cleaner picture of each predictor's marginal contribution.
drop1_result <- drop1(model_interaction, test = "F")
knitr::kable(
	as.data.frame(drop1_result),
	digits  = 4,
	caption = "Model 2 — Type II F-tests (drop1)"
)
```

### 5.4 Full coefficient table

```{r model2-coefs}
coef_m2 <- as.data.frame(int_summary$coefficients)
coef_m2 <- round(coef_m2, 4)
knitr::kable(coef_m2, caption = "Model 2 — Full coefficient table")
```

### 5.5 Budget elasticity by genre

```{r elasticity-table}
# The effective budget slope for each genre is:
#   β_budget (reference genre) + δ_k (genre-specific interaction term)
# A value of 1.0 means a 1 % rise in budget yields a 1 % rise in revenue
# (proportional); values above 1.0 indicate increasing returns.

base_slope  <- coef(model_interaction)["log_budget"]
int_coefs   <- coef(model_interaction)[grep("log_budget:genre", names(coef(model_interaction)))]
genre_names <- sub("log_budget:genre", "", names(int_coefs))

elasticity_table <- data.frame(
	genre       = c(most_common_genre, genre_names),
	elasticity  = round(c(base_slope, base_slope + int_coefs), 3),
	row.names   = NULL
)

# Sort descending so the highest-return genre is at the top
elasticity_table <- elasticity_table[order(-elasticity_table$elasticity), ]

knitr::kable(
	elasticity_table,
	caption = "Effective budget elasticity by genre (Model 2)"
)
```

**How to read this table:**

- An elasticity \> 1 means increasing returns — spending 1 % more on a
  film of that genre is associated with more than 1 % additional
  revenue.
- An elasticity close to 0 means budget has little predictive power for
  that genre's revenue.
- Horror's elasticity is notably lower than the other genres, consistent
  with the EDA finding that many low-budget horror films achieve very
  high ROI.

------------------------------------------------------------------------

## 6 Visualisation — Budget vs Revenue by Genre (log–log scale)

```{r interaction-plot, fig.height=6}
# Each genre gets its own scatter and OLS regression line on the log-log scale.
# The diverging slopes make the interaction effect visible:
#   parallel lines would indicate no interaction (Model 1 assumption);
#   non-parallel lines support Model 2.

genres  <- levels(mlr_data$genre)
colours <- palette.colors(length(genres), palette = "Okabe-Ito")  # colourblind-safe

plot(
	mlr_data$log_budget,
	mlr_data$log_worldwide,
	type = "n",
	xlab = "log(Production Budget)",
	ylab = "log(Worldwide Gross)",
	main = "Budget vs Revenue by Genre (log-log scale)\nNon-parallel lines indicate a budget × genre interaction"
)

for (i in seq_along(genres)) {
	g   <- genres[i]
	sub <- mlr_data[mlr_data$genre == g, ]
	fit <- lm(log_worldwide ~ log_budget, data = sub)

	points(
		sub$log_budget, sub$log_worldwide,
		col = adjustcolor(colours[i], alpha.f = 0.25),
		pch = 16, cex = 0.7
	)
	abline(fit, col = colours[i], lwd = 2)
}

legend("topleft", legend = genres, col = colours, lwd = 2, bty = "n")
```

**What this chart shows:** Each line represents the genre-specific OLS
fit on the log–log scale. The slope of each line is the estimated budget
elasticity for that genre. Lines that are clearly less steep than others
(notably Horror) indicate that budget buys less incremental revenue in
those genres — consistent with the elasticity table above.

------------------------------------------------------------------------

## 7 Model Comparison

```{r model-comparison-table}
# Side-by-side summary of the two models across key criteria.
comparison <- data.frame(
	criterion        = c(
		"R-squared",
		"Adjusted R-squared",
		"Residual Std. Error",
		"Number of parameters",
		"Budget × genre interaction",
		"Interpretability"
	),
	`Model 1 (main effects)` = c(
		round(main_summary$r.squared,     3),
		round(main_summary$adj.r.squared, 3),
		round(main_summary$sigma,         3),
		length(coef(model_main)),
		"No",
		"High — one slope for all genres"
	),
	`Model 2 (interaction)` = c(
		round(int_summary$r.squared,     3),
		round(int_summary$adj.r.squared, 3),
		round(int_summary$sigma,         3),
		length(coef(model_interaction)),
		"Yes",
		"Moderate — separate slope per genre"
	),
	check.names = FALSE
)
knitr::kable(comparison, caption = "Model 1 vs Model 2 — comparison")
```

**Discussion:**

| Dimension | Model 1 | Model 2 |
|------------------------|------------------------|------------------------|
| **Interpretability** | Simple: one budget slope, genre shifts the intercept | Richer: each genre has its own budget elasticity |
| **Goodness of fit** | Slightly lower R² | Higher R² (interaction captures genre heterogeneity) |
| **Parsimony** | Fewer parameters — lower risk of overfitting | More parameters (4 extra interaction terms) |
| **Statistical need** | Baseline | Justified if ANOVA F-test is significant |
| **Practical use** | Suitable for a quick first look | Preferred when genre-specific strategy is the goal |

**Recommendation:** If the ANOVA p-value is significant (see Section
5.2), Model 2 is preferred because the genre-specific slopes provide
actionable insight for production decisions. If the improvement is
marginal and the adjusted R² barely changes, the simpler Model 1 is
sufficient.

------------------------------------------------------------------------

## 8 Key Findings

```{r budget-elasticity-horror}
# Extract Horror's elasticity for inline use in the narrative below.
horror_elasticity <- elasticity_table$elasticity[elasticity_table$genre == "Horror"]
if (length(horror_elasticity) == 0) horror_elasticity <- NA
```

1.  **Budget is a strong, significant predictor of worldwide revenue**
    across all genres. The log–log specification reveals a near-
    proportional relationship for most genres (elasticity ≈ 1).

2.  **Horror is the clear outlier.** With an estimated budget elasticity
    of approximately `r round(horror_elasticity, 2)`, Horror films
    return significantly less additional revenue per additional dollar
    of budget than any other genre in this dataset. This aligns with the
    EDA observation that many of the highest-ROI films are low-budget
    horror productions — they succeed *despite* small budgets, not
    *because of* large ones.

3.  **Action and Adventure show elasticities above 1**, suggesting that
    higher spending (e.g. on visual effects, marketing tie-ins, and
    international distribution) does translate into more than
    proportional revenue gains for those genres.

4.  **Budget and genre together explain roughly 43–44 % of log-revenue
    variance.** The remainder is driven by unobserved factors such as
    marketing expenditure, star power, critical reception, and release
    timing — none of which are available in this dataset.

5.  **The interaction model (Model 2) fits better than the additive
    model (Model 1)**, confirming that genre moderates the
    budget–revenue relationship and that a single universal slope would
    be an oversimplification.

------------------------------------------------------------------------

## 9 Limitations

- **No marketing data.** Marketing budgets can equal or exceed
  production budgets for major releases. Omitting them likely inflates
  the coefficient on production budget.
- **Cross-sectional design.** The models treat all years as equivalent.
  Ticket prices, streaming competition, and global distribution have
  changed substantially over the dataset's time span, introducing
  unmodelled time trends.
- **Genre as a single label.** Many films span multiple genres; forcing
  a single label per film introduces measurement error in the genre
  variable.
- **Homoscedasticity assumption.** Even with log-transformation,
  residual variance may not be constant across the budget range. Robust
  standard errors (e.g. HC3) would provide more reliable inference, but
  are beyond the scope of this report.
- **External validity.** The dataset covers a curated selection of
  films. Straight-to-streaming and international productions are largely
  absent, limiting generalisability.
- **Outliers.** Log-transformation reduces the right-skew of the revenue
  distribution and stabilizes variance, but does not eliminate extreme
  observations. Action and Adventure films appear to dominate the upper
  tail of the revenue distribution; if confirmed, these high-leverage
  points could inflate the estimated budget elasticity for those genres,
  making their interaction terms appear more significant

```{r summary-metadata, include=FALSE}
n_raw       <- nrow(movie_profit)
n_clean     <- nrow(movie_profit_clean)
n_removed   <- n_raw - n_clean
pct_removed <- round(100 * n_removed / n_raw, 1)

year_min <- format(min(movie_profit_clean$release_date, na.rm = TRUE), "%Y")
year_max <- format(max(movie_profit_clean$release_date, na.rm = TRUE), "%Y")
```

# Part 4: Summary and Reflection

## Overview

This document covers **Part 4 — Summary and Reflection**, the closing
part of the Movie Data Analysis project. Parts 1–3 did the technical
work: cleaning the raw data, exploring it in depth, and building two
statistical models to investigate how production budget and genre relate
to worldwide box-office revenue, with a particular focus on the Horror
vs. non-Horror comparison.

This final part steps back from code and numbers and ties everything
together. It is organised into five sections:

1.  **Project summary** — a concise recap of what was done in each part
2.  **Most important findings** — the key results from the EDA and the
    models, brought together in one place
3.  **Main conclusions** — what these findings mean in practical terms
4.  **Limitations and difficulties** — what the data and models could
    *not* tell us, and why
5.  **Reflection on the process** — what was hardest, what was most
    interesting, and what we would do differently with more time,
    knowledge, or resources

------------------------------------------------------------------------

## 1 Project Summary

The project worked with a dataset of `r n_raw` films spanning the five
genres **Action, Adventure, Comedy, Drama, and Horror**, released
between `r year_min` and `r year_max`. The work was carried out in four
stages, each documented in its own part of the report.

**Part 1 — Data cleaning and preparation.** The raw CSV was loaded and
inspected for variable types, missing values, duplicates, and logical
inconsistencies (e.g. `worldwide_gross` smaller than `domestic_gross`).
`release_date` was converted from a text column to a proper `Date`. Rows
with missing values in any column other than `mpaa_rating` (which was
deliberately *kept*, since the rating is not needed for the core
analysis and dropping it would have removed many otherwise-usable rows),
or with a `domestic_gross`/`worldwide_gross` of exactly zero, were
removed. This step removed `r n_removed` rows (`r pct_removed`% of the
original data), leaving `r n_clean` films for analysis — a small enough
fraction that the dataset's overall structure was not materially
affected.

**Part 2 — Exploratory data analysis (EDA).** Every original column was
documented, and **12 new columns** were derived (above the required
minimum of 10), including `international_gross`, two ROI factors
(`roi_domestic_factor`, `roi_worldwide_factor`), two "top 100"
blockbuster flags, the `is_horror_movie` / `movie_type` grouping
variables, a `budget_band` categorical variable, two financial columns
rescaled to thousands of dollars for the interactive table, and a
formatted release date plus a `release_year` integer for time-series
charts. **13 charts** were produced, each accompanied by a short
explanation of why it was created and what it shows — covering genre and
rating distributions, releases over time, distributor market share, and
— most importantly — the Horror vs. non-Horror comparison across counts,
ROI, and budget bands.

**Part 3 — Modeling.** Building on the EDA, the research question *"Does
a higher production budget lead to higher worldwide revenue, and does
this relationship differ across genres?"* was formalised and tested with
two nested linear regression models on log-transformed budget and
revenue: **Model 1**, a main-effects model with a single budget slope
for all genres, and **Model 2**, an interaction model allowing the
budget slope (elasticity) to vary by genre. The two models were compared
on goodness of fit, interpretability, and practical usefulness.

**Part 4 — Summary and reflection.** Brings the threads from Parts 1–3
together, draws overall conclusions, and reflects honestly on the
process — including what was difficult, what was interesting, and what
we would change with more time or resources.

------------------------------------------------------------------------

## 2 Most Important Findings

### 2.1 What the exploratory analysis showed

- **Low-budget Horror films achieve unusually high ROI.** Among the
  cheapest Horror productions (budgets ≤ \$1M), several returned more
  than 100× their production cost, and the large majority returned at
  least 8× — a pattern not seen at the same scale in any other genre.
- **This advantage is structural, not a handful of lucky outliers.**
  Both the median ROI comparison and the budget-band distribution showed
  that Horror consistently occupies a different financial "regime" from
  Action, Adventure, Comedy, and Drama, even after accounting for
  outliers.
- **Bigger budgets do not help Horror the way they help other genres.**
  The scatter of budget vs. revenue for low-budget films suggested that,
  for Horror, spending more does not translate into proportionally more
  revenue — a pattern that motivated the modelling question in Part 3.

### 2.2 What the models showed

- **Budget is a strong, statistically significant predictor of worldwide
  revenue** for every genre (p \< 2×10⁻¹⁶ in both models). On the
  log–log scale, a 1% increase in budget is associated with roughly a 1%
  increase in revenue for most genres — close to a proportional
  relationship.
- **Horror is the clear outlier.** Its estimated budget elasticity (≈
  0.68) is well below that of the reference genre Drama (≈ 0.97) and
  below Comedy, Adventure, and Action, all of which showed elasticities
  at or above 1. In plain terms: for Horror, an extra dollar of budget
  buys noticeably less extra revenue than it does for other genres.
- **The interaction model (Model 2) fits the data better than the
  simpler main-effects model (Model 1)**, confirming that genre
  *moderates* the budget–revenue relationship rather than just shifting
  it up or down.
- **Budget and genre together explain roughly 43–44% of the variance in
  log-revenue** (adjusted R² ≈ 0.436 for the interaction model). This is
  a meaningful share, but it also means more than half of the variation
  in box-office revenue is driven by factors not captured in this
  dataset.

------------------------------------------------------------------------

## 3 Main Conclusions

Putting the EDA and modelling results together, three practical
conclusions stand out:

1.  **Low-budget Horror is a genuinely different segment of the film
    industry, not just a curiosity.** Both the descriptive statistics
    and the regression elasticities point the same direction: Horror
    films can be extremely profitable *relative to their cost*, even
    though they rarely achieve the largest absolute revenues. This
    explains why there are so many cheap Horror productions — money is
    simply not the main lever for this genre's success.

2.  **For most genres, "spend more to earn more" roughly holds — but not
    universally.** Action, Adventure, and Comedy show budget
    elasticities at or above 1, so increased investment is, on average,
    rewarded with at least proportional revenue gains. Horror is the
    exception, which has direct implications for how studios might think
    about genre-specific investment strategy.

3.  **A single "one model fits all genres" view of the budget–revenue
    relationship would be an oversimplification.** The interaction model
    consistently outperformed the additive model, supporting the idea
    that genre-aware models are more useful for anyone trying to use
    budget as a lever for predicting or planning revenue.

At the same time, the moderate R² values are a useful reminder that
production budget and genre — while informative — are far from the whole
story behind a film's box-office performance.

------------------------------------------------------------------------

## 4 Limitations and Difficulties

Several limitations affect how far these conclusions can be generalised:

- **No marketing data.** Marketing spend can rival or exceed production
  budgets for major releases. Its absence likely inflates the apparent
  importance of the production budget coefficient, since some of
  marketing's effect is probably being absorbed into the budget
  variable.
- **Cross-sectional treatment of a 1936–2018 time span.** The models
  treat all years as equivalent, but ticket prices, inflation, streaming
  competition, and global distribution patterns have changed enormously
  over this period. A film's production budget and gross revenue are not
  directly comparable across decades without some form of inflation
  adjustment.
- **Genre as a single label.** Many films could reasonably be classified
  under more than one genre (e.g. an action-horror crossover), but the
  dataset assigns exactly one genre per film. This introduces some
  measurement error into the genre variable and, by extension, into the
  genre-specific elasticities.
- **Homoscedasticity assumption.** Even after the log transformation,
  residual variance may not be perfectly constant across the budget
  range. Robust standard errors (e.g. HC3) would give more reliable
  inference but were outside the scope of this project.
- **Missing `mpaa_rating` values.** `mpaa_rating` had a non-trivial
  number of missing entries and was therefore excluded from the
  row-completeness check in Part 1 (otherwise too many otherwise-usable
  rows would have been dropped). As a result, it was used descriptively
  in the EDA but not as a model predictor.
- **External validity.** The dataset is a curated selection of
  theatrically-released films. Straight-to-streaming releases,
  international productions outside major US distribution, and very
  recent films are largely absent, which limits how far these findings
  can be extended to "the movie industry" as a whole today.

------------------------------------------------------------------------

## 5 Reflection on the Process

### 5.1 What was the most difficult part of the project?

The most challenging stage was the transition between **EDA and
modelling** — specifically, deciding *how* to formalise the pattern
spotted during EDA (cheap Horror films with huge ROI) into a model that
was both statistically sound and easy to interpret. Working on the
log–log scale made the elasticity interpretation cleaner, but it also
required being careful about strictly-positive values, choosing a
sensible reference genre, and explaining what an "elasticity" actually
means in terms a non-statistician would understand. Cleaning decisions
in Part 1 — particularly how to handle the missing `mpaa_rating` values
without discarding a large share of the dataset — also took more
iteration than expected, since every cleaning choice had knock-on
effects for the row counts reported throughout the rest of the report.

### 5.2 What was the most interesting part?

The most interesting moment was seeing the **Horror anomaly confirmed
statistically** after first noticing it informally during EDA. It is one
thing to see a handful of cheap Horror films with very high ROI in a
table; it is more convincing to see the same pattern show up as a
significantly lower budget elasticity in a regression model, controlling
for budget itself. Connecting a simple, intuitive observation ("cheap
Horror movies seem to do really well relative to their cost") to a
formal statistical result was the most rewarding part of the project.

### 5.3 What would we do differently if we started again?

If starting again, it would help to **plan the derived columns and the
research question earlier and together**, rather than designing the EDA
columns first and only formulating the modelling question afterwards. A
few of the columns created in Part 2 (e.g. the "top 100" blockbuster
flags) ended up being more useful for exploration than for the final
models. Sketching the modelling question at the start — even in a rough
form — would have made it easier to design derived columns that serve
both the EDA *and* the modelling stage, and would have reduced some
duplicated logic between Parts 2 and 3.

### 5.4 What would we improve with more time, knowledge, or resources?

With more time, knowledge, or data, the analysis could be extended in
several useful directions:

- **Add marketing budget data** (even an approximate figure) to separate
  the effect of production spending from promotional spending.
- **Adjust monetary values for inflation**, so that budgets and revenues
  from 1936 and 2018 are genuinely comparable, and consider adding
  release year as a covariate or fitting separate models by decade.
- **Allow multiple genre tags per film** instead of a single label, to
  reduce measurement error in the genre variable.
- **Use robust standard errors** (e.g. HC3) to make the significance
  tests more reliable in the presence of heteroscedasticity.
- **Try additional modelling approaches**, such as quantile regression
  to model the *spread* of ROI outcomes (not just the average), or a
  tree-based model (e.g. random forest) to explore non-linear
  interactions and rank predictor importance without assuming a log-log
  functional form in advance.

------------------------------------------------------------------------

## 6 Project at a Glance

The table below summarises the project's headline numbers in one place.
The dataset-derived figures (rows, genres, years) are computed live from
the cleaned data so that they always stay in sync with Part 1; the
remaining rows describe the scope of the analytical work carried out in
Parts 2 and 3.

```{r project-at-a-glance}
n_raw       <- nrow(movie_profit)
n_clean     <- nrow(movie_profit_clean)
n_removed   <- n_raw - n_clean
pct_removed <- round(100 * n_removed / n_raw, 1)

year_min <- format(min(movie_profit_clean$release_date, na.rm = TRUE), "%Y")
year_max <- format(max(movie_profit_clean$release_date, na.rm = TRUE), "%Y")

n_genres <- length(unique(movie_profit_clean$genre))

glance_table <- data.frame(
	metric = c(
		"Films in raw dataset",
		"Films after cleaning (Part 1)",
		"Rows removed during cleaning",
		"Percent of rows removed",
		"Genres analysed",
		"Years covered (cleaned data)",
		"New derived columns created (Part 2)",
		"Charts created (Part 2)",
		"Statistical models built (Part 3)"
	),
	value = c(
		n_raw,
		n_clean,
		n_removed,
		paste0(pct_removed, "%"),
		paste(sort(unique(movie_profit_clean$genre)), collapse = ", "),
		paste0(year_min, "\u2013", year_max),
		"12 (minimum required: 10)",
		"13 (minimum required: 10)",
		"2 (main-effects and interaction linear regression)"
	)
)

knitr::kable(glance_table, caption = "Project at a glance")
```
